---
title: "Module 3: Designing Around Interference"
subtitle: "Cluster Randomization, Switchbacks, and the Bias-Variance Tradeoff"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{css, echo=FALSE}
.remark-code, .remark-inline-code { font-size: 80%; }
.remark-slide-content { padding: 1em 2em; }
.small { font-size: 80%; }
.tiny { font-size: 65%; }
.highlight-box {
  background: #fff3e0;
  border-left: 4px solid #e65100;
  padding: 0.5em 1em;
  margin: 0.5em 0;
}
.blue-box {
  background: #e3f2fd;
  border-left: 4px solid #1565c0;
  padding: 0.5em 1em;
  margin: 0.5em 0;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, fig.align = "center",
  fig.width = 10, fig.height = 4.2, dpi = 150,
  message = FALSE, warning = FALSE
)
library(tidyverse)
theme_set(theme_minimal(base_size = 14))
set.seed(42)
```

# Course Map

<table>
<tr><th>#</th><th>Module</th><th>Status</th></tr>
<tr><td>1</td><td>The Experimental Ideal</td><td>done</td></tr>
<tr><td>2</td><td>SUTVA and When It Breaks</td><td>done</td></tr>
<tr><td><b>3</b></td><td><b>Designing Around Interference</b> <i>(you are here)</i></td><td>current</td></tr>
<tr><td>4</td><td>Power and Sample Size</td><td>upcoming</td></tr>
<tr><td>5</td><td>Analyzing Experiments</td><td>upcoming</td></tr>
<tr><td>6</td><td>Multiple Testing & Subgroups</td><td>upcoming</td></tr>
<tr><td>7</td><td>External Validity</td><td>upcoming</td></tr>
<tr><td>8</td><td>Advanced Topics</td><td>upcoming</td></tr>
</table>

---

# Last Time: SUTVA Breaks Everywhere

Module 2 showed that interference biases the naive ATE:

- **Marketplace:** bonus drivers take rides from control drivers (overestimates effect)
- **Networks:** treated co-authors spill over to control studies (underestimates effect)
- **General equilibrium:** small experiment $\neq$ full rollout

--

Today: **how to design experiments that work despite interference.**

The core idea is simple: randomize at a level where interference is contained.

--

.highlight-box[
Every design in this module trades **bias** (from interference) for **variance** (from fewer independent units). The art is finding the sweet spot.
]

---

# The Setup: 40 Cities, 50 Drivers Each

```{r setup-market, echo=FALSE}
n_cities <- 40
drivers_per_city <- 50
n_total <- n_cities * drivers_per_city

market <- tibble(
  driver_id = 1:n_total,
  city_id = rep(1:n_cities, each = drivers_per_city),
  skill = rnorm(n_total),
  city_effect = rep(rnorm(n_cities, 0, 0.3), each = drivers_per_city)
)

true_ate <- mean(
  plogis(-0.5 + 0.3 * market$skill + market$city_effect + 0.5) -
  plogis(-0.5 + 0.3 * market$skill + market$city_effect)
)
```

A ride-sharing platform operates in **40 cities** with **50 drivers each** (2,000 total). They want to test a driver bonus.

--

**The problem:** within each city, interference is strong. Bonus drivers take rides from non-bonus drivers.

--

**Design options:**

| Design | Unit of randomization | Independent units |
|--------|----------------------|-------------------|
| Individual | Driver | 2,000 (but biased) |
| City cluster | City | 40 (unbiased within city) |
| Region cluster | 5-city region | 8 (less bias, more variance) |

---

# Individual vs Cluster: Side by Side

```{r indiv-vs-cluster, echo=FALSE, fig.height=4}
interference <- 0.3

sim_individual <- function() {
  d <- market |> mutate(bonus = sample(rep(c(0, 1), each = n() / 2)))
  d <- d |>
    group_by(city_id) |>
    mutate(frac_t = mean(bonus),
           interf = -interference * frac_t * (1 - bonus)) |>
    ungroup() |>
    mutate(y = rbinom(n(), 1, prob = plogis(-0.5 + 0.3 * skill +
                                              city_effect + 0.5 * bonus + interf)))
  mean(d$y[d$bonus == 1]) - mean(d$y[d$bonus == 0])
}

sim_cluster <- function() {
  arms <- tibble(city_id = 1:n_cities,
                 bonus = sample(rep(c(0, 1), each = n_cities / 2)))
  d <- market |> left_join(arms, by = "city_id") |>
    mutate(y = rbinom(n(), 1, prob = plogis(-0.5 + 0.3 * skill +
                                              city_effect + 0.5 * bonus)))
  mean(d$y[d$bonus == 1]) - mean(d$y[d$bonus == 0])
}

indiv_ests <- replicate(1000, sim_individual())
cluster_ests <- replicate(1000, sim_cluster())

bind_rows(
  tibble(est = indiv_ests, design = "Individual (biased, low variance)"),
  tibble(est = cluster_ests, design = "City cluster (unbiased, high variance)")
) |>
  ggplot(aes(est, fill = design)) +
  geom_histogram(bins = 40, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = true_ate, color = "firebrick",
             linewidth = 1.2, linetype = "dashed") +
  annotate("text", x = true_ate - 0.005, y = 70,
           label = paste("True ATE =", round(true_ate, 3)),
           color = "firebrick", hjust = 1, size = 4) +
  scale_fill_manual(values = c("steelblue", "darkorange")) +
  labs(title = "Individual vs cluster randomization: bias-variance tradeoff",
       x = "Estimated ATE", y = "Count") +
  theme(legend.position = "top")
```

---

# Why Clustering Eliminates Bias

.pull-left[
**Individual randomization:**
- Within each city, ~50% treated, ~50% control
- Treated drivers take rides from control drivers
- Control retention **drops** due to interference
- Naive ATE = direct effect + interference bias

**City-level clustering:**
- Each city is 100% treated or 100% control
- No within-city interference (everyone has the same status)
- Between-city interference is small (drivers don't cross cities)
- ATE is unbiased
]

.pull-right[
```{r cluster-logic, echo=FALSE, fig.height=5}
tibble(
  Design = rep(c("Individual", "Cluster"), each = 2),
  Group = rep(c("Control", "Treated"), 2),
  `Within-city interference` = c("Yes (harmed)", "No effect",
                                  "None", "None"),
  Outcome = c("Biased down", "Correct", "Correct", "Correct")
) |>
  knitr::kable(align = "llll")
```

.highlight-box[
**Intuition:** clustering works because interference happens *within* clusters. If you randomize the whole cluster, there's no one to interfere with.
]
]

---

# But Clustering Costs You Power

```{r power-cost, echo=FALSE, fig.height=3.8}
tibble(
  Design = c("Individual\n(2,000 units)", "City cluster\n(40 units)",
             "Region cluster\n(8 units)"),
  `Effective N` = c(2000, 40, 8),
  order = c(1, 2, 3)
) |>
  ggplot(aes(reorder(Design, order), `Effective N`)) +
  geom_col(fill = "steelblue", width = 0.5) +
  geom_text(aes(label = `Effective N`), vjust = -0.5, size = 5) +
  labs(title = "Clustering reduces effective sample size dramatically",
       subtitle = "From 2,000 independent units to 40 (or 8) clusters",
       x = "", y = "Number of independent units") +
  ylim(0, 2200)
```

.highlight-box[
**The bias-variance tradeoff:** individual randomization is biased but precise. Cluster randomization is unbiased but noisy. You need to choose the right level.
]

---

# Design Effect and ICC

The **design effect** quantifies the power loss from clustering:

$$n_{eff} = \frac{n}{1 + (m - 1) \cdot \rho}$$

where $m$ = cluster size and $\rho$ = intra-cluster correlation (ICC).

--

```{r icc-table, echo=FALSE}
expand_grid(
  m = c(10, 50, 200),
  rho = c(0.01, 0.05, 0.10, 0.20)
) |>
  mutate(
    design_effect = 1 + (m - 1) * rho,
    n_eff_per_10000 = round(10000 / design_effect),
    pct_lost = paste0(round(100 * (1 - 1/design_effect)), "%")
  ) |>
  select(`Cluster size (m)` = m, `ICC (rho)` = rho,
         `Design effect` = design_effect,
         `Eff. N (of 10,000)` = n_eff_per_10000,
         `% power lost` = pct_lost) |>
  knitr::kable(digits = 1, align = "rrrrr")
```

--

.highlight-box[
Even moderate ICC (0.05) with medium clusters (50) loses ~70% of effective sample size. This is why cluster experiments need many more total observations.
]

<a href="#icc-derivation" style="position:absolute; bottom:12px; left:40px; font-size:11px; background:#e8eaf6; padding:2px 8px; border-radius:3px; z-index:100;">ICC derivation</a>

---

# The Bias-Variance Tradeoff: Simulated

```{r tradeoff-sim, echo=FALSE, fig.height=3.8}
sim_varying <- function(n_clusters) {
  cities_per <- n_cities / n_clusters
  arms <- tibble(cluster_id = 1:n_clusters,
                 bonus = sample(rep(c(0, 1), each = n_clusters / 2)))
  d <- market |>
    mutate(cluster_id = ceiling(city_id / cities_per)) |>
    left_join(arms, by = "cluster_id") |>
    group_by(cluster_id) |>
    mutate(interf = -interference * (1 - mean(bonus)) *
             (1 - bonus) * (n_clusters / n_cities)) |>
    ungroup() |>
    mutate(y = rbinom(n(), 1, prob = plogis(-0.5 + 0.3 * skill +
                                              city_effect + 0.5 * bonus + interf)))
  mean(d$y[d$bonus == 1]) - mean(d$y[d$bonus == 0])
}

cluster_sizes <- c(40, 20, 10, 8, 4, 2)
tradeoff <- map_dfr(cluster_sizes, function(nc) {
  ests <- replicate(500, sim_varying(nc))
  tibble(n_clusters = nc, cities_per = n_cities / nc,
         bias = abs(mean(ests) - true_ate),
         sd = sd(ests),
         rmse = sqrt(mean((ests - true_ate)^2)))
})

tradeoff |>
  pivot_longer(c(bias, sd, rmse), names_to = "metric", values_to = "value") |>
  mutate(metric = factor(metric, levels = c("bias", "sd", "rmse"),
                         labels = c("|Bias|", "Std Dev", "RMSE"))) |>
  ggplot(aes(cities_per, value, color = metric)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(values = c("firebrick", "steelblue", "grey40")) +
  labs(title = "Bias-variance tradeoff as cluster size varies",
       subtitle = "RMSE is minimized at an intermediate cluster size",
       x = "Cities per cluster", y = "Value", color = "") +
  theme(legend.position = "top")
```

.highlight-box[
**RMSE** = $\sqrt{\text{bias}^2 + \text{variance}}$. The optimal cluster size minimizes RMSE, balancing the two forces.
]

---

# Switchback Designs

An alternative to spatial clustering: **randomize over time**.

--

.pull-left[
**How it works:**
1. Divide time into periods (e.g., 2-hour blocks)
2. Each period, randomize each city to treatment or control
3. Compare outcomes in treated vs. control periods
4. Every city serves as its own control (removes city fixed effects)
]

--

.pull-right[
```{r switchback-diagram, echo=FALSE, fig.height=4}
set.seed(123)
sb <- expand_grid(city = paste("City", 1:6), period = 1:12) |>
  mutate(treated = sample(c(0, 1), n(), replace = TRUE))

ggplot(sb, aes(period, city, fill = factor(treated))) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_manual(values = c("grey80", "steelblue"),
                    labels = c("Control", "Treated")) +
  scale_x_continuous(breaks = 1:12) +
  labs(title = "Switchback design: 6 cities x 12 periods",
       x = "Time period", y = "", fill = "") +
  theme(legend.position = "top", panel.grid = element_blank())
```
]

---

# Switchback: Strengths and Risks

**Strengths:**
- Every unit (city) appears in both arms over time -- removes unit-level confounders
- More "independent observations" than pure cluster randomization
- Natural for marketplace experiments where interference is spatial, not temporal

--

**Risks:**

```{r carryover, echo=FALSE, fig.height=2.8}
tibble(
  period = 1:10,
  treatment = c(1, 1, 0, 0, 1, 0, 1, 1, 0, 0),
  true_outcome = c(0.6, 0.65, 0.4, 0.42, 0.62, 0.41, 0.61, 0.64, 0.40, 0.41),
  with_carryover = c(0.6, 0.65, 0.48, 0.43, 0.62, 0.47, 0.61, 0.64, 0.48, 0.42)
) |>
  pivot_longer(c(true_outcome, with_carryover), names_to = "type", values_to = "y") |>
  mutate(type = if_else(type == "true_outcome", "No carryover", "With carryover")) |>
  ggplot(aes(period, y, color = type)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  geom_tile(data = tibble(period = 1:10,
                          treatment = c(1,1,0,0,1,0,1,1,0,0)),
            aes(period, y = 0.35, fill = factor(treatment)),
            height = 0.03, inherit.aes = FALSE, alpha = 0.5) +
  scale_color_manual(values = c("steelblue", "darkorange")) +
  scale_fill_manual(values = c("grey80", "steelblue"), guide = "none") +
  labs(title = "Carryover effect: treatment bleeds into control periods",
       x = "Period", y = "Outcome", color = "") +
  theme(legend.position = "top")
```

.highlight-box[
**Carryover** = treatment in period $t$ affects outcomes in period $t+1$, even if $t+1$ is assigned to control. Fix: add "washout" or "burn-in" periods between switches. Cost: fewer usable periods.
]

---

# Switchback: Simulation

```{r switchback-sim, echo=FALSE, fig.height=3.8}
sim_switchback <- function(carryover = 0) {
  n_periods <- 20; n_sw_cities <- 10
  design <- expand_grid(city_id = 1:n_sw_cities, period = 1:n_periods) |>
    mutate(bonus = sample(c(0, 1), n(), replace = TRUE),
           city_effect = rep(rnorm(n_sw_cities, 0, 0.3), times = n_periods),
           period_effect = rep(rnorm(n_periods, 0, 0.1), each = n_sw_cities))
  design <- design |>
    group_by(city_id) |>
    mutate(prev = lag(bonus, default = 0),
           carry = carryover * prev) |>
    ungroup() |>
    mutate(y = if_else(bonus == 1,
                       plogis(-0.5 + city_effect + period_effect + 0.5),
                       plogis(-0.5 + city_effect + period_effect)) +
             carry + rnorm(n(), 0, 0.03))
  mean(design$y[design$bonus == 1]) - mean(design$y[design$bonus == 0])
}

sb0 <- replicate(500, sim_switchback(0))
sb1 <- replicate(500, sim_switchback(0.05))

bind_rows(
  tibble(est = sb0, scenario = "No carryover"),
  tibble(est = sb1, scenario = "Carryover = 0.05")
) |>
  ggplot(aes(est, fill = scenario)) +
  geom_histogram(bins = 35, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = mean(sb0), color = "firebrick",
             linewidth = 1, linetype = "dashed") +
  scale_fill_manual(values = c("steelblue", "darkorange")) +
  labs(title = "Switchback estimates: carryover creates upward bias",
       subtitle = "Control periods contaminated by prior treatment -> overestimates effect",
       x = "Estimated ATE", y = "Count") +
  theme(legend.position = "top")
```

---

# Two-Sided Marketplace Designs

In a ride-sharing experiment, you can randomize **riders** instead of **drivers**:

--

.pull-left[
**Randomize riders:**
- 50% of riders see new pricing
- 50% see old pricing
- Riders don't interact with each other (mostly)
- No interference on the randomization side

**Measure driver outcomes:**
- Driver earnings, wait times, retention
- These reflect the equilibrium effect
]

--

.pull-right[
**Why it works:**
- Each rider's experience is independent (they don't share rides)
- But their collective demand changes driver supply allocation
- You get the direct effect on riders AND the indirect effect on drivers

**When it doesn't work:**
- If treated riders change market-wide conditions (e.g., surge pricing triggers)
- If the treatment fraction is large enough to move equilibrium prices
]

---

# Geo Experiments

For online advertising and pricing experiments, **geographic units** are natural clusters:

--

```{r geo-diagram, echo=FALSE, fig.height=3.5}
set.seed(42)
geos <- tibble(
  dma = paste("DMA", 1:20),
  treated = sample(rep(c(0, 1), each = 10)),
  baseline = runif(20, 1000, 5000),
  lift = if_else(treated == 1, 0.05, 0),
  revenue = baseline * (1 + lift) + rnorm(20, 0, 100)
)

ggplot(geos, aes(reorder(dma, revenue), revenue,
                 fill = factor(treated))) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = c("grey70", "steelblue"),
                    labels = c("Control", "Treated")) +
  coord_flip() +
  labs(title = "Geo experiment: 20 DMAs, 10 treated, 10 control",
       subtitle = "Each DMA is an independent unit. High variance between DMAs.",
       x = "", y = "Revenue ($)", fill = "") +
  theme(legend.position = "top")
```

--

.blue-box[
**Google/Meta approach:** "GeoX" or "geo-based incrementality." Randomize ad spend across 50-200 DMAs. Estimate lift at the DMA level. Power depends on between-DMA variance and number of DMAs, not number of users.
]

---

# Ego-Cluster Randomization for Networks

For **network interference** (social media, academic co-authorship), standard clustering is hard because network boundaries are unclear.

--

**Ego-cluster approach:**
1. Pick a focal unit (ego)
2. Define the cluster as ego + all immediate neighbors (alter set)
3. Randomize the **entire cluster** to the same arm
4. Interference within the cluster is absorbed

--

```{r ego-cluster, echo=FALSE, fig.height=3}
# Simple network visualization
set.seed(42)
nodes <- tibble(
  id = 1:12,
  x = c(0, 1, 2, 0.5, 1.5, 4, 5, 6, 4.5, 5.5, 3, 3.5),
  y = c(1, 2, 1, 0, 0, 1, 2, 1, 0, 0, 1.5, 0.5),
  cluster = c(rep("Cluster A (treated)", 5), rep("Cluster B (control)", 5),
              rep("Between", 2))
)
edges <- tibble(
  from = c(1,1,2,2,3,4,5, 6,6,7,7,8,9,10, 3,11,11,12),
  to =   c(2,4,3,5,5,5,3, 7,9,8,10,10,10,8, 11,12,6,9)
)

edge_df <- edges |>
  left_join(nodes |> select(id, x, y), by = c("from" = "id")) |>
  rename(x_from = x, y_from = y) |>
  left_join(nodes |> select(id, x, y), by = c("to" = "id")) |>
  rename(x_to = x, y_to = y)

ggplot() +
  geom_segment(data = edge_df, aes(x = x_from, y = y_from,
                                    xend = x_to, yend = y_to),
               color = "grey70", linewidth = 0.5) +
  geom_point(data = nodes, aes(x, y, color = cluster), size = 6) +
  geom_text(data = nodes, aes(x, y, label = id), size = 3, color = "white") +
  scale_color_manual(values = c("grey50", "steelblue", "darkorange")) +
  labs(title = "Ego-cluster randomization: entire cluster gets same arm",
       subtitle = "Nodes 11-12 bridge the clusters -- between-cluster spillover is the residual risk",
       color = "") +
  theme_void(base_size = 13) + theme(legend.position = "top")
```

---

# Application: City-Level Pricing Experiment

A ride-sharing company wants to test a new **surge pricing algorithm**.

--

.pull-left[
**Why cluster at the city level?**
- Surge pricing affects ALL riders and drivers in a city
- If 50% of riders see new surge and 50% see old, drivers face inconsistent signals
- City-level randomization: each city gets one algorithm
- No within-city interference
]

--

.pull-right[
**The power problem:**
- Only 40 cities available
- 20 treatment, 20 control
- ICC of driver outcomes within cities $\approx 0.08$
- Effective $N \approx 40$ (not 2,000)
- Need a **large** effect to detect anything
]

--

.highlight-box[
**The interview question:** "We have 40 cities. Can we run a cluster-randomized experiment?" Answer: probably, but only if the expected effect is large. Calculate the minimum detectable effect (MDE) and check if it's business-relevant.
]

---

# Application: Switchback for Surge Pricing

**Alternative:** use a switchback design within each city.

--

- Divide each day into 2-hour blocks
- Randomize which blocks use the new vs. old surge algorithm
- 12 blocks/day x 30 days x 40 cities = 14,400 city-period cells
- Effective N is much larger than 40 clusters

--

**But:**
- Carryover: if new surge drives drivers away, the next control period starts with fewer drivers
- Need washout periods (waste time and data)
- Confounding with time-of-day effects (surge matters more at peak times)

--

.blue-box[
**Practical design:** stratified switchback. Within each city, pair "similar" time blocks (e.g., Monday 5-7pm with Wednesday 5-7pm) and randomize within pairs. This controls for time-of-day and day-of-week effects.
]

---

# Application: Geo Incrementality for Online Ads

An e-commerce company wants to measure whether their TV ad campaign drives sales.

--

**Design:**
- 100 DMAs (Designated Market Areas) across the US
- Randomly assign 50 DMAs to see ads, 50 to no ads
- Measure sales lift at the DMA level
- Control for baseline sales (CUPED-style adjustment at the DMA level)

--

**Challenges:**
- DMAs vary enormously in size (NYC vs. rural Montana)
- Need to weight by population or stratify by DMA size
- Spillovers: people in control DMAs can see ads on streaming/social media (imperfect compliance)
- Attribution: sales in a DMA might come from visitors from neighboring DMAs

--

.highlight-box[
**Key insight:** geo experiments work best when the treatment is *geographically contained* (TV, billboards, local promotions). They work poorly for digital treatments that cross geo boundaries (social media ads, viral content).
]

---

# Choosing the Right Design

| Interference type | Recommended design | Key tradeoff |
|-------------------|-------------------|--------------|
| Within-market (supply/demand) | City-level cluster | Bias vs. power |
| Temporal (pricing, algorithms) | Switchback | Carryover vs. efficiency |
| Network (social, co-authorship) | Ego-cluster | Cluster size vs. coverage |
| Geographic (ads, promotions) | Geo experiment | Geo count vs. precision |
| Two-sided marketplace | Randomize one side | Indirect effects only |

--

**Decision framework:**

1. **Where does interference happen?** Within cities? Across time? Through networks?
2. **Can you cluster at that level?** How many independent clusters do you have?
3. **Is the effect large enough to detect with that many clusters?**
4. **Is there carryover/leakage between clusters?** If so, switchback may be risky.

---

# Summary of the Bias-Variance Tradeoff

```{r final-tradeoff, echo=FALSE, fig.height=4}
tradeoff_summary <- tibble(
  design = c("Individual", "Small cluster\n(5 cities)", "City\n(1 city)",
             "Region\n(5 cities)", "Full cluster\n(20 cities)"),
  bias = c(0.06, 0.03, 0.005, 0.002, 0.001),
  variance = c(0.008, 0.015, 0.025, 0.045, 0.09),
  order = 1:5
) |>
  mutate(rmse = sqrt(bias^2 + variance^2))

ggplot(tradeoff_summary, aes(variance, bias, label = design)) +
  geom_point(aes(size = rmse), color = "steelblue", alpha = 0.7) +
  geom_text(vjust = -1.2, size = 3.5) +
  scale_size_continuous(range = c(3, 10), guide = "none") +
  labs(title = "Every design sits on the bias-variance frontier",
       subtitle = "Bigger clusters reduce bias but increase variance. RMSE picks the winner.",
       x = "Variance (fewer independent units)", y = "Bias (residual interference)") +
  theme(panel.grid.minor = element_blank())
```

---

# Key Takeaways

1. **Cluster randomization** eliminates interference bias by assigning entire clusters to the same arm. Cost: fewer independent units = more variance.

2. **Switchback designs** randomize over time. Good when interference is spatial. Watch for carryover effects.

3. **Two-sided marketplace designs**: randomize one side, measure the other. Works when one side doesn't interact with itself.

4. **Geo experiments**: natural clusters for advertising. Need many geos (50+) and geographically contained treatments.

5. **Ego-cluster randomization**: for network interference. Cluster = focal unit + neighbors.

6. **The bias-variance tradeoff** is the central tension. Minimize **RMSE**, not just bias or variance alone.

7. The **design effect** formula $n_{eff} = n / [1 + (m-1)\rho]$ tells you how much power you lose from clustering.

---

# Exercise Preview

In the exercise you will:

1. Compare individual vs. city-level randomization in a marketplace simulation
2. Show that clustering eliminates bias but increases variance
3. Sweep over cluster sizes and find the RMSE-minimizing level
4. Simulate a switchback design with and without carryover
5. Compute design effects for different ICC values

See `exercise.R` for the starter code.

---
class: center, middle, inverse

# Backup Slides

---
name: icc-derivation

# Backup: ICC and Design Effect Derivation

The **intra-cluster correlation (ICC)** is:

$$\rho = \frac{\sigma^2_b}{\sigma^2_b + \sigma^2_w}$$

where $\sigma^2_b$ = between-cluster variance and $\sigma^2_w$ = within-cluster variance.

--

For a cluster-level mean with $m$ observations per cluster:

$$\text{Var}(\bar{Y}_c) = \frac{\sigma^2_w}{m} + \sigma^2_b = \frac{\sigma^2_w}{m} + \frac{\rho}{1-\rho} \sigma^2_w$$

--

The variance of the overall mean across $K$ clusters:

$$\text{Var}(\bar{Y}) = \frac{1}{K} \left[\frac{\sigma^2_w}{m}(1 + (m-1)\rho)\right]$$

--

Compare to simple random sampling (SRS) with $n = Km$ units:

$$\text{Var}_{SRS}(\bar{Y}) = \frac{\sigma^2}{n}$$

The ratio $\frac{\text{Var}(\bar{Y})}{\text{Var}_{SRS}(\bar{Y})} = 1 + (m-1)\rho$ is the **design effect (DEFF)**.

The effective sample size is $n_{eff} = n / \text{DEFF}$.

---
name: switchback-formal

# Backup: Switchback Estimator

For a switchback design with cities $c = 1, \ldots, C$ and periods $t = 1, \ldots, T$:

$$Y_{ct} = \mu_c + \gamma_t + \tau \cdot D_{ct} + \delta \cdot D_{c,t-1} + \varepsilon_{ct}$$

--

- $\mu_c$: city fixed effect
- $\gamma_t$: period fixed effect
- $D_{ct}$: treatment indicator for city $c$ at time $t$
- $\delta$: carryover parameter (ideally $\delta = 0$)

--

The standard switchback estimator ignores carryover:

$$\hat{\tau} = \frac{1}{CT}\sum_{c,t} (2D_{ct} - 1) Y_{ct}$$

--

If $\delta \neq 0$, the estimator is biased:

$$E[\hat{\tau}] = \tau + \delta \cdot \text{Corr}(D_{ct}, D_{c,t-1})$$

Under i.i.d. randomization, $\text{Corr}(D_{ct}, D_{c,t-1}) \approx 0$, so the bias is small but not zero in finite samples. Including $D_{c,t-1}$ as a covariate or using burn-in periods mitigates this.

---
name: ego-cluster-formal

# Backup: Ego-Cluster Randomization

Define the **ego network** of unit $i$ as $\mathcal{N}_i = \{i\} \cup \{j : j \text{ is connected to } i\}$.

--

**Ego-cluster randomization** assigns the same treatment to all units in $\mathcal{N}_i$:

$$D_j = D_i \quad \forall j \in \mathcal{N}_i$$

--

This ensures that, for unit $i$, all first-degree neighbors have the same treatment status. If interference is limited to first-degree connections:

$$Y_i(D_i, \mathbf{D}_{\mathcal{N}_i}) = Y_i(D_i, D_i, \ldots, D_i) = Y_i(D_i)$$

and SUTVA is restored within the ego cluster.

--

**Challenges:**
- Ego clusters overlap: if $i$ and $j$ are neighbors, $\mathcal{N}_i \cap \mathcal{N}_j \neq \emptyset$
- Must resolve overlaps (e.g., graph coloring, independent set sampling)
- Power depends on the number of *non-overlapping* ego clusters
- Second-degree interference is not addressed
