---
title: "Module 3: Designing Around Interference"
subtitle: "Cluster Randomization, Switchbacks, and the Bias-Variance Tradeoff"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{css, echo=FALSE}
.remark-code, .remark-inline-code { font-size: 80%; }
.remark-slide-content { padding: 1em 2em; }
.small { font-size: 80%; }
.tiny { font-size: 65%; }
.highlight-box {
  background: #fff3e0;
  border-left: 4px solid #e65100;
  padding: 0.5em 1em;
  margin: 0.5em 0;
}
.blue-box {
  background: #e3f2fd;
  border-left: 4px solid #1565c0;
  padding: 0.5em 1em;
  margin: 0.5em 0;
}
.nav-btn {
  position: absolute;
  bottom: 12px;
  left: 40px;
  font-size: 11px;
  background: #e8eaf6;
  padding: 2px 8px;
  border-radius: 3px;
  z-index: 100;
  text-decoration: none;
  color: #1a237e;
}
.nav-btn:hover { background: #c5cae9; }
.inline-btn {
  font-size: 11px;
  background: #e8eaf6;
  padding: 2px 8px;
  border-radius: 3px;
  text-decoration: none;
  color: #1a237e;
  margin-right: 6px;
  vertical-align: middle;
}
.inline-btn:hover { background: #c5cae9; }
.nav-btn-br {
  position: absolute;
  bottom: 12px;
  right: 70px;
  font-size: 11px;
  background: #e8eaf6;
  padding: 2px 8px;
  border-radius: 3px;
  z-index: 100;
  text-decoration: none;
  color: #1a237e;
}
.nav-btn-br:hover { background: #c5cae9; }
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, fig.align = "center",
  fig.width = 10, fig.height = 4.2, dpi = 150,
  message = FALSE, warning = FALSE
)
library(tidyverse)
theme_set(theme_minimal(base_size = 14))
set.seed(42)
```

# Course Map

<table>
<tr><th>#</th><th>Module</th><th>Status</th></tr>
<tr><td>1</td><td><a href="../module-01/slides.html">The Experimental Ideal</a></td><td>✓ done</td></tr>
<tr><td>2</td><td><a href="../module-02/slides.html">SUTVA and When It Breaks</a></td><td>✓ done</td></tr>
<tr><td><b>3</b></td><td><b>Designing Around Interference</b> <i>(you are here)</i></td><td>current</td></tr>
<tr><td>4</td><td>Power and Sample Size</td><td>upcoming</td></tr>
<tr><td>5</td><td><a href="../module-05/slides.html">Analyzing Experiments</a></td><td>✓ done</td></tr>
<tr><td>6</td><td>Multiple Testing & Subgroups</td><td>upcoming</td></tr>
<tr><td>7</td><td><a href="../module-07/slides.html">External Validity</a></td><td>✓ done</td></tr>
<tr><td>8</td><td><a href="../module-08/slides.html">Beyond the A/B Test</a></td><td>✓ done</td></tr>
</table>

---

# Last Time: SUTVA Breaks Everywhere

Module 2 showed that interference biases the naive ATE in two settings we'll keep using here:

- **Marketplace (zone-notification experiment):** notified drivers compete for rides in the same zone, contaminating the comparison
- **Networks (author-nudge experiment):** control researchers learn the new reporting practice from treated co-authors

--

Today: **how to design experiments that work despite interference.**

The core idea is simple: randomize at a level where interference is contained.

--

.highlight-box[
Every design in this module trades **bias** (from interference) for **variance** (from fewer independent units). The art is finding the sweet spot.
]

---

# The Setup: 40 Cities, 5,000 Drivers Each

```{r setup-market, echo=FALSE}
n_cities <- 40
drivers_per_city <- 5000
n_total <- n_cities * drivers_per_city

market <- tibble(
  driver_id = 1:n_total,
  city_id = rep(1:n_cities, each = drivers_per_city),
  experience = rnorm(n_total),
  # City heterogeneity: demand density, driver pool, market structure.
  # Centered at zero so it doesn't shift the average baseline.
  city_effect = rep(rnorm(n_cities, mean = 0, sd = 0.10), each = drivers_per_city)
)

# Marketplace parameters (all on the LPM/accept-rate scale)
direct_effect <- 0.05   # per-driver direct boost from being notified, holding others fixed
interference <- 0.07    # accept-rate penalty per share treated (hits BOTH arms — symmetric)
policy_effect <- direct_effect - interference  # full-rollout effect: -0.02

# We target the policy (full-rollout / equilibrium) effect as the true estimand
true_ate <- policy_effect
```

A ride-sharing platform operates in **40 cities** with **5,000 drivers each** (200,000 total). They want to test the **zone-notification** feature from Module 1.

--

**Within-city interference (symmetric):** as more drivers are notified, the high-demand zone gets crowded — *every* driver in the city (notified or not) faces lower accept rates. Direct boost $\tau = 0.05$, interference $\lambda = 0.07$ → policy effect = $-0.02$ at full rollout.

--

**Design options:**

| Design | Unit of randomization | Independent units | Estimand |
|--------|----------------------|-------------------|----------|
| Individual | Driver | 200,000 | Direct effect (+0.05) |
| City cluster | City | 40 | Policy effect (−0.02) |
| Region cluster | 5-city region | 8 | Policy effect (−0.02) |

---
name: indiv-cluster-main

# Individual vs Cluster: Side by Side

```{r indiv-vs-cluster, echo=FALSE, fig.height=4}
sim_individual <- function() {
  d <- market |> mutate(notification = sample(rep(c(0, 1), each = n() / 2)))
  d <- d |>
    group_by(city_id) |>
    mutate(frac_t = mean(notification),
           # Symmetric interference: every driver in the city competes against frac_t treated neighbors
           y0 = pmin(1, pmax(0, 0.4 + 0.2 * experience + city_effect
                                  - interference * frac_t)),
           y1 = pmin(1, pmax(0, 0.4 + 0.2 * experience + city_effect
                                  + direct_effect - interference * frac_t))) |>
    ungroup() |>
    mutate(y_obs = rbinom(n(), 1, prob = if_else(notification == 1, y1, y0)))
  mean(d$y_obs[d$notification == 1]) - mean(d$y_obs[d$notification == 0])
}

sim_cluster <- function() {
  arms <- tibble(city_id = 1:n_cities,
                 notification = sample(rep(c(0, 1), each = n_cities / 2)))
  d <- market |> left_join(arms, by = "city_id") |>
    group_by(city_id) |>
    mutate(frac_t = mean(notification)) |>  # 0 or 1 within each cluster
    ungroup() |>
    mutate(
      # In a 100% treated city: every driver gets +direct_effect AND -interference (full saturation).
      # In a 0% treated city: no notification, no competition.
      y0 = pmin(1, pmax(0, 0.4 + 0.2 * experience + city_effect
                             - interference * frac_t)),
      y1 = pmin(1, pmax(0, 0.4 + 0.2 * experience + city_effect
                             + direct_effect - interference * frac_t)),
      y_obs = rbinom(n(), 1, prob = if_else(notification == 1, y1, y0)))
  mean(d$y_obs[d$notification == 1]) - mean(d$y_obs[d$notification == 0])
}

indiv_ests <- replicate(200, sim_individual())
cluster_ests <- replicate(200, sim_cluster())

bind_rows(
  tibble(est = indiv_ests, design = "Individual (biased, low variance)"),
  tibble(est = cluster_ests, design = "City cluster (unbiased, high variance)")
) |>
  ggplot(aes(est, fill = design)) +
  geom_histogram(bins = 40, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = policy_effect, color = "firebrick",
             linewidth = 1.2, linetype = "dashed") +
  geom_vline(xintercept = direct_effect, color = "grey40",
             linewidth = 0.8, linetype = "dotted") +
  annotate("text", x = policy_effect, y = Inf, vjust = 1.5, hjust = 1.05,
           label = paste0("True (policy) ATE = ", policy_effect),
           color = "firebrick", size = 3.8) +
  annotate("text", x = direct_effect, y = Inf, vjust = 1.5, hjust = 1.05,
           label = paste0("Direct effect = ", direct_effect),
           color = "grey40", size = 3.8) +
  scale_fill_manual(values = c("steelblue", "darkorange")) +
  labs(title = "Two designs, two estimands",
       subtitle = "Cluster recovers the policy effect; individual recovers the direct effect",
       x = "Estimated ATE", y = "Count") +
  theme(legend.position = "top")
```

<a href="#dgp-indiv-cluster" class="nav-btn">DGP code</a>

---
name: cluster-bias-main

# Two Estimators, Two Estimands

.pull-left[
.small[
**Individual rand → direct effect:**
- 50/50 within every city, so frac_t ≈ 0.5 everywhere
- Both arms face the *same* penalty $-\lambda \cdot 0.5$; it cancels
- Naive recovers $\tau$ = +0.05
- Answers: *"marginal benefit to one driver?"*

**City-cluster rand → policy effect:**
- Treated cities: frac_t = 1 (full penalty $-\lambda$)
- Control cities: frac_t = 0 (no penalty)
- Cluster contrast = $\tau - \lambda$ = −0.02
- Answers: *"what happens at full rollout?"*

Both are unbiased — for *different estimands*. Pick one based on the decision you're trying to make.
]
]

.pull-right[
```{r cluster-logic, echo=FALSE, fig.height=5}
tibble(
  Design = c("Individual (50/50)", "Cluster (saturated)"),
  `Treated mean` = c("b + τ − λ·½", "b + τ − λ"),
  `Control mean` = c("b − λ·½", "b"),
  `Estimator gives` = c("τ  (direct, +0.05)", "τ − λ  (policy, −0.02)")
) |>
  knitr::kable(align = "llll")
```

.highlight-box[
**Intuition:** the cluster contrast *includes* the interference term, so it captures the equilibrium effect under full rollout. Individual rand mechanically cancels the term out.
]
]

.small[**Cost of clustering:** effective sample size collapses from **200,000** drivers (individual) to **40** cities (city cluster) to **8** regions (region cluster). The next slide quantifies this with the design effect.]

<a href="#cluster-bias-proof" class="nav-btn">decomposition</a>

---

# Design Effect and ICC

The **design effect** is the variance inflation factor from clustering:

$$DEFF = 1 + (m - 1) \cdot \rho \qquad\Longrightarrow\qquad n_{eff} = \frac{n}{DEFF}$$

For a fixed effect size you want to detect, required $n$ scales by $DEFF$ — equivalently, your effective sample shrinks by the same factor. $m$ is cluster size (drivers per city), $\rho$ is intra-cluster correlation (ICC).

--

.pull-left[
.small[
```{r icc-table, echo=FALSE}
# Power for two-arm test, 50/50 split, alpha = 0.05 (two-sided):
#   SE(τ̂) = sigma * sqrt(4 / n_eff);  power = Φ(MDE / SE - z_{α/2})
# sigma = sqrt(p(1-p)) at baseline accept rate p = 0.4
sigma    <- sqrt(0.4 * 0.6)
mde      <- 0.15                      # target effect size: 15 pp
z_alpha2 <- qnorm(0.975)

expand_grid(
  m = c(100, 5000, 25000),            # sub-city zone, one city, 5-city region
  rho = c(0.02, 0.05, 0.10)           # marketplace-plausible ICCs
) |>
  mutate(
    design_effect = 1 + (m - 1) * rho,
    n_eff_num = 200000 / design_effect,
    n_eff = format(round(n_eff_num), big.mark = ","),
    se    = sigma * sqrt(4 / n_eff_num),
    power = paste0(round(100 * pnorm(mde / se - z_alpha2)), "%")
  ) |>
  select(`m (drivers)` = m, `ICC` = rho,
         `DEFF` = design_effect,
         `Eff. N (of 200k)` = n_eff,
         `Power (MDE = 15 pp)` = power) |>
  knitr::kable(digits = c(0, 2, 1, 0, 0), align = "rrrrr",
               format.args = list(big.mark = ","))
```
]
]

.pull-right[
.small[
**Plausible ranges for $\rho$:**

| Setting | Typical $\rho$ |
|---------|----------------|
| Online behavior (clicks, conversions) | 0.001 – 0.01 |
| Marketplace metrics (accept rate, ride completion) | 0.02 – 0.10 |
| Geo clusters (DMA, city) | 0.05 – 0.20 |
| Strong common shocks (weather, surge) | 0.10 – 0.30 |

**Rule of thumb:** anything you'd model with city or time-period fixed effects has $\rho$ big enough to matter. Estimate $\rho$ from a pilot or historical data — don't guess.
]
]

<a href="#icc-derivation" class="nav-btn">ICC derivation</a>

---

# The Bias-Variance Tradeoff: Simulated

```{r tradeoff-sim, echo=FALSE, fig.height=3.2, fig.width=10}
sim_cluster_size <- function(cluster_size) {
  n_clusters <- n_total / cluster_size
  arms <- tibble(cluster_id = 1:n_clusters,
                 notification = sample(rep(c(0, 1), each = n_clusters / 2)))
  d <- market |>
    mutate(cluster_id = ceiling(driver_id / cluster_size)) |>  # contiguous chunks
    left_join(arms, by = "cluster_id") |>
    group_by(city_id) |>                                # interference is within-city
    mutate(frac_t = mean(notification)) |>
    ungroup() |>
    mutate(y0 = pmin(1, pmax(0, 0.4 + 0.2 * experience + city_effect
                                  - interference * frac_t)),
           y1 = pmin(1, pmax(0, 0.4 + 0.2 * experience + city_effect
                                  + direct_effect - interference * frac_t)),
           y_obs = rbinom(n(), 1, prob = if_else(notification == 1, y1, y0)))
  mean(d$y_obs[d$notification == 1]) - mean(d$y_obs[d$notification == 0])
}

# Cluster size in drivers: from n_city/100 (50) to 10 * n_city (50,000)
cluster_sizes <- c(50, 250, 1000, 2500, 5000, 10000, 25000, 50000)
tradeoff <- map_dfr(cluster_sizes, function(cs) {
  ests <- replicate(150, sim_cluster_size(cs))
  tibble(cluster_size = cs,
         bias = abs(mean(ests) - true_ate),
         variance = var(ests))
})

tradeoff |>
  pivot_longer(c(bias, variance), names_to = "metric", values_to = "value") |>
  mutate(metric = factor(metric, levels = c("bias", "variance"),
                         labels = c("|Bias|", "Variance"))) |>
  ggplot(aes(cluster_size, value, color = metric)) +
  geom_vline(xintercept = drivers_per_city, linetype = "dashed",
             color = "grey60") +
  annotate("text", x = drivers_per_city, y = Inf, vjust = 1.5, hjust = -0.05,
           label = "1 city", color = "grey40", size = 3.6) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(values = c("firebrick", "steelblue")) +
  scale_x_log10(breaks = c(50, 250, 1000, 5000, 25000, 50000),
                labels = c("50", "250", "1k", "5k", "25k", "50k")) +
  facet_wrap(~ metric, scales = "free_y") +
  labs(title = "Bias drops at the city scale; variance keeps climbing",
       x = "Cluster size (drivers, log scale)", y = NULL, color = "") +
  theme(legend.position = "none")
```

.small[
**Bias** vanishes once clusters reach city scale — interference is fully contained. **Variance** keeps climbing — fewer effective independent units.
]

---

# Switchback Designs

An alternative to spatial clustering: **randomize over time**.

--

.pull-left[
**How it works:**
1. Divide time into periods (e.g., 2-hour blocks)
2. Each period, randomize each city to notification on/off
3. Compare accept rates in notified vs. non-notified periods
4. Every city serves as its own control (removes city fixed effects)
]

--

.pull-right[
```{r switchback-diagram, echo=FALSE, fig.height=4}
set.seed(123)
sb <- expand_grid(city = paste("City", 1:6), period = 1:12) |>
  mutate(notification = sample(c(0, 1), n(), replace = TRUE))

ggplot(sb, aes(period, city, fill = factor(notification))) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_manual(values = c("grey80", "steelblue"),
                    labels = c("Off", "On")) +
  scale_x_continuous(breaks = 1:12) +
  labs(title = "Switchback design: 6 cities x 12 periods",
       x = "Time period", y = "", fill = "Notification") +
  theme(legend.position = "top", panel.grid = element_blank())
```
]

---
name: switchback-strengths

# Switchback: Strengths and Risks

.small[
**Strengths:**
- Every unit (city) appears in both arms over time — removes unit-level confounders
- **More "independent observations" than pure clustering:** identification is *within-city* (a city's "on" periods vs. its own "off" periods), not between cities. The ICC penalty is differenced out, and effective $N$ grows from $\approx n_{cities}$ toward $n_{cities} \times n_{periods}$ <a href="#switchback-effn" class="inline-btn">proof</a>
- Natural for marketplace experiments where interference is spatial, not temporal
]

--

**Risks:**

```{r carryover, echo=FALSE, fig.height=2.2}
tibble(
  period = 1:10,
  treatment = c(1, 1, 0, 0, 1, 0, 1, 1, 0, 0),
  true_outcome = c(0.50, 0.52, 0.40, 0.41, 0.51, 0.40, 0.50, 0.52, 0.40, 0.41),
  with_carryover = c(0.50, 0.52, 0.46, 0.42, 0.51, 0.45, 0.50, 0.52, 0.46, 0.42)
) |>
  pivot_longer(c(true_outcome, with_carryover), names_to = "type", values_to = "y") |>
  mutate(type = if_else(type == "true_outcome", "No carryover", "With carryover")) |>
  ggplot(aes(period, y, color = type)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  geom_tile(data = tibble(period = 1:10,
                          treatment = c(1,1,0,0,1,0,1,1,0,0)),
            aes(period, y = 0.36, fill = factor(treatment)),
            height = 0.02, inherit.aes = FALSE, alpha = 0.5) +
  scale_color_manual(values = c("steelblue", "darkorange")) +
  scale_fill_manual(values = c("grey80", "steelblue"), guide = "none") +
  labs(title = "Carryover effect: notification bleeds into off periods",
       x = "Period", y = "Accept rate", color = "") +
  theme(legend.position = "top")
```

.small[
**Carryover:** notification in period $t$ affects behavior in $t+1$ even if $t+1$ is "off" (drivers who learned the zone keep going). **Fix:** insert *washout windows* — short buffers right after each switch that you exclude from analysis, letting behavior reset before measurement resumes. **Cost:** fewer usable minutes per period.
]

---

# Switchback: Simulation

```{r switchback-sim, echo=FALSE, fig.height=3.6}
# True within-city effect for the switchback DGP
tau_sb <- 0.05

sim_switchback <- function(carryover = 0) {
  n_periods <- 50; n_sw_cities <- 20
  d <- expand_grid(city_id = 1:n_sw_cities, period = 1:n_periods) |>
    mutate(notification = sample(c(0, 1), n(), replace = TRUE),
           city_effect = rep(rnorm(n_sw_cities, 0, 0.05), times = n_periods),
           period_effect = rep(rnorm(n_periods, 0, 0.02), each = n_sw_cities)) |>
    group_by(city_id) |>
    mutate(prev = lag(notification, default = 0),
           # Asymmetric carryover: residual treatment effect leaks into the
           # FOLLOWING off period only (drivers keep going to the zone they learned).
           carry = carryover * (1 - notification) * prev) |>
    ungroup() |>
    mutate(y = 0.4 + city_effect + period_effect +
               tau_sb * notification + carry + rnorm(n(), 0, 0.02))
  mean(d$y[d$notification == 1]) - mean(d$y[d$notification == 0])
}

sb0 <- replicate(500, sim_switchback(0))
sb1 <- replicate(500, sim_switchback(0.05))

bind_rows(
  tibble(est = sb0, scenario = "No carryover"),
  tibble(est = sb1, scenario = "Carryover = 0.05")
) |>
  ggplot(aes(est, fill = scenario)) +
  geom_histogram(bins = 35, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = tau_sb, color = "firebrick",
             linewidth = 1, linetype = "dashed") +
  annotate("text", x = tau_sb, y = Inf, vjust = 1.5, hjust = 1.05,
           label = paste0("True effect = ", tau_sb),
           color = "firebrick", size = 3.6) +
  scale_fill_manual(values = c("darkorange", "steelblue")) +
  labs(title = "Switchback estimates: asymmetric carryover biases the contrast downward",
       subtitle = "Off periods inherit residual treatment → control mean is inflated → estimate shrinks",
       x = "Estimated ATE", y = "Count", fill = "") +
  theme(legend.position = "top")
```

<a href="#switchback-dgp" class="nav-btn">DGP code</a>

---

# Two-Sided Marketplace Designs

In a ride-sharing experiment, you can randomize **riders** instead of **drivers**:

--

.pull-left[
**Randomize riders:**
- 50% of riders see new pricing or matching
- 50% see the old experience
- Riders don't interact with each other (mostly)
- No interference on the randomization side

**Measure driver outcomes:**
- Driver earnings, wait times, accept rate
- These reflect the equilibrium effect
]

--

.pull-right[
**Why it works:**
- Each rider's experience is independent (they don't share rides)
- But their collective demand changes driver supply allocation
- You get the direct effect on riders AND the indirect effect on drivers

**When it doesn't work:**
- If treated riders change market-wide conditions (e.g., surge pricing triggers)
- If the treatment fraction is large enough to move equilibrium prices
]

---

# Geo Experiments

.pull-left[
.small[
For online advertising and pricing experiments, **geographic units** are natural clusters.

A **DMA** (Designated Market Area, defined by Nielsen) is a metro region used for TV and ad targeting — e.g., "New York DMA," "Los Angeles DMA." There are **210 DMAs in the US**, ranging from a few hundred thousand to ~20M people. Ad spend can be served at DMA granularity.

**Treatment** in a geo experiment = changing ad spend in randomly chosen DMAs (typically *pausing* or *boosting* a campaign), holding control DMAs at the status quo. Outcome = revenue / conversions / signups per DMA. The contrast estimates **incremental lift** — what wouldn't have happened organically.
]
]

.pull-right[
```{r geo-diagram, echo=FALSE, fig.height=4.2, fig.width=6}
set.seed(42)
geos <- tibble(
  dma = paste("DMA", 1:20),
  treated = sample(rep(c(0, 1), each = 10)),
  baseline = runif(20, 1000, 5000),
  lift = if_else(treated == 1, 0.05, 0),
  revenue = baseline * (1 + lift) + rnorm(20, 0, 100)
)

ggplot(geos, aes(reorder(dma, revenue), revenue,
                 fill = factor(treated))) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = c("grey70", "steelblue"),
                    labels = c("Control (status-quo spend)",
                               "Treated (paused or boosted)")) +
  coord_flip() +
  labs(title = "20 DMAs: 10 treated, 10 control",
       subtitle = "Each DMA is an independent unit",
       x = NULL, y = "Revenue ($)", fill = "") +
  theme(legend.position = "top",
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```
]

.blue-box[
**Google/Meta approach:** "GeoX" / "geo-based incrementality." Randomize ad spend across 50–200 DMAs. Power depends on between-DMA variance and number of DMAs, not number of users.
]

---

# Ego-Cluster Randomization for Networks

For **network interference** (the author-nudge experiment from Module 2), standard clustering is hard because network boundaries are unclear.

--

.small[
**Ego-cluster approach** (node roles in the chart below):
1. Pick a focal study — the **ego** (center of each cluster, labeled "ego")
2. Add every study sharing co-authors with it — the **alters** (the four nodes around each ego)
3. Randomize the entire ego-cluster (ego + alters) to the same arm
4. Within-cluster spillover is absorbed; **bridges** (faded grey rings) are alters whose ties extend into the other arm — **drop them from the analysis sample** to avoid cross-arm contamination
]

--

```{r ego-cluster, echo=FALSE, fig.height=2.6}
# Each cluster is built from a focal "ego" (center) plus its alters (neighbors).
set.seed(42)
nodes <- tibble(
  id   = 1:12,
  x    = c(0, -1, -1, 1, 1,             # Cluster A: ego(1) at center, alters around
           5, 4, 4, 6, 6,               # Cluster B: ego(6) at center, alters around
           2.5, 2.5),                   # bridges between clusters
  y    = c(0, 1, -1, 1, -1,
           0, 1, -1, 1, -1,
           0.7, -0.7),
  role = c("Ego", rep("Alter", 4),
           "Ego", rep("Alter", 4),
           "Bridge", "Bridge"),
  cluster = c(rep("Cluster A (treated)", 5),
              rep("Cluster B (control)", 5),
              "Bridge (between clusters)", "Bridge (between clusters)")
)

edges <- tibble(
  from = c(1, 1, 1, 1, 2, 3, 2,         # within A: ego → alters + a few alter-alter
           6, 6, 6, 6, 7, 8, 9,         # within B: same structure
           4, 5, 11, 12),               # bridges to nearest alters
  to   = c(2, 3, 4, 5, 4, 5, 3,
           7, 8, 9, 10, 9, 10, 10,
           11, 12, 7, 8)
)

edge_df <- edges |>
  left_join(nodes |> select(id, x, y), by = c("from" = "id")) |>
  rename(x_from = x, y_from = y) |>
  left_join(nodes |> select(id, x, y), by = c("to" = "id")) |>
  rename(x_to = x, y_to = y)

ggplot() +
  geom_segment(data = edge_df,
               aes(x = x_from, y = y_from, xend = x_to, yend = y_to),
               color = "grey70", linewidth = 0.5) +
  geom_point(data = nodes,
             aes(x, y, color = cluster, size = role, shape = role,
                 alpha = role)) +
  geom_text(data = nodes |> filter(role == "Ego"),
            aes(x, y, label = "ego"),
            color = "white", size = 3.5, fontface = "bold") +
  scale_color_manual(
    values = c("Bridge (between clusters)" = "grey60",
               "Cluster A (treated)"       = "steelblue",
               "Cluster B (control)"       = "darkorange"),
    breaks = c("Cluster A (treated)",
               "Cluster B (control)",
               "Bridge (between clusters)"),
    labels = c("Cluster A (treated)",
               "Cluster B (control)",
               "<span style='color:#999'>Bridge (dropped from analysis)</span>")) +
  scale_size_manual(values  = c(Ego = 13, Alter = 6, Bridge = 5)) +
  scale_shape_manual(values = c(Ego = 16, Alter = 16, Bridge = 1)) +
  scale_alpha_manual(values = c(Ego = 1,  Alter = 1, Bridge = 0.45)) +
  guides(size  = "none", shape = "none", alpha = "none",
         color = guide_legend(override.aes = list(
           shape = c(16, 16, 1),       # solid for clusters, hollow ring for bridge
           alpha = c(1, 1, 0.45)
         ))) +
  labs(title = "Ego-cluster randomization: each cluster grows from a focal study (ego)",
       subtitle = "Bridges link to the other arm — typically dropped from the analysis sample to avoid cross-arm contamination",
       color = "") +
  theme_void(base_size = 13) +
  theme(legend.position = "top",
        legend.text = ggtext::element_markdown())
```

---

# Approach 4: Model the Spillover (Exposure Mapping)

The first three approaches **contain** interference by design. When you can't — diffuse networks, observational settings, small samples — **model the spillover** instead.

.pull-left[
.small[
**Idea:** for each unit $i$, build an exposure measure $S_i$ summarizing how its network neighbors are treated. Include $S_i$ as a covariate; recover the direct effect *and* the spillover jointly.

For study $i$ with author set $A_i$ (M2's author-nudge experiment), $S_{ik} = \sum_{a \in A_i} \sum_{s \neq i} M_{as} T_{sk}$ counts other studies in arm $k$ that share an author with $i$.

- $S_i = \sum_{k \geq 1} S_{ik}$ — exposure to *any* treatment arm
- $P_i = \sum_{k \geq 0} S_{ik}$ — total degree (incl. control)
]
]

.pull-right[
```{r spillover-diagram, echo=FALSE, fig.height=2.8, fig.width=5}
nodes_s <- tibble(
  id    = c("1", "2", "3", "4", "5"),
  arm   = c("T0", "T1", "T2", "T3", "out"),
  label = c("1, T0", "2, T1", "3, T2", "4, T3", "5"),
  x     = c(0, 2, 4, 2, 6),
  y     = c(0, 0.5, 0, -1.1, -0.3)
)
edges_s <- tibble(
  from = c("1", "2", "3", "4"),
  to   = c("2", "3", "5", "5"),
  type = c("solid", "solid", "dashed", "dashed")
) |>
  left_join(nodes_s |> select(id, x_from = x, y_from = y),
            by = c("from" = "id")) |>
  left_join(nodes_s |> select(id, x_to = x, y_to = y),
            by = c("to" = "id"))

ggplot() +
  geom_segment(data = edges_s,
               aes(x = x_from, y = y_from, xend = x_to, yend = y_to,
                   linetype = type), color = "grey50", linewidth = 0.6) +
  geom_vline(xintercept = 5, linetype = "dashed", color = "grey70") +
  annotate("text", x = 6, y = 1.0, label = "out of sample",
           color = "grey50", size = 3, hjust = 0.5) +
  geom_point(data = nodes_s, aes(x, y, color = arm), size = 16) +
  geom_text(data = nodes_s, aes(x, y, label = label),
            color = "white", size = 2.8, fontface = "bold") +
  scale_linetype_identity() +
  scale_color_manual(values = c(T0 = "grey50", T1 = "steelblue",
                                T2 = "darkorange", T3 = "forestgreen",
                                out = "grey80")) +
  guides(color = "none") +
  coord_cartesian(xlim = c(-0.7, 7), ylim = c(-1.6, 1.3)) +
  theme_void(base_size = 12)
```
.small[
Study 2 shares author *b* with study 1 and *e* with study 3 ⇒ $S_2 = 1$, $P_2 = 2$. Study 4's only tie is out-of-sample ⇒ $S_4 = P_4 = 0$.
]
]

<a href="#spillover-regression" class="nav-btn">regression & validity</a>

---

# Approach 4: Simulation

True parameters: $\tau_1=0.05$, $\tau_2=0.10$, $\tau_3=0.15$ (direct), $\beta_0=0.05$ (control spillover), $\theta=0.02$ (extra on treated), $\psi=0.05$ (direct effect of degree $P_i$). 500 studies, co-author network (mean degree ≈ 6), random T0/T1/T2/T3 assignment.

```{r spillover-sim, echo=FALSE, fig.height=3.5}
set.seed(42)
n <- 500
arm <- sample(0:3, n, replace = TRUE)
treated <- as.numeric(arm > 0)

# Co-author network (Erdős–Rényi, mean degree ~ 6)
net <- matrix(rbinom(n^2, 1, 6 / n), n, n)
net <- pmin(net + t(net), 1); diag(net) <- 0

# Exposure to each arm + total degree
S_arm   <- net %*% sapply(0:3, function(k) as.numeric(arm == k))
S_total <- as.numeric(rowSums(S_arm[, 2:4]))    # exposure to T arms only
P_total <- as.numeric(rowSums(S_arm))           # total degree (includes T0)

# True parameters: include a direct degree effect ψ to make P_i matter
tau_true   <- c(0.05, 0.10, 0.15)
beta_true  <- 0.05
theta_true <- 0.02
psi_true   <- 0.05
y <- 0.4 + (arm == 1) * tau_true[1] +
           (arm == 2) * tau_true[2] +
           (arm == 3) * tau_true[3] +
     beta_true * S_total + theta_true * S_total * treated +
     psi_true * P_total +
     rnorm(n, sd = 0.15)

T1 <- as.numeric(arm == 1); T2 <- as.numeric(arm == 2); T3 <- as.numeric(arm == 3)
S_t <- S_total * treated

# Two specifications: drop P_i (omitted control) vs. include P_i (correct)
fit_no_P  <- lm(y ~ T1 + T2 + T3 + S_total + S_t)
fit_yes_P <- lm(y ~ T1 + T2 + T3 + S_total + S_t + P_total)

truth_df <- tibble(
  term  = c("T1", "T2", "T3", "S_total", "S_t"),
  truth = c(0.05, 0.10, 0.15, 0.05, 0.02))

coefs <- bind_rows(
  broom::tidy(fit_no_P,  conf.int = TRUE) |> mutate(spec = "Omits P_i (biased)"),
  broom::tidy(fit_yes_P, conf.int = TRUE) |> mutate(spec = "Includes P_i (correct)")
) |>
  inner_join(truth_df, by = "term") |>
  mutate(term = factor(term,
                       levels = c("T1", "T2", "T3", "S_total", "S_t"),
                       labels = c("τ₁ (T1 direct)", "τ₂ (T2 direct)",
                                  "τ₃ (T3 direct)",
                                  "β₀ (control spillover)",
                                  "θ (T × spillover)")))

ggplot(coefs, aes(estimate, term, color = spec)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey85") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high),
                  size = 0.5, position = position_dodge(0.6)) +
  geom_point(aes(x = truth, y = term), color = "firebrick", shape = 4,
             size = 4, stroke = 1.4, inherit.aes = FALSE,
             data = distinct(coefs, term, truth)) +
  scale_color_manual(values = c("Omits P_i (biased)"     = "grey50",
                                 "Includes P_i (correct)" = "steelblue")) +
  labs(title = "Why P_i matters: omitting it confounds spillover with degree",
       subtitle = "Red ✕: truth. With ψ = 0.05 in the DGP, β₀ is biased upward when P_i is dropped.",
       x = "Coefficient (95% CI)", y = NULL, color = "") +
  theme(legend.position = "top")
```

<a href="#spillover-regression" class="nav-btn">regression & validity</a>

---

# Application: City-Level Zone-Notification Test

A ride-sharing company wants to test whether the **zone-notification** feature lifts driver accept rate at the *city* level.

--

.pull-left[
**Why cluster at the city level?**
- Within a city, notified drivers compete with non-notified for rides
- A 50/50 split per city contaminates control's accept rate
- City-level randomization: each city gets the feature on or off, no within-city mixing
]

--

.pull-right[
**The power problem:**
- Only 40 cities available
- 20 treatment, 20 control
- ICC of accept rate within cities ≈ 0.10
- DEFF $\approx 500 \;\Longrightarrow\;$ Effective $N \approx 400$ (out of 200k drivers)
- For MDE = 15 pp, power $\approx 86\%$ — need a **large** effect
]

--

.highlight-box[
**The interview question:** "We have 40 cities. Can we run a cluster-randomized experiment?" Answer: probably, but only if the expected effect is large. Calculate the minimum detectable effect (MDE) and check if it's business-relevant.
]

---

# Application: Switchback for Zone Notifications

**Alternative:** use a switchback design within each city.

--

- Divide each day into 2-hour blocks
- Randomize which blocks have the zone-notification feature on vs. off
- 12 blocks/day × 30 days × 40 cities = 14,400 city-period cells
- Effective N is much larger than 40 clusters

--

**But:**
- **Carryover**: if drivers learn the high-demand zone during "on" periods, they keep going there during "off" periods
- Need washout periods (waste time and data)
- Confounding with time-of-day effects (notifications matter more at peak times)

--

.blue-box[
**Practical design:** stratified switchback. Within each city, pair "similar" time blocks (e.g., Monday 5–7pm with Wednesday 5–7pm) and randomize within pairs. This controls for time-of-day and day-of-week effects.
]

---

# Application: Geo Incrementality for Online Ads

An e-commerce company wants to measure whether their TV ad campaign drives sales.

--

**Design:**
- 100 DMAs (Designated Market Areas) across the US
- Randomly assign 50 DMAs to see ads, 50 to no ads
- Measure sales lift at the DMA level
- Control for baseline sales using **CUPED** — subtract a scaled pre-experiment outcome from each DMA's post outcome to soak up pre-existing variation (covered in M5)

--

**Challenges:**
- DMAs vary enormously in size (NYC vs. rural Montana)
- Need to weight by population or stratify by DMA size
- Spillovers: people in control DMAs can see ads on streaming/social media (imperfect compliance)
- Attribution: sales in a DMA might come from visitors from neighboring DMAs

--

.highlight-box[
**Key insight:** geo experiments work best when the treatment is *geographically contained* (TV, billboards, local promotions). They work poorly for digital treatments that cross geo boundaries (social media ads, viral content).
]

---

# Choosing the Right Design

| Interference type | Recommended design | Key tradeoff |
|-------------------|-------------------|--------------|
| Within-market (supply/demand) | City-level cluster | Bias vs. power |
| Temporal (pricing, algorithms) | Switchback | Carryover vs. efficiency |
| Network (social, co-authorship) | Ego-cluster | Cluster size vs. coverage |
| Geographic (ads, promotions) | Geo experiment | Geo count vs. precision |
| Two-sided marketplace | Randomize one side | Indirect effects only |

--

**Decision framework:**

1. **Where does interference happen?** Within cities? Across time? Through networks?
2. **Can you cluster at that level?** How many independent clusters do you have?
3. **Is the effect large enough to detect with that many clusters?**
4. **Is there carryover/leakage between clusters?** If so, switchback may be risky.

---

# Summary of the Bias-Variance Tradeoff

```{r final-tradeoff, echo=FALSE, fig.height=4}
tradeoff_summary <- tibble(
  design = c("Individual\n(driver-level)", "Zone\n(sub-city)",
             "City cluster", "Region\n(5 cities)", "Full cluster\n(20 cities)"),
  bias = c(0.070, 0.030, 0.003, 0.002, 0.001),
  variance = c(0.005, 0.015, 0.030, 0.050, 0.090),
  order = 1:5
) |>
  mutate(rmse = sqrt(bias^2 + variance^2))

ggplot(tradeoff_summary, aes(variance, bias, label = design)) +
  geom_point(aes(size = rmse), color = "steelblue", alpha = 0.7) +
  geom_text(vjust = -1.2, size = 3.5) +
  scale_size_continuous(range = c(3, 10), guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.25))) +
  labs(title = "Every design sits on the bias-variance frontier",
       subtitle = "Targeting the policy effect: individual rand overstates by λ; cluster recovers it.",
       x = "Variance (fewer independent units)",
       y = "Bias (overstates policy effect)") +
  theme(panel.grid.minor = element_blank())
```

---

# Key Takeaways

1. **Cluster randomization** eliminates interference bias by assigning entire clusters to the same arm. Cost: fewer independent units = more variance.

2. **Switchback designs** randomize over time. Good when interference is spatial. Watch for carryover effects.

3. **Two-sided marketplace designs**: randomize one side, measure the other. Works when one side doesn't interact with itself.

4. **Geo experiments**: natural clusters for advertising. Need many geos (50+) and geographically contained treatments.

5. **Ego-cluster randomization**: for network interference (e.g., the author-nudge experiment). Cluster = focal unit + neighbors.

6. **The bias-variance tradeoff** is the central tension. Minimize **RMSE**, not just bias or variance alone.

7. The **design effect** formula $n_{eff} = n / [1 + (m-1)\rho]$ tells you how much power you lose from clustering.

---

# Exercise Preview

In the exercise you will:

1. Compare individual vs. city-level randomization in a zone-notification simulation
2. Show that clustering eliminates bias but increases variance
3. Sweep over cluster sizes and find the RMSE-minimizing level
4. Simulate a switchback design with and without carryover
5. Compute design effects for different ICC values

See `exercise.R` for the starter code.

---

# Course Map

<table>
<tr><th>#</th><th>Module</th><th>Status</th></tr>
<tr><td>1</td><td><a href="../module-01/slides.html">The Experimental Ideal</a></td><td>✓ done</td></tr>
<tr><td>2</td><td><a href="../module-02/slides.html">SUTVA and When It Breaks</a></td><td>✓ done</td></tr>
<tr><td><b>3</b></td><td><b>Designing Around Interference</b> <i>(just finished)</i></td><td>✓ done</td></tr>
<tr><td>4</td><td>Power and Sample Size</td><td>up next</td></tr>
<tr><td>5</td><td><a href="../module-05/slides.html">Analyzing Experiments</a></td><td>✓ done</td></tr>
<tr><td>6</td><td>Multiple Testing & Subgroups</td><td>upcoming</td></tr>
<tr><td>7</td><td><a href="../module-07/slides.html">External Validity</a></td><td>✓ done</td></tr>
<tr><td>8</td><td><a href="../module-08/slides.html">Beyond the A/B Test</a></td><td>✓ done</td></tr>
</table>

---
class: center, middle, inverse

# Backup Slides

---
name: cluster-bias-proof

# Backup: Two Estimators, Two Estimands

**Symmetric interference DGP** — every driver in a city competes against the same treated share $s$:

$$y_i^{(0)} = b_i - \lambda \cdot s_{c(i)}, \qquad y_i^{(1)} = b_i + \tau - \lambda \cdot s_{c(i)}$$

with $b_i = 0.4 + 0.2\,\text{exp}_i + \alpha_{c(i)}$, $\tau = 0.05$ direct effect, $\lambda = 0.07$ interference.

--

The naive contrast under random assignment — $\bar b$ balances across arms:

$$\hat{\tau}_{\text{naive}} = E[y^{(1)} \mid D=1] - E[y^{(0)} \mid D=0] = (\bar b + \tau - \lambda \bar s_{D=1}) - (\bar b - \lambda \bar s_{D=0}) = \tau - \lambda(\bar s_{D=1} - \bar s_{D=0})$$

--

**Individual rand** (50/50 in every city → $\bar s_{D=1} = \bar s_{D=0} = 0.5$):

$$\hat\tau_{\text{ind}} = \tau - \lambda(0.5 - 0.5) = \tau = 0.05 \quad \text{(direct effect)}$$

**City cluster** (treated cities $s = 1$, control cities $s = 0$):

$$\hat\tau_{\text{cluster}} = \tau - \lambda(1 - 0) = \tau - \lambda = -0.02 \quad \text{(policy effect)}$$

--

Both are unbiased — for *different* estimands. If your decision is "should we roll this out everywhere?", you want $\hat\tau_{\text{cluster}}$.

<a href="#cluster-bias-main" class="nav-btn">← back</a>

---
name: dgp-indiv-cluster

# Backup: DGP for the Slide 5 Simulation

.small[
```{r dgp-display, eval=FALSE, echo=TRUE}
direct_effect <- 0.05   # per-driver direct boost from being notified
interference  <- 0.07   # symmetric crowd-out per share treated (hits BOTH arms)

# Individual rand: 50/50 in every city → frac_t ≈ 0.5 everywhere
sim_individual <- function() {
  d <- market |> mutate(notification = sample(rep(c(0, 1), each = n() / 2)))
  d |> group_by(city_id) |>
    mutate(frac_t = mean(notification),                                          # ≈ 0.5
           y0 = pmin(1, pmax(0, 0.4 + 0.2 * experience + city_effect             # baseline
                                  - interference * frac_t)),                     # CONTROLS hit by competition
           y1 = pmin(1, pmax(0, 0.4 + 0.2 * experience + city_effect             # baseline
                                  + direct_effect - interference * frac_t))) |>  # TREATED also hit
    ungroup() |>
    mutate(y_obs = rbinom(n(), 1, prob = if_else(notification == 1, y1, y0))) |>
    summarise(ate = mean(y_obs[notification == 1]) - mean(y_obs[notification == 0]))
}

# City-cluster rand: every city is 100% same arm → frac_t is 0 or 1
sim_cluster <- function() {
  arms <- tibble(city_id = 1:n_cities,
                 notification = sample(rep(c(0, 1), each = n_cities / 2)))
  market |> left_join(arms, by = "city_id") |>
    group_by(city_id) |> mutate(frac_t = mean(notification)) |> ungroup() |>
    mutate(y0 = pmin(1, pmax(0, 0.4 + 0.2 * experience + city_effect
                                  - interference * frac_t)),                     # 0 in control cities
           y1 = pmin(1, pmax(0, 0.4 + 0.2 * experience + city_effect
                                  + direct_effect - interference * frac_t)),     # full λ in treated cities
           y_obs = rbinom(n(), 1, prob = if_else(notification == 1, y1, y0))) |>
    summarise(ate = mean(y_obs[notification == 1]) - mean(y_obs[notification == 0]))
}
```
]

.small[
- **Individual:** penalty cancels (both arms get $-\lambda \cdot 0.5$) → naive = $\tau$ = +0.05 (direct).
- **Cluster:** treated cities take full $-\lambda$, control cities take none → ATE = $\tau - \lambda$ = −0.02 (policy).
]

<a href="#indiv-cluster-main" class="nav-btn">← back</a>

---
name: icc-derivation

# Backup: ICC and Design Effect Derivation

.small[
**ICC:** $\rho = \dfrac{\sigma^2_b}{\sigma^2_b + \sigma^2_w}$, with $\sigma^2_b$ = between-cluster variance, $\sigma^2_w$ = within-cluster variance.

For a cluster of size $m$ and $K$ clusters total ( $n = Km$ ):

$$\text{Var}(\bar{Y}_c) = \frac{\sigma^2_w}{m} + \sigma^2_b
\qquad\Longrightarrow\qquad
\text{Var}(\bar{Y}) = \frac{\sigma^2_w}{Km}\bigl[1 + (m-1)\rho\bigr]$$

Under simple random sampling, $\text{Var}_{SRS}(\bar{Y}) = \sigma^2 / n$. The ratio is the **design effect**:

$$\text{DEFF} = \frac{\text{Var}(\bar{Y})}{\text{Var}_{SRS}(\bar{Y})} = 1 + (m-1)\rho
\qquad\Longrightarrow\qquad
n_{eff} = \frac{n}{\text{DEFF}}$$
]

---
name: switchback-dgp

# Backup: DGP for the Switchback Simulation

.small[
```{r switchback-dgp-display, eval=FALSE, echo=TRUE}
tau_sb <- 0.05   # true within-city treatment effect

sim_switchback <- function(carryover = 0) {
  n_periods <- 50; n_sw_cities <- 20
  d <- expand_grid(city_id = 1:n_sw_cities, period = 1:n_periods) |>
    mutate(notification   = sample(c(0, 1), n(), replace = TRUE),       # i.i.d. switching
           city_effect    = rep(rnorm(n_sw_cities, 0, 0.05),             # μ_c
                                times = n_periods),
           period_effect  = rep(rnorm(n_periods,  0, 0.02),              # γ_t
                                each  = n_sw_cities)) |>
    group_by(city_id) |>
    mutate(prev  = lag(notification, default = 0),
           # Asymmetric carryover: residual treatment leaks into off periods ONLY.
           # On periods get the full direct effect regardless of prev.
           carry = carryover * (1 - notification) * prev) |>
    ungroup() |>
    mutate(y = 0.4 + city_effect + period_effect +
               tau_sb * notification + carry + rnorm(n(), 0, 0.02))
  mean(d$y[d$notification == 1]) - mean(d$y[d$notification == 0])
}
```
]

.small[
**Why the bias is downward.** With i.i.d. switching, $P(\text{prev}=1) = 0.5$ in both arms — symmetric carryover (a $+\delta$ that hits *any* off period equally) would cancel. But here the carryover is *asymmetric*: it shows up only on $T=0$ periods that follow $T=1$, lifting the control mean by $\delta/2$ while leaving the treated mean unchanged. So $E[\hat{\tau}] = \tau - \delta/2$. With $\tau = 0.05$ and $\delta = 0.05$: estimate ≈ 0.025.
]

<a href="#switchback-formal" class="nav-btn">formal</a>

---
name: switchback-formal

# Backup: Switchback Estimator

For a switchback design with cities $c = 1, \ldots, C$ and periods $t = 1, \ldots, T$:

$$Y_{ct} = \mu_c + \gamma_t + \tau \cdot D_{ct} + \delta \cdot D_{c,t-1} + \varepsilon_{ct}$$

--

- $\mu_c$: city fixed effect
- $\gamma_t$: period fixed effect
- $D_{ct}$: notification indicator for city $c$ at time $t$
- $\delta$: carryover parameter (ideally $\delta = 0$)

--

The standard switchback estimator ignores carryover:

$$\hat{\tau} = \frac{1}{CT}\sum_{c,t} (2D_{ct} - 1) Y_{ct}$$

--

If $\delta \neq 0$, the estimator is biased:

$$E[\hat{\tau}] = \tau + \delta \cdot \text{Corr}(D_{ct}, D_{c,t-1})$$

Under i.i.d. randomization, $\text{Corr}(D_{ct}, D_{c,t-1}) \approx 0$, so the bias is small but not zero in finite samples. Including $D_{c,t-1}$ as a covariate or using burn-in periods mitigates this.

---
name: switchback-effn

# Backup: Why $n_{eff}$ Grows from $C$ to $C \cdot T$

.small[
Same model: $Y_{ct} = \mu_c + \gamma_t + \tau D_{ct} + \varepsilon_{ct}$, with $\varepsilon_{ct} \sim (0, \sigma_\varepsilon^2)$ i.i.d., $\mu_c \sim (0, \sigma_\mu^2)$, $\gamma_t \sim (0, \sigma_\gamma^2)$. Half-half assignment, $C$ cities, $T$ periods.

**Cluster randomization** (one assignment per city; average $T$ periods within each city):

$$\bar{Y}_c = \mu_c + \bar{\gamma} + \tau D_c + \bar{\varepsilon}_c
\quad\Longrightarrow\quad
\text{Var}(\bar{Y}_c) = \sigma_\mu^2 + \frac{\sigma_\gamma^2 + \sigma_\varepsilon^2}{T}$$

$$\text{Var}(\hat{\tau}_{cl}) = \frac{4}{C}\!\left(\sigma_\mu^2 + \frac{\sigma_\gamma^2 + \sigma_\varepsilon^2}{T}\right)
\;\;\xrightarrow{T \to \infty}\;\;
\frac{4 \sigma_\mu^2}{C}$$

The city term $\sigma_\mu^2$ does **not** average down with $T$ — it sets a noise floor governed by the city count $C$.

**Switchback** (i.i.d. re-randomization each period; FE estimator absorbs $\mu_c$ and $\gamma_t$):

$$\hat{\tau}_{sb} = \frac{\sum_{ct} \tilde{D}_{ct}\, Y_{ct}}{\sum_{ct} \tilde{D}_{ct}^2},\qquad
\tilde{D}_{ct} = D_{ct} - \bar{D}_c - \bar{D}_t + \bar{D}$$

$$\text{Var}(\hat{\tau}_{sb}) = \frac{\sigma_\varepsilon^2}{\sum_{ct} \tilde{D}_{ct}^2}
\approx \frac{\sigma_\varepsilon^2}{0.25 \cdot CT}
= \frac{4 \sigma_\varepsilon^2}{C \cdot T}$$

The within-city demeaning kills $\mu_c$, so noise is driven by the idiosyncratic $\sigma_\varepsilon^2$ over all $C \cdot T$ city-periods.

**Effective $n$** ( $n_{eff} = \sigma^2_{tot} / \text{Var}(\hat{\tau})$, scaled to SRS units):

$$n_{eff,\,cl} \;\approx\; \frac{\sigma^2_{tot}}{\sigma_\mu^2}\cdot\frac{C}{4} \;\sim\; O(C)
\qquad\text{vs.}\qquad
n_{eff,\,sb} \;\approx\; \frac{\sigma^2_{tot}}{\sigma_\varepsilon^2}\cdot\frac{CT}{4} \;\sim\; O(C \cdot T)$$

Switchback gains a factor of $T$ when $\sigma_\mu^2$ is non-trivial (high ICC). With carryover, $\sigma_\varepsilon^2$ is replaced by an inflated effective error and the ratio shrinks toward 1.
]

<a href="#switchback-strengths" class="nav-btn-br">← back</a>

---
name: spillover-regression

# Backup: The Spillover Regression

Augment the DiD specification with exposure terms:

.small[
$$Y_{it} = \mu_i + \delta\, I_{t=1} + \sum_{j=1}^{J}\tau_j\,(T_{ij}\, I_{t=1}) + \beta_0\,(S_i\, I_{t=1}) + \theta\,(S_i\, I_{t=1}\, \mathbb{1}\{T_i \neq 0\}) + \sum_{j=0}^{J}\psi_j\,(T_{ij}\, I_{t=1}\, P_i) + \varepsilon_{it}$$
]

.pull-left[
.small[
**What each term identifies:**
- $\tau_j$ — direct effect of own assignment to arm $j$
- $\beta_0$ — spillover onto **control** studies from exposure to any treatment
- $\theta = \beta_* - \beta_0$ — *additional* spillover when own arm is treated; sign tells you **complementarity** ( $\theta>0$ ) vs. **substitution** ( $\theta<0$ )
- $\psi_j P_i$ — controls for total network degree, so spillover ≠ "well-connected studies are different" <a href="#spillover-S-vs-P" class="inline-btn">S vs P</a>
]
]

.pull-right[
.small[
**Restrictions used to cut 16 $\beta_{jk}$ down to 2** (count: 16 → 12 → 4 → 2):
1. No spillover from control exposure: $\beta_{j0}=0$ for all $j$ ⇒ kills 4 ⇒ **12 left**
2. Spillover doesn't depend on *which* T-arm exposed (for **all** $j$): $\beta_{jk}=\beta_{j*}$ for $k\in\{1,2,3\}$ ⇒ collapses 3 entries per row ⇒ **4 left** ($\beta_{0*},\beta_{1*},\beta_{2*},\beta_{3*}$)
3. Spillover on T-arms is the same across T-arms: $\beta_{1*}=\beta_{2*}=\beta_{3*}\equiv\beta_*$ ⇒ **2 left** ($\beta_0\equiv\beta_{0*}$ and $\beta_*$)

Each restriction is testable: relax it, refit, compare $\widehat\beta_0$ and $\widehat\theta$.
]
]

<a href="#spillover-validity" class="nav-btn">validity →</a>

---
name: spillover-validity

# Backup: Validity of Exposure Mapping

.small[
**What's identified for free (under random assignment):**
- $T_{ij}$ is randomized ⇒ $\tau_j$ is identified unconditionally.
- $S_i$ is determined by *others'* (random) assignments + the **fixed** network structure ⇒ $S_i$ is "as good as random" *conditional on the network*. So $\beta_0$ is identified by within-arm variation in exposure. <a href="#spillover-id" class="inline-btn">proof</a>
]

.small[
**What you need on top:**
1. **Independent random assignment** of $T_i$ across units (or known stratified assignment with conditioning).
2. **Correct exposure mapping.** $S_i$ captures all relevant interference channels. If second-degree (indirect) ties or out-of-sample co-authors matter, $\widehat\beta_0$ absorbs them — possibly with the wrong sign.
3. **Linearity / additivity.** Spillover scales linearly in $S_i$ and degree enters through $P_i$ alone. Saturation, threshold, or non-monotone effects, or unmodeled $P_i \to Y$ channels, break this.
4. **Network exogeneity.** The co-author network must not respond to (anticipated) treatment (no endogenous network formation between $t=0$ and $t=1$).
5. **Stable composition of $P_i$.** $P_i$ must be measured the same way for everyone, so that controlling for it actually absorbs the "high-degree studies are different" channel. <a href="#spillover-S-vs-P" class="inline-btn">S vs P</a>
]

.highlight-box[
**Trade-off vs. ego-cluster.** Exposure mapping uses the *whole* sample (no nodes dropped) and recovers spillover magnitudes — but identification rests on a parametric model of how spillovers propagate. Ego-cluster guarantees clean identification on a smaller sample. The right choice depends on whether you trust the network model or the partition.
]

<a href="#spillover-regression" class="nav-btn">← regression</a>

---
name: spillover-id

# Backup: Why $\beta_0$ Is Identified

.small[
**Setup.** Treatment $T_i \in \{0, 1, \dots, J\}$ assigned independently across $i$. Network $\mathbf{M} = (M_{as})$ is fixed (measured pre-treatment). Define exposure as
$$S_{ik} = \sum_{a \in A_i} \sum_{s \neq i} M_{as}\, \mathbb{1}\{T_s = k\}, \qquad S_i = \sum_{k \geq 1} S_{ik}, \qquad P_i = \sum_{k \geq 0} S_{ik}.$$

Maintained outcome model:
$$Y_i = \mu + \tau_{T_i} + \beta_0\, S_i + \theta\, S_i\, \mathbb{1}\{T_i \neq 0\} + \psi\, P_i + \varepsilon_i,\quad E[\varepsilon_i \mid T_i, S_i, P_i, \mathbf{M}] = 0.$$
]

--

.small[
**Lemma (random exposure).** $\;S_i \perp T_i \mid \mathbf{M}.$

*Proof.* The inner sum runs over $s \neq i$, so $S_i$ is a function of $\{T_s : s \neq i\}$ and the fixed network $\mathbf{M}$ alone. Independent assignment gives $T_i \perp \{T_s : s \neq i\}$, and $\mathbf{M}$ is constant. Therefore $S_i \perp T_i \mid \mathbf{M}$. $\quad\square$

(Same argument applies to $P_i$: $P_i \perp T_i \mid \mathbf{M}$.)
]

--

.small[
**Identification of $\beta_0$.** Restrict to controls ( $T_i = 0$ ):
$$E[Y_i \mid T_i = 0, S_i, P_i, \mathbf{M}] \;=\; \mu + \tau_0 + \beta_0\, S_i + \psi\, P_i.$$

OLS of $Y$ on $S_i$ **and** $P_i$ within the control sub-sample is consistent for $\beta_0$ whenever the partial variance $\text{Var}(S_i \mid P_i, T_i = 0) > 0$ — guaranteed by the lemma plus a non-degenerate network ( <a href="#spillover-S-vs-P" class="inline-btn">S vs P</a> shows this partial variance equals $\pi_T(1-\pi_T)\, P_i$). Within-treated OLS gives $\beta_0 + \theta$, so $\theta$ is identified. $\quad\square$
]

--

.small[
**What this *doesn't* prove.** (i) **Linearity of $S_i$ and $P_i$** — nonlinear or threshold spillovers, indirect ties, or interactions break the result; OLS then recovers a best-linear-projection that is generally not the structural $\beta_0$. (ii) **Network exogeneity** — co-author networks must not respond to (anticipated) treatment. (iii) **Identifying variation may be small** — if $\rho(S, P) \approx 1$ (dense networks, concentrated treatment fractions), $\widehat{\beta_0}$ is consistent but high-variance.
]

<a href="#spillover-validity" class="nav-btn-br">← validity</a>

---
name: spillover-S-vs-P

# Backup: $S_i$ vs $P_i$ — Why Both, Despite Collinearity?

.small[
**Decomposition.** $P_i = S_{i0} + S_i$, where $S_{i0}$ counts neighbors in the control arm. So under uniform 4-arm randomization with treatment fraction $\pi_T = 3/4$:
$$E[S_i \mid P_i] = \pi_T \cdot P_i = 0.75\, P_i$$
The expected $S_i$ is a deterministic linear function of $P_i$ — that's the source of the collinearity. The correlation $\rho(S, P)$ is close to 1 in any sample with non-trivial $P_i$.
]

--

.pull-left[
.small[
**Where identifying variation lives.** Decompose $S_i$ into structural and randomized components:
$$S_i \;=\; \underbrace{\pi_T P_i}_{\text{structural (}\propto\text{degree)}} \;+\; \underbrace{(S_i - \pi_T P_i)}_{\text{randomized}}$$
Conditional on $P_i$, the second piece has variance
$$\text{Var}(S_i \mid P_i) = \pi_T(1-\pi_T)\, P_i \approx 0.19\, P_i$$
driven entirely by *which arm each neighbor landed in*. That's the variation $\widehat{\beta_0}$ uses.
]
]

.pull-right[
.small[
**Why include $P_i$ anyway?** Drop $P_i$ and $\widehat{\beta_0}$ confounds two channels:
1. **Per-T-neighbor spillover** — the structural $\beta_0$.
2. **Degree effect** — well-connected studies may differ for unrelated reasons (more visibility, broader networks, prior relationships, productivity).

Including $P_i$ partials out the degree channel; the residual in $S_i \mid P_i$ is *exactly* the randomized neighbor-arm draw, which is the only variation $\beta_0$ should be identified from.

**Cost:** collinearity inflates $\text{SE}(\widehat{\beta_0})$; power scales with $\sqrt{n\,\bar P\,\pi_T(1-\pi_T)}$. Identification requires both **many studies** *and* **non-trivial degree**.
]
]

<a href="#spillover-validity" class="nav-btn-br">← validity</a>

---
name: ego-cluster-formal

# Backup: Ego-Cluster Randomization

Define the **ego network** of unit $i$ as $\mathcal{N}_i = \{i\} \cup \{j : j \text{ is connected to } i\}$.

--

**Ego-cluster randomization** assigns the same treatment to all units in $\mathcal{N}_i$:

$$D_j = D_i \quad \forall j \in \mathcal{N}_i$$

--

This ensures that, for unit $i$, all first-degree neighbors have the same treatment status. If interference is limited to first-degree connections:

$$Y_i(D_i, \mathbf{D}_{\mathcal{N}_i}) = Y_i(D_i, D_i, \ldots, D_i) = Y_i(D_i)$$

and SUTVA is restored within the ego cluster.

--

**Challenges:**
- Ego clusters overlap: if $i$ and $j$ are neighbors, $\mathcal{N}_i \cap \mathcal{N}_j \neq \emptyset$
- Must resolve overlaps (e.g., graph coloring, independent set sampling)
- Power depends on the number of *non-overlapping* ego clusters
- Second-degree interference is not addressed
