---
title: "Module 1: The Experimental Ideal"
subtitle: "Potential Outcomes, Selection Bias, and Why Randomization Works"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{css, echo=FALSE}
.remark-code, .remark-inline-code { font-size: 80%; }
.remark-slide-content { padding: 1em 2em; }
.small { font-size: 80%; }
.tiny { font-size: 65%; }
.highlight-box {
  background: #fff3e0;
  border-left: 4px solid #e65100;
  padding: 0.5em 1em;
  margin: 0.5em 0;
}
.blue-box {
  background: #e3f2fd;
  border-left: 4px solid #1565c0;
  padding: 0.5em 1em;
  margin: 0.5em 0;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, fig.align = "center",
  fig.width = 10, fig.height = 4.2, dpi = 150,
  message = FALSE, warning = FALSE
)
library(tidyverse)
library(estimatr)
theme_set(theme_minimal(base_size = 14))
set.seed(42)
```

# Course Map

<table>
<tr><th>#</th><th>Module</th><th>Status</th></tr>
<tr><td><b>1</b></td><td><b>The Experimental Ideal</b> <i>(you are here)</i></td><td>current</td></tr>
<tr><td>2</td><td>SUTVA and When It Breaks</td><td>upcoming</td></tr>
<tr><td>3</td><td>Designing Around Interference</td><td>upcoming</td></tr>
<tr><td>4</td><td>Power and Sample Size</td><td>upcoming</td></tr>
<tr><td>5</td><td>Analyzing Experiments</td><td>upcoming</td></tr>
<tr><td>6</td><td>Multiple Testing & Subgroups</td><td>upcoming</td></tr>
<tr><td>7</td><td>External Validity</td><td>upcoming</td></tr>
<tr><td>8</td><td>Advanced Topics</td><td>upcoming</td></tr>
</table>

---

# The Setup: A Driver Bonus Experiment

A ride-sharing platform wants to know: **does offering a $500 sign-on bonus increase driver retention at 6 months?**

--

The product team looks at the data:

```{r naive-data, echo=FALSE}
n <- 1000
drivers <- tibble(
  motivation = rnorm(n),
  # More motivated drivers are more likely to seek out the bonus
  got_bonus = rbinom(n, 1, prob = plogis(0.8 * motivation)),
  # Retention depends on motivation AND bonus (true effect = 10pp)
  retained_6mo = rbinom(n, 1, prob = plogis(-0.3 + 0.4 * motivation +
                                               0.4 * got_bonus))
)

drivers |>
  group_by(got_bonus) |>
  summarise(n = n(), retention_rate = mean(retained_6mo)) |>
  mutate(got_bonus = if_else(got_bonus == 1, "Got bonus", "No bonus")) |>
  knitr::kable(col.names = c("Group", "N", "Retention rate"), digits = 3)
```

--

"The bonus increases retention by ~15 percentage points! Ship it!"

...or does it?

---

# The Fundamental Problem

We want to know the **causal effect** of the bonus on retention for each driver $i$:

$$\tau_i = Y_i(1) - Y_i(0)$$

--

| Driver | $Y_i(1)$ (with bonus) | $Y_i(0)$ (without) | Effect |
|--------|----------------------|---------------------|--------|
| Alice  | 1 (retained)         | ?                   | ?      |
| Bob    | ?                    | 0 (churned)         | ?      |
| Carol  | 1 (retained)         | ?                   | ?      |
| Dave   | ?                    | 1 (retained)        | ?      |

--

We only observe **one** potential outcome per driver. The other is the *counterfactual* — it doesn't exist in the data. We can never compute $\tau_i$ directly.

---

# What We *Can* Estimate: Average Effects

Even though individual effects are unobservable, we can estimate **averages**:

--

- **ATE** = $E[Y_i(1) - Y_i(0)]$ — average effect across *everyone*

--

- **ATT** = $E[Y_i(1) - Y_i(0) \mid D_i = 1]$ — average effect on those who *got* treatment

--

- **ATU** = $E[Y_i(1) - Y_i(0) \mid D_i = 0]$ — average effect on the *untreated*

--

.highlight-box[
**When do these differ?** When (1) treatment effects are heterogeneous AND (2) who gets treated is correlated with the effect. Example: if motivated drivers both seek the bonus AND benefit more from it, then ATT > ATE > ATU.
]

---

# Selection Bias: The Decomposition

The naive comparison — difference in observed means — is:

$$E[Y_i \mid D_i = 1] - E[Y_i \mid D_i = 0]$$

--

This equals:

$$= \underbrace{E[Y_i(1) - Y_i(0) \mid D_i = 1]}_{\text{ATT}} + \underbrace{E[Y_i(0) \mid D_i = 1] - E[Y_i(0) \mid D_i = 0]}_{\text{selection bias}}$$

--

.highlight-box[
**Selection bias** = the difference in *baseline* outcomes between the groups. Drivers who sought out the bonus were more motivated → they would have had higher retention *even without the bonus*.
]

---

# Selection Bias: Visualized

```{r selection-bias-viz, echo=FALSE, fig.height=4}
drivers |>
  mutate(Group = if_else(got_bonus == 1, "Got bonus", "No bonus")) |>
  ggplot(aes(motivation, fill = Group)) +
  geom_density(alpha = 0.5) +
  geom_vline(data = drivers |>
               group_by(got_bonus) |>
               summarise(m = mean(motivation)) |>
               mutate(Group = if_else(got_bonus == 1, "Got bonus", "No bonus")),
             aes(xintercept = m, color = Group),
             linetype = "dashed", linewidth = 1) +
  labs(title = "The groups are NOT comparable",
       subtitle = "Drivers who got the bonus have higher baseline motivation",
       x = "Motivation (unobserved)", y = "Density") +
  theme(legend.position = "top")
```

The bonus group has higher motivation on average → higher retention *regardless of the bonus*. The naive estimate conflates the bonus effect with pre-existing differences.

---

# Randomization: The Fix

Randomly assign 500 drivers to bonus, 500 to control:

```{r randomize}
n <- 1000
rct <- tibble(
  driver_id = 1:n,
  motivation = rnorm(n),
  # Random assignment — independent of motivation
  bonus = sample(rep(c(0, 1), each = n/2)),
  # True effect of bonus = 10pp (0.4 on logit scale)
  retained = rbinom(n, 1, prob = plogis(-0.3 + 0.4 * motivation +
                                          0.4 * bonus))
)
```

--

```{r rct-balance, echo=FALSE}
rct |>
  group_by(bonus) |>
  summarise(n = n(), mean_motivation = mean(motivation),
            retention = mean(retained)) |>
  mutate(bonus = if_else(bonus == 1, "Bonus", "Control")) |>
  knitr::kable(digits = 3, col.names = c("Group", "N", "Mean motivation",
                                          "Retention"))
```

Motivation is balanced (by construction). The difference in retention is an unbiased estimate of the ATE.

---

# Why Does Randomization Work?

Random assignment makes treatment independent of potential outcomes:

$$(Y_i(1), Y_i(0)) \perp D_i$$

--

This kills selection bias:

$$E[Y_i(0) \mid D_i = 1] = E[Y_i(0) \mid D_i = 0]$$

--

So the simple difference in means equals the ATE:

$$E[Y \mid D=1] - E[Y \mid D=0] = E[Y(1)] - E[Y(0)] = \text{ATE}$$

--

.blue-box[
**But:** randomization guarantees this *in expectation*, not in every sample. Any given experiment can have imbalanced covariates by chance. That's what inference (confidence intervals, p-values) is for.
]

---

# Simulation: Randomization Works in Expectation

```{r sim-many-rcts, echo=FALSE, fig.height=3.8}
true_ate <- mean(plogis(-0.3 + 0.4 * rct$motivation + 0.4) -
                   plogis(-0.3 + 0.4 * rct$motivation))

sim_rct <- function(i) {
  bonus <- sample(rep(c(0, 1), each = n/2))
  retained <- rbinom(n, 1, prob = plogis(-0.3 + 0.4 * rct$motivation +
                                           0.4 * bonus))
  mean(retained[bonus == 1]) - mean(retained[bonus == 0])
}

estimates <- tibble(
  estimate = map_dbl(1:1000, sim_rct)
)

ggplot(estimates, aes(estimate)) +
  geom_histogram(bins = 40, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = true_ate, color = "firebrick",
             linewidth = 1.2, linetype = "dashed") +
  annotate("text", x = true_ate + 0.005, y = 60,
           label = paste("True ATE =", round(true_ate, 3)),
           color = "firebrick", hjust = 0, size = 4) +
  labs(title = "1,000 simulated RCTs: estimates centered on the truth",
       subtitle = "Each bar is one experiment. Variation = sampling noise, not bias.",
       x = "Estimated ATE (difference in retention rates)", y = "Count")
```

---

# The Naive Estimate Is Biased

```{r sim-naive, echo=FALSE, fig.height=3.8}
sim_naive <- function(i) {
  # Non-random: motivated drivers more likely to get bonus
  bonus <- rbinom(n, 1, prob = plogis(0.8 * rct$motivation))
  retained <- rbinom(n, 1, prob = plogis(-0.3 + 0.4 * rct$motivation +
                                           0.4 * bonus))
  mean(retained[bonus == 1]) - mean(retained[bonus == 0])
}

naive_estimates <- tibble(estimate = map_dbl(1:1000, sim_naive))

bind_rows(
  estimates |> mutate(method = "Randomized"),
  naive_estimates |> mutate(method = "Observational (naive)")
) |>
  ggplot(aes(estimate, fill = method)) +
  geom_histogram(bins = 40, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = true_ate, color = "firebrick",
             linewidth = 1.2, linetype = "dashed") +
  labs(title = "Randomized vs Naive: bias is visible",
       subtitle = "Naive estimates are systematically too high (selection bias)",
       x = "Estimated effect", y = "Count") +
  theme(legend.position = "top")
```

The naive distribution is shifted right — it **overestimates** the effect because motivated drivers both seek the bonus and retain more.

---

# Numerical Decomposition

.pull-left[
```{r decompose, echo=FALSE}
# Compute the decomposition for one observational sample
set.seed(99)
bonus_obs <- rbinom(n, 1, prob = plogis(0.8 * rct$motivation))
y1 <- plogis(-0.3 + 0.4 * rct$motivation + 0.4)  # potential outcome under treatment
y0 <- plogis(-0.3 + 0.4 * rct$motivation)          # potential outcome under control

naive_est <- mean(y1[bonus_obs == 1]) - mean(y0[bonus_obs == 0])
att <- mean(y1[bonus_obs == 1] - y0[bonus_obs == 1])
sel_bias <- mean(y0[bonus_obs == 1]) - mean(y0[bonus_obs == 0])

tibble(
  Component = c("Naive estimate", "= ATT", "+ Selection bias"),
  Value = c(naive_est, att, sel_bias)
) |>
  knitr::kable(digits = 3)
```
]

.pull-right[
- **ATT** = the real causal effect on treated drivers
- **Selection bias** = treated drivers had higher $Y(0)$ (they would have retained more even without the bonus)
- **Naive** = ATT + bias → overstates the effect
]

---

# ATE vs ATT vs ATU: When It Matters

```{r ate-att, echo=FALSE, fig.height=3.5}
# Heterogeneous effects: high-motivation drivers benefit less
rct_het <- tibble(
  motivation = rnorm(n),
  effect = 0.6 - 0.3 * motivation,  # less motivated benefit more
  y0 = plogis(-0.3 + 0.4 * motivation),
  y1 = plogis(-0.3 + 0.4 * motivation + effect),
  tau = y1 - y0,
  got_bonus = rbinom(n, 1, prob = plogis(0.8 * motivation))
)

tibble(
  Estimand = c("ATE", "ATT", "ATU"),
  Value = c(mean(rct_het$tau),
            mean(rct_het$tau[rct_het$got_bonus == 1]),
            mean(rct_het$tau[rct_het$got_bonus == 0]))
) |>
  ggplot(aes(Estimand, Value, fill = Estimand)) +
  geom_col(show.legend = FALSE, width = 0.5) +
  geom_text(aes(label = round(Value, 3)), vjust = -0.5, size = 5) +
  scale_fill_manual(values = c("steelblue", "firebrick", "grey60")) +
  labs(title = "ATE, ATT, and ATU can differ substantially",
       subtitle = "Here: less motivated drivers benefit more, but motivated drivers seek the bonus",
       y = "Average treatment effect") +
  ylim(0, 0.15)
```

**Interview question:** "We ran an experiment and found a 5% lift. Should we roll out to everyone?" The answer depends on whether the experimental effect (≈ ATE) equals the effect on the marginal users who would adopt (≈ ATU).

---

# SUTVA: A Preview

Everything so far assumed:

1. **No interference:** Alice's bonus doesn't affect Bob's retention
2. **No hidden versions:** the bonus is the same for everyone

--

.highlight-box[
In marketplace experiments, **both assumptions are routinely violated:**
- Giving bonuses to some drivers changes the market for all drivers (more supply → fewer rides per driver)
- The "same" bonus might be worth different amounts in different cities
]

--

This is Module 2's topic — and it's what tech interviewers care about most. The experimental ideal from this module is the *starting point*, not the ending point.

---

# Key Takeaways

1. **Potential outcomes** define the causal effect: $\tau_i = Y_i(1) - Y_i(0)$. We can never observe both.

2. The naive comparison = **ATT + selection bias**. Selection bias = pre-existing differences between groups.

3. **Randomization** makes treatment independent of potential outcomes → kills selection bias → simple difference in means is unbiased for ATE.

4. Randomization works **in expectation**, not in every sample. Inference handles the sampling variation.

5. **ATE ≠ ATT ≠ ATU** when effects are heterogeneous and treatment is correlated with the effect. Always ask "for whom?"

6. **SUTVA** (no interference, no hidden versions) is assumed throughout — and almost always violated in tech. That's next.

---

# Exercise Preview

In the exercise you will:

1. Simulate potential outcomes for 1,000 drivers with heterogeneous effects
2. Show that the naive comparison is biased (reproduce the decomposition)
3. Show that randomization eliminates bias in expectation (1,000 simulated RCTs)
4. Compute ATE, ATT, ATU and show when they diverge
5. Break SUTVA in a simple simulation and see what happens to the estimate

See `exercise.R` for the starter code.

---
class: center, middle, inverse

# Backup Slides

---
name: selection-proof

# Backup: Deriving the Selection Bias Decomposition

Start with the naive estimand (difference in observed conditional means):

$$E[Y \mid D=1] - E[Y \mid D=0]$$

--

Since $Y = D \cdot Y(1) + (1-D) \cdot Y(0)$, the treated group shows $Y(1)$ and the control group shows $Y(0)$:

$$= E[Y(1) \mid D=1] - E[Y(0) \mid D=0]$$

--

Add and subtract $E[Y(0) \mid D=1]$:

$$= \underbrace{E[Y(1) \mid D=1] - E[Y(0) \mid D=1]}_{\text{ATT}} + \underbrace{E[Y(0) \mid D=1] - E[Y(0) \mid D=0]}_{\text{selection bias}}$$

--

Under randomization, $E[Y(0) \mid D=1] = E[Y(0) \mid D=0]$, so selection bias = 0 and ATT = ATE.
