---
title: "Module 5: Analyzing Experiments"
subtitle: "From Difference in Means to CUPED, ITT, and LATE"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{css, echo=FALSE}
.remark-code, .remark-inline-code { font-size: 80%; }
.remark-slide-content { padding: 1em 2em; }
.small { font-size: 80%; }
.tiny { font-size: 65%; }
.highlight-box {
  background: #fff3e0;
  border-left: 4px solid #e65100;
  padding: 0.5em 1em;
  margin: 0.5em 0;
}
.blue-box {
  background: #e3f2fd;
  border-left: 4px solid #1565c0;
  padding: 0.5em 1em;
  margin: 0.5em 0;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, fig.align = "center",
  fig.width = 10, fig.height = 4.2, dpi = 150,
  message = FALSE, warning = FALSE
)
library(tidyverse)
theme_set(theme_minimal(base_size = 14))
set.seed(42)
```

# Course Map

<table>
<tr><th>#</th><th>Module</th><th>Status</th></tr>
<tr><td>1</td><td><a href="../module-01/slides.html">The Experimental Ideal</a></td><td>✓ done</td></tr>
<tr><td>2</td><td>SUTVA and When It Breaks</td><td>done</td></tr>
<tr><td>3</td><td>Designing Around Interference</td><td>done</td></tr>
<tr><td>4</td><td>Power and Sample Size</td><td>done</td></tr>
<tr><td><b>5</b></td><td><b>Analyzing Experiments</b> <i>(you are here)</i></td><td>current</td></tr>
<tr><td>6</td><td>Multiple Testing & Subgroups</td><td>upcoming</td></tr>
<tr><td>7</td><td>External Validity</td><td>upcoming</td></tr>
<tr><td>8</td><td>Advanced Topics</td><td>upcoming</td></tr>
</table>

---

# Start Simple: Difference in Means

A ride-sharing company tests a $50 driver bonus. 2,000 drivers randomized equally.

```{r dim-setup}
n <- 2000
drivers <- tibble(
  id = 1:n,
  pre_trips = rpois(n, lambda = 30),  # pre-treatment trips
  bonus = rep(c(0, 1), each = n / 2),
  # True effect: 3 extra trips. Noisy outcome correlated with pre_trips.
  post_trips = pre_trips + 3 * bonus + rnorm(n, sd = 12)
)
```

--

```{r dim-estimate}
mean(drivers$post_trips[drivers$bonus == 1]) -
  mean(drivers$post_trips[drivers$bonus == 0])
```

--

This is unbiased for the ATE. But the standard error is large because `post_trips` is noisy.

---

# Regression: Same Thing, More Flexible

```{r regression-basic}
fit_simple <- lm(post_trips ~ bonus, data = drivers)
summary(fit_simple)$coefficients |> round(3)
```

--

The coefficient on `bonus` is identical to the difference in means.

--

.blue-box[
**Key fact:** In a randomized experiment with no covariates, OLS regression of $Y$ on $D$ gives exactly the difference in means. The regression framework just makes it easier to add covariates and compute standard errors.
]

---

# Adding Covariates: Why and How

Pre-treatment trips (`pre_trips`) predict post-treatment trips. Adding it as a covariate reduces residual variance:

```{r regression-covariate}
fit_covar <- lm(post_trips ~ bonus + pre_trips, data = drivers)
summary(fit_covar)$coefficients |> round(3)
```

--

.pull-left[
**Without covariate:**
```{r compare-se, echo=FALSE}
cat("SE(bonus):", round(summary(fit_simple)$coefficients[2, 2], 3))
```
]

.pull-right[
**With covariate:**
```{r compare-se2, echo=FALSE}
cat("SE(bonus):", round(summary(fit_covar)$coefficients[2, 2], 3))
```
]

--

.highlight-box[
The standard error dropped substantially. Same data, same experiment, but a **tighter** estimate -- just by using information we already had.
]

---

# Lin (2013): Covariate Adjustment Done Right

Simply adding covariates to a regression can be slightly wrong when treatment effects are heterogeneous. Lin (2013) showed the right way:

--

.blue-box[
**Lin's estimator:** Interact the (demeaned) covariate with treatment:

$$Y_i = \alpha + \tau D_i + \beta (X_i - \bar{X}) + \gamma D_i (X_i - \bar{X}) + \varepsilon_i$$

This is **always** at least as precise as the simple difference in means, regardless of whether the covariate matters.
]

--

.small[
```{r lin-estimator}
drivers <- drivers |> mutate(pre_trips_dm = pre_trips - mean(pre_trips))
fit_lin <- lm(post_trips ~ bonus * pre_trips_dm, data = drivers)
summary(fit_lin)$coefficients |> round(3)
```
]

The treatment effect estimate is the coefficient on `bonus`.

---

# CUPED: Variance Reduction Using Pre-Experiment Data

.blue-box[
**CUPED** (Controlled-experiment Using Pre-Experiment Data): Use the pre-treatment outcome to reduce variance of the treatment effect estimate. Widely used at Microsoft, Netflix, and other tech companies.
]

--

The idea: construct a variance-reduced outcome:

$$\tilde{Y}_i = Y_i - \theta (X_i - \bar{X})$$

where $X_i$ is the pre-treatment outcome and $\theta = \frac{\text{Cov}(Y, X)}{\text{Var}(X)}$.

--

```{r cuped-implement}
theta <- cov(drivers$post_trips, drivers$pre_trips) / var(drivers$pre_trips)
drivers <- drivers |>
  mutate(y_cuped = post_trips - theta * (pre_trips - mean(pre_trips)))

# Compare variance
cat("Var(post_trips):", round(var(drivers$post_trips), 1), "\n")
cat("Var(y_cuped):   ", round(var(drivers$y_cuped), 1), "\n")
cat("Variance reduction:", round(1 - var(drivers$y_cuped)/var(drivers$post_trips), 3))
```

---

# CUPED = Regression (It's the Same Thing)

```{r cuped-vs-regression}
# CUPED estimate
cuped_est <- mean(drivers$y_cuped[drivers$bonus == 1]) -
             mean(drivers$y_cuped[drivers$bonus == 0])

# Regression with covariate
reg_est <- coef(fit_covar)["bonus"]

cat("CUPED estimate:     ", round(cuped_est, 3), "\n")
cat("Regression estimate:", round(reg_est, 3), "\n")
```

--

.highlight-box[
**CUPED is just regression adjustment repackaged.** The optimal $\theta$ is the OLS coefficient of $Y$ on $X$. The CUPED-adjusted difference in means equals the regression coefficient on treatment.

Why use CUPED framing? It's easier to implement in production pipelines (you transform the metric once, then compute simple means).
]

---

# Simulation: CUPED Variance Reduction

```{r cuped-sim, echo=FALSE, fig.height=4}
sim_cuped <- function(i, use_cuped = FALSE) {
  pre <- rpois(2000, lambda = 30)
  treat <- rep(c(0, 1), each = 1000)
  post <- pre + 3 * treat + rnorm(2000, sd = 12)
  if (use_cuped) {
    th <- cov(post, pre) / var(pre)
    post <- post - th * (pre - mean(pre))
  }
  mean(post[treat == 1]) - mean(post[treat == 0])
}

raw_ests <- map_dbl(1:2000, ~ sim_cuped(.x, FALSE))
cuped_ests <- map_dbl(1:2000, ~ sim_cuped(.x, TRUE))

bind_rows(
  tibble(estimate = raw_ests, method = "Raw difference in means"),
  tibble(estimate = cuped_ests, method = "CUPED-adjusted")
) |>
  ggplot(aes(estimate, fill = method)) +
  geom_histogram(bins = 50, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = 3, linetype = "dashed", color = "firebrick", linewidth = 1) +
  annotate("text", x = 3.1, y = 100, label = "True effect = 3",
           color = "firebrick", hjust = 0, size = 4) +
  labs(title = "CUPED tightens the distribution of estimates",
       subtitle = paste0("SD(raw) = ", round(sd(raw_ests), 2),
                        " vs SD(CUPED) = ", round(sd(cuped_ests), 2)),
       x = "Estimated treatment effect", y = "Count") +
  theme(legend.position = "top")
```

Both centered on the truth, but CUPED has a much narrower spread.

---

# Stratification and Post-Stratification

**Stratified randomization:** Randomize *within* blocks defined by important covariates. Guarantees exact balance.

--

```{r stratification}
# Stratify by pre_trips tercile
drivers_strat <- drivers |>
  mutate(stratum = ntile(pre_trips, 3),
         bonus_strat = ave(1:n(), stratum, FUN = function(x)
           sample(rep(c(0, 1), length.out = length(x)))))

# Post-stratification: add stratum fixed effects
fit_strat <- lm(post_trips ~ bonus + factor(stratum), data = drivers_strat)
summary(fit_strat)$coefficients[1:2, ] |> round(3)
```

--

.highlight-box[
**Post-stratification** adjusts for strata after the fact (even if randomization wasn't stratified). It's a free variance reduction -- always include randomization strata as fixed effects.
]

---

# Robust Standard Errors

OLS standard errors assume homoskedasticity. In experiments, we should use **heteroskedasticity-robust** (HC) standard errors.

--

.small[
```{r robust-se}
# HC2 standard errors (recommended for experiments)
# Using the sandwich formula manually:
X <- model.matrix(fit_covar)
e <- residuals(fit_covar)
hat_vals <- hatvalues(fit_covar)
meat <- t(X) %*% diag(e^2 / (1 - hat_vals)) %*% X
bread <- solve(t(X) %*% X)
vcov_hc2 <- bread %*% meat %*% bread
se_hc2 <- sqrt(diag(vcov_hc2))

cat("OLS SE: ", round(summary(fit_covar)$coefficients[2, 2], 3), "\n")
cat("HC2 SE: ", round(se_hc2[2], 3), "\n")
```
]

--

For **cluster-randomized** designs, cluster the standard errors at the level of randomization:

.small[
```{r cluster-se-note, eval=FALSE}
# If randomized at city level:
# lm(y ~ treatment, data = df) with cluster-robust SEs at city level
# Or use fixest::feols(y ~ treatment, cluster = ~city, data = df)
```
]

---

# ITT vs LATE: Non-Compliance

Not everyone assigned to treatment actually takes the treatment.

```{r noncompliance-setup}
n <- 2000
experiment <- tibble(
  id = 1:n,
  assigned = rep(c(0, 1), each = n / 2),
  # 70% compliance: only 70% of those assigned to bonus actually use it
  complied = if_else(assigned == 1, rbinom(n, 1, prob = 0.70), 0L),
  # Effect only for compliers: 5 extra trips
  post_trips = rpois(n, lambda = 30) + 5 * complied + rnorm(n, sd = 8)
)
```

--

.pull-left[
**ITT** (Intent-to-Treat):
```{r itt}
itt <- mean(experiment$post_trips[experiment$assigned == 1]) -
       mean(experiment$post_trips[experiment$assigned == 0])
cat("ITT:", round(itt, 2))
```
]

.pull-right[
**LATE** (Local Average Treatment Effect):
```{r late}
compliance_rate <- mean(experiment$complied[experiment$assigned == 1])
late <- itt / compliance_rate
cat("LATE:", round(late, 2), "\n")
cat("Compliance:", round(compliance_rate, 2))
```
]

---

# ITT vs LATE: What to Report?

.blue-box[
**ITT** = effect of *assigning* treatment. Always identified by randomization. This is the policy-relevant effect: "what happens if we offer this to everyone?"

**LATE** = effect of *taking* treatment on compliers. Requires additional assumptions (monotonicity, exclusion restriction). This is the "efficacy" effect.
]

--

```{r itt-late-iv, echo=FALSE}
# IV regression (2SLS) gives LATE
fit_iv_first <- lm(complied ~ assigned, data = experiment)
experiment$complied_hat <- fitted(fit_iv_first)
fit_iv_second <- lm(post_trips ~ complied_hat, data = experiment)

tibble(
  Estimand = c("ITT (OLS on assignment)", "LATE (IV / Wald)",
               "Naive (OLS on compliance)"),
  Estimate = c(coef(lm(post_trips ~ assigned, data = experiment))[2],
               coef(fit_iv_second)[2],
               coef(lm(post_trips ~ complied, data = experiment))[2]),
  Issue = c("Diluted by non-compliance", "Accounts for non-compliance",
            "BIASED: compliance is endogenous")
) |> knitr::kable(digits = 2)
```

--

.highlight-box[
**Never** regress on actual treatment take-up -- it's endogenous. Use assignment as an instrument (2SLS) to get LATE, or report ITT.
]

---

# Application: Ride-Sharing Driver Bonus with CUPED

```{r rideshare-cuped, echo=FALSE, fig.height=4}
n <- 3000
set.seed(123)
ride_data <- tibble(
  driver_id = 1:n,
  city = sample(c("SF", "LA", "NYC", "CHI"), n, replace = TRUE),
  pre_earnings = rlnorm(n, meanlog = 6, sdlog = 0.5),
  bonus = rep(c(0, 1), each = n / 2),
  # True effect: $200 boost, but noisy
  post_earnings = pre_earnings * exp(rnorm(n, sd = 0.3)) + 200 * bonus + rnorm(n, sd = 100)
)

# Raw
raw_effect <- mean(ride_data$post_earnings[ride_data$bonus == 1]) -
              mean(ride_data$post_earnings[ride_data$bonus == 0])
raw_se <- sqrt(var(ride_data$post_earnings[ride_data$bonus == 1]) / (n/2) +
               var(ride_data$post_earnings[ride_data$bonus == 0]) / (n/2))

# CUPED
theta_ride <- cov(ride_data$post_earnings, ride_data$pre_earnings) /
              var(ride_data$pre_earnings)
ride_data$y_cuped <- ride_data$post_earnings -
                     theta_ride * (ride_data$pre_earnings - mean(ride_data$pre_earnings))
cuped_effect <- mean(ride_data$y_cuped[ride_data$bonus == 1]) -
                mean(ride_data$y_cuped[ride_data$bonus == 0])
cuped_se <- sqrt(var(ride_data$y_cuped[ride_data$bonus == 1]) / (n/2) +
                 var(ride_data$y_cuped[ride_data$bonus == 0]) / (n/2))

tibble(
  Method = c("Raw", "CUPED"),
  Estimate = c(raw_effect, cuped_effect),
  SE = c(raw_se, cuped_se),
  CI_lower = Estimate - 1.96 * SE,
  CI_upper = Estimate + 1.96 * SE
) |>
  ggplot(aes(Method, Estimate)) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2, linewidth = 1) +
  geom_hline(yintercept = 200, linetype = "dashed", color = "firebrick") +
  annotate("text", x = 2.3, y = 200, label = "True effect = $200",
           color = "firebrick", size = 4) +
  labs(title = "CUPED narrows the confidence interval",
       subtitle = paste0("Variance reduction: ",
                        round((1 - cuped_se^2/raw_se^2) * 100), "%"),
       y = "Estimated effect ($)")
```

---

# Application: DiD for Pre/Post Intervention

Simulated academic reporting experiment: 200 studies, 4 treatment arms (T0-T3), measure hypothesis reporting rates before and after intervention.

.small[
```{r did-setup}
n_studies <- 200
studies <- tibble(
  study_id = 1:n_studies,
  has_pap = rbinom(n_studies, 1, 0.4),
  arm = sample(rep(c("T0", "T1", "T2", "T3"), each = n_studies / 4)),
  # Pre-period: ~42% of hypotheses reported on average
  y_pre = rbeta(n_studies, shape1 = 3, shape2 = 4),
  # Post-period: T0 no change, T1-T3 increase reporting
  effect = case_when(
    arm == "T0" ~ 0,
    arm == "T1" ~ 0.03,
    arm == "T2" ~ 0.06,
    arm == "T3" ~ 0.10
  ),
  y_post = pmin(1, y_pre + effect + rnorm(n_studies, sd = 0.08))
)
```
]

---

# DiD: Estimating Treatment Effects

```{r did-regression}
# Reshape to panel
panel <- studies |>
  pivot_longer(cols = c(y_pre, y_post), names_to = "period",
               values_to = "y_report", names_prefix = "y_") |>
  mutate(post = if_else(period == "post", 1, 0))

# DiD regression with study fixed effects
fit_did <- lm(y_report ~ factor(study_id) + post:factor(arm), data = panel)

# Extract treatment effects (post x arm interactions)
did_coefs <- coef(fit_did)
treat_effects <- did_coefs[grep("post:factor\\(arm\\)", names(did_coefs))]
names(treat_effects) <- c("T1 vs T0", "T2 vs T0", "T3 vs T0")
round(treat_effects, 3)
```

--

.highlight-box[
**DiD** removes time-invariant study-level confounders. Each study is its own control. The identifying assumption: absent treatment, reporting rates would have changed equally across arms (parallel trends).
]

---

# Pre-Analysis Plans

.blue-box[
**Pre-analysis plan (PAP):** A document specifying the analysis *before* seeing the data. Includes: primary outcomes, estimating equations, subgroup analyses, and how to handle complications.
]

--

A good PAP specifies:

1. **Primary outcome(s)** and how they're constructed
2. **Estimating equation** (e.g., the DiD specification above)
3. **Standard errors** (robust? clustered? at what level?)
4. **Multiple testing** corrections (if applicable -- Module 6)
5. **Subgroup analyses** (which ones are confirmatory vs. exploratory?)
6. **Rules for deviations** (what do you do if compliance is lower than expected?)

--

.highlight-box[
**Why PAPs matter:** Without pre-commitment, researchers (or product teams) can run dozens of specifications and report only the significant ones. A PAP distinguishes confirmatory from exploratory analysis.
]

---

# Summary: The Analysis Toolkit

| Method | When to Use | Key Benefit |
|--------|------------|-------------|
| Difference in means | Simplest case | Unbiased, transparent |
| Regression + covariates | Pre-treatment predictors available | Variance reduction |
| Lin (2013) | Heterogeneous effects possible | Robust to misspecification |
| CUPED | Pre-treatment outcome available | Large variance reduction |
| Stratification FEs | Stratified randomization | Uses design information |
| HC2 standard errors | Always (in experiments) | Robust to heteroskedasticity |
| Cluster-robust SEs | Cluster randomization | Correct inference |
| ITT | Non-compliance present | Always identified |
| LATE / IV | Want "efficacy" effect | Handles non-compliance |
| DiD | Panel data (pre/post) | Controls for time-invariant confounders |

---

# Key Takeaways

1. **Difference in means** is the foundation. Regression gives the same answer but scales to covariates.

2. **Covariate adjustment** (Lin 2013, CUPED) reduces variance without introducing bias. Always use pre-treatment data when available.

3. **CUPED is regression.** The $\theta$ is just the OLS coefficient. CUPED is a convenient production implementation.

4. **Robust SEs** (HC2) should be standard. Cluster at the level of randomization.

5. **ITT vs LATE:** Report ITT as the primary result. Use IV for LATE when compliance matters. Never naively regress on take-up.

6. **Pre-analysis plans** separate confirmatory from exploratory analysis. They protect you from yourself.

---

# Exercise Preview

In the exercise you will:

1. Implement CUPED from scratch and verify it matches regression
2. Compare precision with and without covariate adjustment across 1,000 simulations
3. Implement Lin's (2013) estimator and compare to naive covariate adjustment
4. Estimate ITT and LATE with simulated non-compliance data
5. Run a DiD specification on simulated panel data

See `exercise.R` for the starter code.

---
class: center, middle, inverse

# Backup Slides

---
name: cuped-derivation

# Backup: CUPED Derivation

We want to estimate $\tau = E[Y(1)] - E[Y(0)]$ with minimum variance.

**Step 1:** Define a new random variable:
$$\tilde{Y}_i = Y_i - \theta(X_i - E[X])$$

where $X_i$ is a pre-treatment covariate (e.g., last period's outcome).

--

**Step 2:** The variance of the difference in means of $\tilde{Y}$ is:
$$\text{Var}(\hat{\tau}_{\text{CUPED}}) = \text{Var}(\hat{\tau}_{\text{raw}}) + \theta^2 \text{Var}(\bar{X}_T - \bar{X}_C) - 2\theta \text{Cov}(\hat{\tau}_{\text{raw}}, \bar{X}_T - \bar{X}_C)$$

--

**Step 3:** Minimize over $\theta$:
$$\theta^* = \frac{\text{Cov}(Y, X)}{\text{Var}(X)}$$

This is the OLS coefficient from regressing $Y$ on $X$ -- CUPED is regression.

--

**Step 4:** Variance reduction:
$$\frac{\text{Var}(\hat{\tau}_{\text{CUPED}})}{\text{Var}(\hat{\tau}_{\text{raw}})} = 1 - \text{Corr}(Y, X)^2$$

If $\text{Corr}(Y, X) = 0.7$, variance reduction = 51%.

---
name: lin-proof

# Backup: Why Lin (2013) Works

The standard covariate-adjusted estimator regresses $Y$ on $D$ and $X$:
$$Y_i = \alpha + \tau D_i + \beta X_i + \varepsilon_i$$

**Problem:** This constrains the coefficient on $X$ to be the same in treatment and control groups. If the relationship between $X$ and $Y$ differs by treatment status, $\hat{\tau}$ can be biased.

--

**Lin's fix:** Interact the demeaned covariate with treatment:
$$Y_i = \alpha + \tau D_i + \beta_0 \tilde{X}_i + \beta_1 D_i \tilde{X}_i + \varepsilon_i$$
where $\tilde{X}_i = X_i - \bar{X}$.

--

**Properties:**
1. $\hat{\tau}$ is consistent for the ATE regardless of the true data-generating process.
2. $\hat{\tau}$ is at least as precise as the unadjusted estimator (asymptotically).
3. With HC2 standard errors, inference is valid even under heteroskedasticity.

**Intuition:** Demeaning ensures the interaction term doesn't shift the intercept. The interaction allows different slopes in each group, avoiding functional form bias.

---
name: wald-estimator

# Backup: The Wald Estimator (LATE)

With one-sided non-compliance (only treatment group can take treatment):

$$\text{LATE} = \frac{\text{ITT}}{\text{compliance rate}} = \frac{E[Y | Z=1] - E[Y | Z=0]}{E[D | Z=1] - E[D | Z=0]}$$

--

where $Z$ = assignment, $D$ = actual treatment take-up.

This is equivalent to 2SLS with $Z$ as an instrument for $D$.

**Required assumptions:**
1. **Relevance:** $Z$ affects $D$ (first stage is strong)
2. **Independence:** $Z$ is randomly assigned
3. **Exclusion restriction:** $Z$ affects $Y$ only through $D$
4. **Monotonicity:** No "defiers" (no one does the opposite of their assignment)

Under these assumptions, LATE identifies the effect on **compliers** -- those who take treatment when assigned to it and don't when assigned to control.
