---
title: "Module 7: External Validity & Generalizability"
subtitle: "Your Zone-Notification Worked in 8 Cities. Will It Work in the Other 32?"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{css, echo=FALSE}
.remark-code, .remark-inline-code { font-size: 80%; }
.remark-slide-content { padding: 1em 2em; }
.small { font-size: 80%; }
.tiny { font-size: 65%; }
.highlight-box {
  background: #fff3e0;
  border-left: 4px solid #e65100;
  padding: 0.5em 1em;
  margin: 0.5em 0;
}
.blue-box {
  background: #e3f2fd;
  border-left: 4px solid #1565c0;
  padding: 0.5em 1em;
  margin: 0.5em 0;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, fig.align = "center",
  fig.width = 10, fig.height = 4.2, dpi = 150,
  message = FALSE, warning = FALSE
)
library(tidyverse)
theme_set(theme_minimal(base_size = 14))
set.seed(42)
```

```{r m7-setup, include=FALSE}
# Threads the M1-M3 zone-notification example through M7.
# Reuses M3's 40-city / accept-rate framework; adds a city-level effect modifier
# (local_effect = direct_effect + hte_slope * city_effect) so external-validity
# questions are concrete instead of generic.
n_cities <- 40
drivers_per_city <- 200            # smaller than M3 (5,000) for fast sims
n_total <- n_cities * drivers_per_city

selected_metros <- c("San Francisco", "Los Angeles", "New York City",
                     "Chicago", "Boston", "Seattle", "Austin", "Miami")

direct_effect <- 0.05              # average direct effect (M3-consistent)
interference  <- 0.07              # symmetric crowd-out per share treated
hte_slope     <- 0.40              # effect-modifier slope (NEW in M7)

set.seed(42)
cities_df <- tibble(
  city_id     = 1:n_cities,
  city_effect = rnorm(n_cities, mean = 0, sd = 0.10)
) |>
  arrange(desc(city_effect)) |>
  mutate(rank         = row_number(),
         city_name    = if_else(rank <= 8, selected_metros[rank],
                                paste0("City ", city_id)),
         selected     = rank <= 8,
         local_effect = direct_effect + hte_slope * city_effect)
```

# Course Map

<table>
<tr><th>#</th><th>Module</th><th>Status</th></tr>
<tr><td>1</td><td><a href="../module-01/slides.html">The Experimental Ideal</a></td><td>✓ done</td></tr>
<tr><td>2</td><td><a href="../module-02/slides.html">SUTVA and When It Breaks</a></td><td>✓ done</td></tr>
<tr><td>3</td><td><a href="../module-03/slides.html">Designing Around Interference</a></td><td>✓ done</td></tr>
<tr><td>4</td><td>Power and Sample Size</td><td>done</td></tr>
<tr><td>5</td><td>Analyzing Experiments</td><td>done</td></tr>
<tr><td>6</td><td>Multiple Testing & Subgroups</td><td>done</td></tr>
<tr><td><b>7</b></td><td><b>External Validity & Generalizability</b> <i>(you are here)</i></td><td>current</td></tr>
<tr><td>8</td><td>Advanced Topics</td><td>upcoming</td></tr>
</table>

---

# The Setup: Zone-Notification in 8 of 40 Cities

The platform from Modules 1–3 ran the **zone-notification experiment** in just **8 hand-picked metros** (the ones with the most engineering bandwidth — and the densest markets). Pooled estimate of the direct effect on accept rate: **≈ +0.11** — about double the +0.05 the M3 model implied for an average city.

```{r zone-experiment, echo=FALSE, fig.height=3.2}
selected_summary <- cities_df |> filter(selected) |>
  summarise(ate = mean(local_effect),
            se  = sd(local_effect) / sqrt(n()))

tibble(group = c("Control (no zone notif.)", "Zone notification"),
       p_accept = c(0.40, 0.40 + selected_summary$ate),
       se = c(0.005, 0.006)) |>
  ggplot(aes(group, p_accept, fill = group)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_errorbar(aes(ymin = p_accept - 1.96 * se,
                     ymax = p_accept + 1.96 * se), width = 0.15) +
  scale_fill_manual(values = c("grey60", "steelblue")) +
  labs(title = "Pooled estimate from the 8 selected metros",
       subtitle = paste0("Selected: SF, LA, NYC, Chicago, Boston, Seattle, Austin, Miami. ",
                         "ATE ≈ +", round(selected_summary$ate, 3)),
       x = NULL, y = "Accept rate")
```

--

The CEO asks: "Great, let's roll this out to all 40 cities." Should you?

---

# Internal vs External Validity

.pull-left[
.blue-box[
**Internal validity:** Is the causal estimate correct *within the study*?
- Randomization was done properly
- No attrition, spillovers, or compliance issues
- The estimated ATE is unbiased for the study population
]
]

.pull-right[
.blue-box[
**External validity:** Does the effect generalize *beyond the study*?
- To other populations (cities, countries)
- To other time periods
- To other implementations of the treatment
- To other outcome measures
]
]

--

.highlight-box[
**The tradeoff:** Lab experiments maximize internal validity (tight control) but may have low external validity (artificial setting). Field experiments in one site have high internal validity but limited external validity. Multi-site experiments trade some precision for generalizability.
]

---

# Why 8-City Results Might Not Generalize

```{r city-heterogeneity, echo=FALSE, fig.height=4}
ggplot(cities_df, aes(city_effect, local_effect,
                       color = selected, shape = selected)) +
  geom_hline(yintercept = direct_effect, linetype = "dotted", color = "grey60") +
  annotate("text", x = max(cities_df$city_effect), y = direct_effect,
           label = "population mean = 0.05", vjust = -0.7, hjust = 1,
           color = "grey50", size = 3) +
  geom_smooth(data = cities_df, aes(group = 1), method = "lm",
              color = "steelblue", se = TRUE, linetype = "dashed") +
  geom_point(size = 3) +
  geom_text(data = filter(cities_df, selected),
            aes(label = city_name), vjust = -0.9, size = 2.8,
            show.legend = FALSE) +
  scale_color_manual(values = c("FALSE" = "grey60", "TRUE" = "firebrick"),
                     labels = c("Not in experiment", "In experiment (8 metros)")) +
  scale_shape_manual(values = c("FALSE" = 16, "TRUE" = 17),
                     labels = c("Not in experiment", "In experiment (8 metros)")) +
  labs(title = "Local treatment effect rises with city density",
       subtitle = "The 8 selected metros sit at the top of the city_effect distribution",
       x = "City effect (demand density)",
       y = "Local direct effect on accept rate",
       color = NULL, shape = NULL) +
  theme(legend.position = "top")
```

--

The selected metros are systematically denser than the typical city. Cities with thin demand or sparse driver pools (negative `city_effect`) see much smaller effects — sometimes near zero.

---

# Site Selection Bias

.blue-box[
.small[
**Site selection bias:** experiment sites are not random samples of the target population — they're chosen for convenience, feasibility, or where the treatment is expected to work.
]
]

.pull-left[
.small[
**Common patterns:**
- Tech companies test in the densest US metros first
- Development RCTs run in places with cooperative governments
- Drug trials recruit from academic medical centers, not community clinics

In our setup: the 8 selected metros sit in the top quintile of `city_effect`. Their average local effect overstates the all-cities target.
]
]

.pull-right[
```{r site-selection, echo=FALSE, fig.height=2.4, fig.width=5}
tibble(
  site = c("8 selected metros\n(experiment)", "All 40 cities\n(target)"),
  mean_effect = c(mean(cities_df$local_effect[cities_df$selected]),
                  mean(cities_df$local_effect)),
  se = c(sd(cities_df$local_effect[cities_df$selected]) / sqrt(8),
         sd(cities_df$local_effect) / sqrt(40))
) |>
  ggplot(aes(site, mean_effect)) +
  geom_col(width = 0.5, fill = c("firebrick", "steelblue"), alpha = 0.7) +
  geom_errorbar(aes(ymin = mean_effect - 1.96 * se,
                     ymax = mean_effect + 1.96 * se), width = 0.15) +
  labs(title = "Selected metros are not representative",
       x = NULL, y = "Mean local direct effect") +
  coord_flip()
```
]

---

# Heterogeneous Treatment Effects: The Key

```{r hte-simulation, echo=FALSE, fig.height=3.3}
ggplot(cities_df, aes(city_effect, local_effect,
                       color = selected, shape = selected)) +
  geom_smooth(data = cities_df, aes(group = 1), method = "lm",
              color = "steelblue", se = TRUE, linetype = "dashed") +
  geom_point(size = 3) +
  scale_color_manual(values = c("FALSE" = "grey60", "TRUE" = "firebrick"),
                     labels = c("Not in experiment", "In experiment (top 8)")) +
  scale_shape_manual(values = c("FALSE" = 16, "TRUE" = 17),
                     labels = c("Not in experiment", "In experiment (top 8)")) +
  labs(title = "If you only experiment in the densest cities, you overestimate the effect",
       subtitle = "The slope (= hte_slope) is what makes the experiment-sample mean exceed the target mean",
       x = "city_effect (demand density)",
       y = "Local direct effect on accept rate",
       color = NULL, shape = NULL) +
  theme(legend.position = "top")
```

--

.small[
**If treatment effects are homogeneous** (same everywhere), external validity is free. The problem is entirely driven by **heterogeneous treatment effects** that correlate with site characteristics.
]

---

# Transportability: The Formal Framework

.blue-box[
**Transportability** (Pearl & Bareinboim, 2014): When can we use a causal effect estimated in population $A$ to predict the effect in population $B$?
]

--

**Intuition:** You can transport if:

1. You know **which variables** moderate the treatment effect (effect modifiers)
2. You can **measure** those variables in both populations
3. You can **adjust** for differences in those variables

--

$$\text{ATE}_B = \sum_x \text{CATE}(x) \cdot P_B(X = x)$$

Re-weight the site-specific conditional effects by the target population's covariate distribution.

--

.highlight-box[
**In practice:** This requires knowing the effect modifiers. If you miss an important one, transportability fails. Domain knowledge is essential.
]

---

# Transportability: Simulation

```{r transport-sim, echo=FALSE, fig.height=3.6}
# Source = 8 selected metros, Target = remaining 32 cities.
# Naive extrapolation uses the source mean. Transportability reweights:
# integrate the source-fitted CATE(city_effect) over the target's
# city_effect distribution.
fit_cate <- lm(local_effect ~ city_effect, data = filter(cities_df, selected))
cates <- tibble(
  source_mean   = mean(cities_df$local_effect[cities_df$selected]),
  target_mean   = mean(cities_df$local_effect[!cities_df$selected]),
  reweighted    = mean(predict(fit_cate, newdata = filter(cities_df, !selected)))
) |>
  pivot_longer(everything(), names_to = "estimate", values_to = "ate") |>
  mutate(label = recode(estimate,
                         source_mean = "Source ATE\n(8 metros)",
                         target_mean = "Target ATE\n(32 cities, truth)",
                         reweighted  = "Reweighted prediction\n(CATE × target distribution)"),
         label = factor(label, levels = c("Source ATE\n(8 metros)",
                                          "Reweighted prediction\n(CATE × target distribution)",
                                          "Target ATE\n(32 cities, truth)")))

ggplot(cates, aes(label, ate, fill = label)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  scale_fill_manual(values = c("firebrick", "steelblue", "grey50")) +
  labs(title = "Naive transport overstates the target; reweighting fixes it",
       subtitle = "Reweight source CATE(city_effect) by the target's density distribution",
       x = NULL, y = "Mean direct effect") +
  coord_flip()
```

--

Naive transport from the 8-metro source overestimates the 32-city target by ≈ `r round(cates$ate[1] - cates$ate[3], 3)`. Reweighting the source-fitted CATE by the target's `city_effect` distribution recovers the truth.

---

# Dose-Response and Mechanism

.pull-left[
**Why does it work?** matters more than **does it work?** for generalizability.

If you understand the **mechanism**, you can predict *where* the treatment will work:

- Price cut works because it shifts demand → only works if supply can absorb it
- Zone notification works because it routes drivers to demand → works wherever demand is concentrated
- A nudge works because it reduces friction → works wherever friction is the bottleneck
]

.pull-right[
```{r dose-response, echo=FALSE, fig.height=4}
set.seed(42)
dose_data <- tibble(
  dose = rep(seq(0, 1, by = 0.1), each = 200),
  y = 2 * dose / (0.3 + dose) + rnorm(11 * 200, sd = 0.3)
)

dose_data |>
  group_by(dose) |>
  summarise(mean_y = mean(y), se = sd(y)/sqrt(n())) |>
  ggplot(aes(dose, mean_y)) +
  geom_point(size = 2) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_ribbon(aes(ymin = mean_y - 1.96*se, ymax = mean_y + 1.96*se),
              alpha = 0.2, fill = "steelblue") +
  labs(title = "Dose-response curve: diminishing returns",
       subtitle = "Understanding the shape helps predict effects at different doses",
       x = "Treatment intensity (dose)", y = "Mean outcome")
```
]

--

.highlight-box[
**Interview tip:** "If you ran the experiment at one dose level, can you predict the effect at another?" Only if you have a model of the dose-response relationship. A binary experiment gives you one point on the curve.
]

---

# Temporal Validity: Effects That Decay

```{r novelty-effect, echo=FALSE, fig.height=3.2}
set.seed(42)
weeks <- 1:24
# Drivers gradually learn the high-demand zones — the *direct* effect of the
# notification decays toward a small steady-state premium as the routing info
# becomes common knowledge.
peak_effect    <- 0.12
steady_effect  <- 0.02
true_weekly_effect <- steady_effect + (peak_effect - steady_effect) * exp(-0.18 * weeks)
observed_effect <- true_weekly_effect + rnorm(24, sd = 0.008)

tibble(week = weeks, effect = observed_effect, true = true_weekly_effect) |>
  ggplot(aes(week, effect)) +
  geom_point(color = "steelblue", size = 2) +
  geom_line(aes(y = true), color = "firebrick", linewidth = 1,
            linetype = "dashed") +
  geom_hline(yintercept = steady_effect, linetype = "dotted") +
  annotate("text", x = 20, y = steady_effect + 0.005,
           label = "Steady-state effect ≈ 0.02",
           color = "grey40", size = 3.4) +
  annotate("rect", xmin = 1, xmax = 4, ymin = -Inf, ymax = Inf,
           alpha = 0.1, fill = "steelblue") +
  annotate("text", x = 2.5, y = 0.13, label = "4-week\nexperiment",
           color = "steelblue", size = 3) +
  labs(title = "Zone-notification: drivers learn, novelty fades",
       subtitle = "A 4-week experiment captures the peak, not the steady state",
       x = "Week", y = "Effect on accept rate")
```

--

.small[
**Novelty effects in tech:** drivers who didn't know about the high-demand zone now do — but once everyone knows (and the platform's other surfaces start showing the same info), the *marginal* informational value of the notification shrinks. The 4-week experiment's average (≈ 0.07) is roughly **3× the long-run effect** (≈ 0.02). Fix: run longer, or model the decay curve.
]

---

# Hawthorne and Demand Effects

.pull-left[
.blue-box[
**Hawthorne effect:** People change behavior because they know they are being observed, not because of the treatment itself.
]

- Lab experiments are most susceptible
- A/B tests where users don't know they're in an experiment are less susceptible
- But surveys, feedback forms, and opt-in studies are vulnerable
]

.pull-right[
.blue-box[
**Demand effect:** Participants guess the hypothesis and behave accordingly (or contrarily).
]

- "I think they want me to click more ads, so I will"
- Or: "I think they want me to click more, so I won't (reactance)"
- Mitigated by: blinding, behavioral (not self-reported) outcomes, deception
]

--

.highlight-box[
**For tech interviews:** These are less of a concern in standard A/B tests (users don't know their assignment). But they matter for: employee experiments, opt-in trials, user research studies, and any experiment with self-reported outcomes.
]

---

# What Would Help Generalizability?

.small[
**Strategy 1: Multi-site experiments.** Run in 5–10 cities **stratified by `city_effect`**, not just the densest 8. Estimate site-specific effects and model the heterogeneity.

**Strategy 2: Identify effect modifiers.** Use the 8 estimated ATEs as data points; fit ATE on `city_effect`; plug a *target* city's `city_effect` into the line to predict its ATE — instead of just averaging the 8.

```{r effect-modifiers}
sample_8 <- filter(cities_df, selected)
fit <- lm(local_effect ~ city_effect, data = sample_8)
predict(fit, data.frame(city_effect = -0.05))   # moderator   → ≈ 0.03
mean(sample_8$local_effect)                      # naive 8-city → ≈ 0.11
```

Same 8 cities of data → two different answers. Moderator says a thin-market city's ATE is ≈ 4× smaller than the naive guess.

**Strategy 3: Understand the mechanism.** Zone-notification routes drivers to concentrated demand. **Strongest** where demand is concentrated (high `city_effect`), **weakest** in thin markets — exactly the heterogeneity we see in the data.
]

---

# When Extrapolation Fails: A Simulation

```{r extrapolation-fail, echo=FALSE, fig.height=4}
# Same 40-city framework, but the TRUE local effect is non-linear in
# city_effect (saturates at high density, dips below zero at very low density —
# e.g., crowding *hurts* in already-thin markets).
set.seed(42)
nl_sites <- cities_df |>
  mutate(true_cate_nl = 0.05 + 0.5 * tanh(4 * city_effect) - 0.02,
         est = true_cate_nl + rnorm(n(), sd = 0.012))

ggplot(nl_sites, aes(city_effect, est, color = selected)) +
  geom_smooth(data = filter(nl_sites, selected), method = "lm",
              se = TRUE, color = "steelblue", fullrange = TRUE) +
  geom_line(aes(y = true_cate_nl, group = 1), color = "firebrick",
            linewidth = 1, linetype = "dashed") +
  geom_point(size = 3) +
  geom_vline(xintercept = min(nl_sites$city_effect[nl_sites$selected]),
             linetype = "dotted") +
  annotate("text", x = -0.18, y = 0.05,
           label = "Extrapolation\nzone", color = "firebrick", size = 3.4) +
  annotate("text", x = 0.20, y = -0.10,
           label = "Experiment\nsample", color = "steelblue", size = 3.4) +
  scale_color_manual(values = c("FALSE" = "grey60", "TRUE" = "steelblue"),
                     guide = "none") +
  labs(title = "Linear extrapolation fails when the true CATE is non-linear in city_effect",
       subtitle = "Red dashed = truth; blue band = linear fit from the 8 experiment metros",
       x = "city_effect (demand density)",
       y = "Local treatment effect")
```

---

# Strategies for Improving External Validity

| Strategy | How it helps | Cost |
|----------|-------------|------|
| **Multi-site experiments** | Directly measures heterogeneity | Expensive, complex logistics |
| **Stratified site selection** | Ensures diversity on key dimensions | Need to know the key dimensions |
| **Effect modifier analysis** | Models why effects vary | Needs theory + data on moderators |
| **Mechanism tests** | Predicts where treatment will work | Requires understanding of mechanism |
| **Replication** | Confirms or refutes in new settings | Time, resources |
| **Longer experiments** | Captures steady state, not novelty | Opportunity cost |

--

.highlight-box[
**The meta-point:** External validity cannot be tested — only argued for. Internal validity can be designed in (randomization). External validity requires theory, multiple studies, and honest uncertainty about generalizability.
]

---

# Key Takeaways

1. **Internal validity** (is the estimate right?) and **external validity** (does it generalize?) are distinct. High internal validity does not imply high external validity.

2. External validity fails when **treatment effects are heterogeneous** and the experimental population differs from the target population on the effect modifiers.

3. **Site selection bias** is pervasive: experiments run in convenient or favorable locations overstate average effects.

4. **Transportability** requires knowing the effect modifiers and adjusting for distributional differences. Without this knowledge, extrapolation is guesswork.

5. **Novelty effects** and **Hawthorne effects** threaten temporal external validity. Short experiments may overstate long-run impacts.

6. **Understanding the mechanism** is the most powerful tool for generalizability: *why* it works tells you *where* it will work.

---

# Exercise Preview

In the exercise (`exercise.R`) you will:

1. Simulate treatment effects that vary across 30 sites based on site characteristics
2. Run an experiment in a biased subset of sites and show the estimate doesn't generalize
3. Apply a transportability reweighting approach and show it recovers the target ATE
4. Simulate novelty effects and show how experiment duration affects the estimate
5. Compare multi-site vs single-site experimental designs

See `exercise.R` for the starter code.

---
class: center, middle, inverse

# Backup Slides

---
name: transport-formula

# Backup: Transportability Formula

**Setup:** Experiment in population $A$, target population $B$. Let $X$ be observed effect modifiers.

The ATE in $B$ is:

$$\tau_B = E_B[Y(1) - Y(0)] = \sum_x E_A[Y(1) - Y(0) \mid X = x] \cdot P_B(X = x)$$

--

**Assumptions required:**

1. **Conditional exchangeability across populations:** $\tau(x) = E_A[Y(1) - Y(0) \mid X = x] = E_B[Y(1) - Y(0) \mid X = x]$

2. **Positivity:** $P_A(X = x) > 0$ for all $x$ with $P_B(X = x) > 0$

3. **Correct specification** of the effect modifier set $X$

--

**In practice:** Estimate $\hat{\tau}(x)$ from the experimental data, then reweight:

$$\hat{\tau}_B = \frac{1}{n_B} \sum_{i \in B} \hat{\tau}(x_i)$$

This is analogous to survey reweighting / inverse probability weighting, but for treatment effects rather than means.

---
name: multisite-design

# Backup: Multi-Site Experimental Design

**Option 1: Same treatment, many sites**

- Estimate site-specific effects $\hat{\tau}_s$ for $s = 1, \ldots, S$
- The overall ATE is a weighted average: $\hat{\tau} = \sum_s w_s \hat{\tau}_s$
- Weights can be equal (average site effect) or proportional to site size (population ATE)
- Key: report the **variance** of site-specific effects, not just the average

--

**Option 2: Vary the treatment across sites (factorial design)**

- Different doses, implementations, or contexts
- Allows estimation of dose-response curves
- Requires more sites and/or larger samples

--

**Power consideration:**

- With $S$ sites, you have $S$ "observations" for estimating heterogeneity
- Individual-level $N$ does not help with site-level inference
- Rule of thumb: $S \ge 20$ for meaningful heterogeneity analysis

---
name: novelty-modeling

# Backup: Modeling Novelty / Decay Effects

**Exponential decay model:**

$$\tau(t) = \tau_{\infty} + (\tau_0 - \tau_{\infty}) \cdot e^{-\lambda t}$$

where:
- $\tau_0$ = initial treatment effect (includes novelty)
- $\tau_{\infty}$ = long-run (steady state) treatment effect
- $\lambda$ = decay rate
- $t$ = time since treatment onset

--

**Estimation:** Run a long experiment (or multiple cohorts with staggered start). Estimate $\tau(t)$ for each time period and fit the decay curve.

**Practical note:** If $\tau_{\infty} = 0$, the treatment has no lasting effect. This is surprisingly common for UI changes, notifications, and other "attention-grabbing" interventions. A 4-week experiment cannot distinguish a slowly-decaying effect from a permanent one.
