---
title: "Module 8: Beyond the A/B Test"
subtitle: "Modern DiD, Synthetic Control, and Causal Forests"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{css, echo=FALSE}
.remark-code, .remark-inline-code { font-size: 80%; }
.remark-slide-content { padding: 1em 2em; }
.small { font-size: 80%; }
.tiny { font-size: 65%; }
.highlight-box {
  background: #fff3e0;
  border-left: 4px solid #e65100;
  padding: 0.5em 1em;
  margin: 0.5em 0;
}
.blue-box {
  background: #e3f2fd;
  border-left: 4px solid #1565c0;
  padding: 0.5em 1em;
  margin: 0.5em 0;
}
.nav-btn {
  position: absolute;
  bottom: 12px;
  left: 40px;
  font-size: 11px;
  background: #e8eaf6;
  padding: 2px 8px;
  border-radius: 3px;
  z-index: 100;
  text-decoration: none;
  color: #1a237e;
}
.nav-btn:hover { background: #c5cae9; }
.nav-btn-br {
  position: absolute;
  bottom: 12px;
  right: 70px;
  font-size: 11px;
  background: #e8eaf6;
  padding: 2px 8px;
  border-radius: 3px;
  z-index: 100;
  text-decoration: none;
  color: #1a237e;
}
.nav-btn-br:hover { background: #c5cae9; }
.inline-btn {
  font-size: 11px;
  background: #e8eaf6;
  padding: 2px 8px;
  border-radius: 3px;
  text-decoration: none;
  color: #1a237e;
  margin-right: 6px;
  vertical-align: middle;
}
.inline-btn:hover { background: #c5cae9; }
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, fig.align = "center",
  fig.width = 10, fig.height = 4.2, dpi = 150,
  message = FALSE, warning = FALSE
)
library(tidyverse)
library(fixest)
theme_set(theme_minimal(base_size = 14))
set.seed(42)
```

# Course Map

<table>
<tr><th>#</th><th>Module</th><th>Status</th></tr>
<tr><td>1</td><td><a href="../module-01/slides.html">The Experimental Ideal</a></td><td>✓ done</td></tr>
<tr><td>2</td><td><a href="../module-02/slides.html">SUTVA and When It Breaks</a></td><td>✓ done</td></tr>
<tr><td>3</td><td><a href="../module-03/slides.html">Designing Around Interference</a></td><td>✓ done</td></tr>
<tr><td>4</td><td><a href="../module-04/slides.html">Power and Sample Size</a></td><td>✓ done</td></tr>
<tr><td>5</td><td><a href="../module-05/slides.html">Analyzing Experiments</a></td><td>✓ done</td></tr>
<tr><td>6</td><td><a href="../module-06/slides.html">Multiple Testing & Subgroups</a></td><td>✓ done</td></tr>
<tr><td>7</td><td><a href="../module-07/slides.html">External Validity</a></td><td>✓ done</td></tr>
<tr><td><b>8</b></td><td><b>Beyond the A/B Test</b> <i>(you are here)</i></td><td>current</td></tr>
</table>

---

# When Randomization Isn't Enough

Three settings where the clean experiment from M1–M5 isn't available:

.blue-box[
**1. Staggered rollouts** — Treatment turns on city-by-city over months. There's no clean "post" period that's the same for everyone. Naive two-way fixed effects (TWFE) is **biased** when effects are heterogeneous over time.

**2. One treated unit** — A policy hits a single city. No control group to randomize against. *Synthetic control* builds a counterfactual from a weighted donor pool.

**3. Heterogeneity that matters for policy** — Knowing the average effect isn't enough; you need to know *who responds*. *Causal forests* estimate treatment effects as a function of covariates.
]

--

This module: the modern toolkit for each, with proofs and DGP code in backup slides.

---

class: center, middle, inverse

# Part 1 — Modern Difference-in-Differences

---
name: classic-22-did

# Classic 2×2 DiD: Two Groups, Two Periods

One city gets the zone-notification feature at $t=2$; one doesn't. Driver weekly earnings before and after:

```{r did-22-data, echo=FALSE, fig.height=3}
did22 <- tibble(
  city = rep(c("Treated", "Control"), each = 2),
  period = rep(c("Pre", "Post"), 2),
  t = rep(c(1, 2), 2),
  y = c(800, 870, 780, 800)
)
ggplot(did22, aes(t, y, color = city)) +
  geom_line(linewidth = 1.2) + geom_point(size = 4) +
  geom_segment(aes(x = 2.05, xend = 2.05, y = 800, yend = 870),
               color = "black", linewidth = 0.4,
               arrow = arrow(ends = "both", length = unit(0.1, "in"))) +
  geom_segment(aes(x = 2.1, xend = 2.1, y = 800, yend = 820),
               color = "black", linewidth = 0.4, linetype = "dotted") +
  annotate("text", x = 2.15, y = 845, label = "Total\nchange = $70", hjust = 0, size = 3.5) +
  annotate("text", x = 2.15, y = 810, label = "Counterfactual\ntrend = $20", hjust = 0, size = 3.5) +
  scale_x_continuous(breaks = 1:2, labels = c("Pre", "Post")) +
  scale_color_manual(values = c("grey50", "firebrick")) +
  labs(x = NULL, y = "Weekly earnings ($)", color = NULL) +
  xlim(1, 2.7)
```

$$\widehat{\text{ATT}} = (\bar y_{T,\text{post}} - \bar y_{T,\text{pre}}) - (\bar y_{C,\text{post}} - \bar y_{C,\text{pre}}) = 70 - 20 = 50$$

---
name: parallel-trends-main

# Parallel Trends — The Identifying Assumption

The DiD estimator equals the ATT *only if* the treated and control groups would have followed the same trend, on average, in the absence of treatment.

.highlight-box[
**Parallel trends:** $\;\;E[Y_t(0) - Y_{t-1}(0) \mid \text{treated}] = E[Y_t(0) - Y_{t-1}(0) \mid \text{control}]$

It's an assumption about *counterfactuals* — fundamentally untestable. We can only test pretrends as a proxy.
]

--

**The pretrend test:** check whether the two groups moved in parallel *before* treatment. Failing this is bad news. **Passing it is weaker evidence than people think** — Roth (2022) shows pretrend tests have low power against the most-worrying violations.

<a href="#pretrend-power-detail" class="nav-btn">Roth pretrend critique</a>

---

name: staggered-main

# Staggered Adoption — What Changes

Real rollouts don't happen all at once. Cities adopt zone-notification at different times.

```{r staggered-dgp, echo=FALSE, fig.height=3.6}
n_cities <- 12; n_t <- 24
cohorts <- tibble(
  city = 1:n_cities,
  g = c(rep(8, 3), rep(12, 3), rep(16, 3), rep(Inf, 3))
)
panel <- expand_grid(city = 1:n_cities, t = 1:n_t) |>
  left_join(cohorts, by = "city") |>
  mutate(
    treated = t >= g,
    cohort = case_when(g == 8 ~ "Early (g=8)", g == 12 ~ "Mid (g=12)",
                       g == 16 ~ "Late (g=16)", TRUE ~ "Never"),
    # heterogeneous, time-varying effect: bigger for earlier cohorts
    eff = if_else(treated, 0.6 * (1 + 0.15 * (t - g)) *
                            (1 + 0.7 * (g == 8) - 0.7 * (g == 16)), 0),
    y = 5 + 0.05 * t + city * 0.1 + eff + rnorm(n(), 0, 0.3)
  )
ggplot(panel, aes(t, y, group = city, color = cohort)) +
  geom_line(alpha = 0.6) +
  geom_vline(data = distinct(panel, cohort, g) |> filter(is.finite(g)),
             aes(xintercept = g, color = cohort), linetype = "dashed", alpha = 0.7) +
  scale_color_manual(values = c("Early (g=8)" = "#e41a1c", "Mid (g=12)" = "#ff7f00",
                                 "Late (g=16)" = "#4daf4a", "Never" = "grey60")) +
  labs(x = "Period", y = "Outcome", color = "Adoption cohort")
```

Three cohorts adopt at $t=8, 12, 16$; three never adopt. The effect grows over time and is **larger for earlier-adopting cohorts**.

<a href="#staggered-dgp-detail" class="nav-btn-br">DGP code</a>

---

name: twfe-main

# The TWFE Estimator — Looks Innocent, Isn't

The "default" DiD with staggered timing:

$$Y_{it} = \alpha_i + \lambda_t + \beta\, D_{it} + \varepsilon_{it}$$

```{r twfe-fit}
fit_twfe <- feols(y ~ treated | city + t, data = panel)
coef(fit_twfe)["treatedTRUE"]
```

--

The true effects are positive and grow over time. The TWFE coefficient is wrong (sometimes the *wrong sign*) when:

- effects are heterogeneous *across cohorts*, or
- effects grow/shrink *within a cohort over time*.

--

**Goodman-Bacon (2021):** $\hat\beta_{\text{TWFE}}$ is a weighted average of all possible 2×2 DiDs in the data. Some of those weights are **negative** — earlier-treated units act as "controls" for later-treated ones, which subtracts a treated trend.

<a href="#goodman-bacon-derivation" class="nav-btn">decomposition</a>

---

name: bacon-viz-main

# Visualizing the Bacon Decomposition

$\hat\beta_{\text{TWFE}}$ is a **weighted average of every 2×2 DiD** in the data — each cohort vs every other cohort, plus each vs never-treated. With cohorts $g \in \{8, 12, 16, \infty\}$, that's **9 distinct 2×2 estimates**:

```{r bacon-viz, echo=FALSE, fig.height=2.9}
get_2x2 <- function(g_t, g_c, df, t_max) {
  if (is.infinite(g_c)) {
    pre <- 1:(g_t - 1); post <- g_t:t_max
  } else if (g_c > g_t) {
    pre <- 1:(g_t - 1); post <- g_t:(g_c - 1)
  } else {
    pre <- (g_c):(g_t - 1); post <- g_t:t_max
  }
  ya <- mean(df$y[df$g == g_t & df$t %in% pre])
  yb <- mean(df$y[df$g == g_t & df$t %in% post])
  yc <- mean(df$y[df$g == g_c & df$t %in% pre])
  yd <- mean(df$y[df$g == g_c & df$t %in% post])
  (yb - ya) - (yd - yc)
}
bacon <- expand_grid(g_t = c(8, 12, 16), g_c = c(8, 12, 16, Inf)) |>
  filter(g_t != g_c) |>
  rowwise() |> mutate(est = get_2x2(g_t, g_c, panel, n_t)) |> ungroup() |>
  mutate(comparison = case_when(
    is.infinite(g_c) ~ "Clean: treated vs never",
    g_c > g_t ~ "Clean: treated vs not-yet-treated",
    TRUE ~ "Forbidden: treated vs already-treated"
  ),
  label = paste0("treated g=", g_t, ",  control g=",
                  ifelse(is.infinite(g_c), "never", g_c))) |>
  arrange(comparison, est) |>
  mutate(label = factor(label, levels = label))

true_att <- mean(panel$eff[panel$treated])
twfe_est <- coef(feols(y ~ treated | city + t, data = panel))["treatedTRUE"]

ggplot(bacon, aes(label, est, fill = comparison)) +
  geom_col() + coord_flip() +
  geom_hline(aes(yintercept = true_att, linetype = "Truth"), color = "#e65100", linewidth = 0.8) +
  geom_hline(aes(yintercept = twfe_est, linetype = "TWFE estimate"), color = "black", linewidth = 0.8) +
  scale_fill_manual(values = c("Clean: treated vs never" = "#4daf4a",
                               "Clean: treated vs not-yet-treated" = "#377eb8",
                               "Forbidden: treated vs already-treated" = "#e41a1c")) +
  scale_linetype_manual(values = c("Truth" = "dashed", "TWFE estimate" = "dotted")) +
  labs(x = NULL, y = "2×2 DiD estimate", fill = NULL, linetype = NULL) +
  theme(legend.position = "right", legend.text = element_text(size = 8),
        legend.box = "vertical")
```

The **red 2×2s** use an already-treated cohort as control — they difference out a treated trend, not a counterfactual. TWFE (dotted) averages them in anyway, landing far below truth (dashed). <a href="#goodman-bacon-derivation" class="inline-btn">why</a>

---
name: modern-toolkit

# The Modern Toolkit: One Slide

All four heterogeneity-robust estimators solve the same problem — only use **clean** comparisons (treated vs not-yet-treated or never-treated):

.small[
| Estimator | Year | Core idea | R package |
|---|---|---|---|
| **Callaway & Sant'Anna** | 2021 | Group-time ATT for each cohort g and time t, then aggregate | `did` |
| **Sun & Abraham** | 2021 | Saturated event-study with cohort×event-time interactions | `fixest` (`sunab()`) |
| **de Chaisemartin & D'Haultfœuille** | 2020 | Switchers vs not-yet-switchers | `DIDmultiplegt` |
| **Borusyak, Jaravel, Spiess** | 2024 | Impute the untreated outcome from never-treated, then average residuals | `didimputation` |
]

.highlight-box[
All four converge in simple cases. Pick one, report the event-study plot, and check robustness with another. **Callaway & Sant'Anna is the most-cited workhorse.**
]

---

name: cs-main

# Callaway-Sant'Anna in Practice

.small[
**Estimand: group-time ATT** $\text{ATT}(g, t)$ — effect on units treated at $g$, evaluated at $t$. Two indices because TWFE collapsed this 2D structure into one biased number.

**Construction.** For each cohort $g$ and post-period $t \geq g$, a clean 2×2 DiD vs the never-treated, anchored at $g-1$:

$$\text{ATT}(g, t) = \big[\bar Y_{g,\,t} - \bar Y_{g,\,g-1}\big] - \big[\bar Y_{\text{nev},\,t} - \bar Y_{\text{nev},\,g-1}\big]$$

*All four indices use $g-1$ (cohort $g$'s last pre-treatment period), not $t-1$* — the **universal base period** version of CS. Same baseline for both groups → the formula is a direct parallel-trends test: "did the gap between cohort-$g$ and never-treated change from $g{-}1$ to $t$?" A separate "varying base period" version with $t-1$ exists but drifts the anchor as $t$ moves.

Plot against *event time* $e = t - g$ to align cohorts at $e=0$. No negative weights — never-treated is uncontaminated.
]

```{r cs-handcoded, echo=FALSE, fig.height=2.5}
cohort_means <- panel |> filter(is.finite(g)) |>
  group_by(g, t) |> summarise(y = mean(y), .groups = "drop")
never_means <- panel |> filter(is.infinite(g)) |>
  group_by(t) |> summarise(y_nev = mean(y), .groups = "drop")

att_gt <- cohort_means |> left_join(never_means, by = "t") |>
  group_by(g) |>
  mutate(att = (y - y[t == g - 1]) - (y_nev - y_nev[t == g - 1])) |>
  ungroup() |> mutate(e = t - g)

ggplot(att_gt, aes(e, att, color = factor(g))) +
  geom_line(linewidth = 0.9) + geom_point() +
  geom_hline(yintercept = 0) + geom_vline(xintercept = -0.5, linetype = "dashed") +
  labs(x = "Event time (t - g)", y = "ATT(g,t)", color = "Cohort g")
```

.small[
Each cohort's ATT grows post-treatment — the dynamic heterogeneity TWFE smeared. Aggregate over $g$ at each $e$ for an event study, or over both for an overall ATT. <a href="#cs-code" class="inline-btn">code</a>
]

---
name: honest-did-main

# Honest DiD — Bounds Under Partial PT Violations

Pretrend tests have low power. What if parallel trends is "almost but not quite" right?

.blue-box[
**Rambachan & Roth (2023):** Don't pretend PT holds exactly. Posit a bound on how much the post-treatment differential trend can deviate from the pre-trend, then report a *robust confidence interval*.
]

--

Two common restrictions:
- **Smoothness, parameter M:** the post-treatment violation can be at most M times larger than the largest pre-treatment violation.
- **Relative magnitudes:** the post-period violation cannot exceed the worst pre-period violation.

The output is a *sensitivity plot*: the CI grows as you allow more violation. If your conclusion survives plausible M, you're robust.

<a href="#honest-did-detail" class="nav-btn">visualization</a>

---
name: staggered-application

# Application: Staggered Zone-Notification Rollout

Three city cohorts adopt zone-notification at $t=8, 12, 16$. The true effect is positive and **larger for earlier adopters** (these cities had more underserved zones). What does each estimator say?

.small[
```{r app-twfe-vs-cs, echo=FALSE}
# True average post-treatment effect across treated obs
true_att <- panel |> filter(t >= g, is.finite(g)) |>
  summarise(true = mean(eff)) |> pull(true)
twfe <- coef(feols(y ~ treated | city + t, data = panel))["treatedTRUE"]
# Sun-Abraham via fixest::sunab — average post-treatment effect
panel_sa <- panel |> mutate(g_sa = if_else(is.infinite(g), 10000, g))
fit_sa <- feols(y ~ sunab(g_sa, t) | city + t, data = panel_sa)
sa_avg <- summary(fit_sa, agg = "att")$coeftable["ATT", "Estimate"]
# Hand-coded CS aggregate (simple average of post-treatment ATT(g,t))
cs_avg <- att_gt |> filter(t >= g) |> summarise(m = mean(att, na.rm = TRUE)) |> pull(m)

tibble(
  Estimator = c("Truth", "TWFE (biased)", "Sun-Abraham", "Callaway-Sant'Anna (hand)"),
  Estimate  = c(true_att, twfe, sa_avg, cs_avg)
) |> knitr::kable(digits = 3)
```
]

TWFE is downward-biased — early cohorts (with the largest effects) get re-used as controls for later cohorts, dragging the estimate down. CS and SA recover the truth.

---

# The Same Problem at an Online Retailer

**The membership program.**
- A randomized discount experiment generates heterogeneous effects: causal forest estimates which customers respond most to the signup offer based on tenure, category mix, and pre-period spend.
- Policy learning turns that heterogeneity into a targeting rule: whom to offer the next discount under a fixed promotional budget.

**The delivery rollout.**
- The staggered metro-by-metro rollout is the staggered-adoption DiD setting; TWFE is biased when delivery effects grow as customers learn the new promise (Module 8 of the sibling course covers the fix).
- A single flagship metro receiving same-day delivery before any others is the synthetic-control setting: build the counterfactual from donor metros still on two-day.
- Both settings appear in the decision tree for choosing among modern causal methods, which is the subject of the companion deep-dive course.

---
name: dd-interview-questions

# Sample Interview Questions: Staggered DiD

.small[
Realistic 45-min round questions on Goodman-Bacon + Callaway-Sant'Anna / Sun-Abraham. The right column gives what's being tested and a one-paragraph answer skeleton.

| Question | What it tests / answer skeleton |
|---|---|
| **1.** *"You ran TWFE on a staggered rollout of a new pricing algorithm in 30 cities (2022–24). Coefficient is +1.2%. A colleague says you can't trust it. Why?"* | Goodman-Bacon (2021) decomposition: TWFE is a weighted avg of all possible 2×2 DiDs, including ones that use *early-treated as the control* for late-treated. Under treatment-effect heterogeneity those comparisons get *negative* weights → bias of unknown sign. Ask to see the Bacon decomp + the share of negative weights. |
| **2.** *"When IS TWFE a valid estimator for staggered designs?"* | Iff (a) treatment effects are homogeneous across cohorts AND across event time, AND (b) parallel trends holds. Marketplace cases routinely fail (a) — novelty effects, learning curves — so default to Callaway-Sant'Anna, Sun-Abraham, dCDH, or BJS. |
| **3.** *"Walk me through Callaway-Sant'Anna at a high level."* | Estimate group-time $\text{ATT}(g, t)$ as a clean 2×2 DiD between cohort $g$ and the never-treated, anchored at the pre-period $g-1$. Aggregate to taste: average over $g$ at each event time $e = t - g$ (event-study) or over both for an overall ATT. No negative weights because the control group is uncontaminated. |
| **4.** *"Pre-trends look noisy but not flat. What now?"* | Don't fail-to-reject and move on. Run HonestDiD (Rambachan-Roth 2023): bound post-period bias under credible parallel-trends violations; report the smoothness parameter $M$ at which the CI just covers zero. If $M$ is implausibly large, the result is robust. |
| **5.** *"Drivers from treated cities migrate to nearby untreated cities. What breaks?"* | Parallel trends becomes a *no-cross-unit-spillover* statement. Control trends bend down → DiD **overstates** TTE. See [marketplace-matrix](#marketplace-matrix). Mitigate with distance-based donor exclusion, HonestDiD with a spillover-driven $\bar M$, or switch to switchback if randomization is feasible. |
]

---

class: center, middle, inverse

# Part 2 — Synthetic Control

---
name: sc-one-treated

# When You Have *One* Treated Unit

A single city introduces a policy (e.g., a minimum-wage floor for drivers). No randomization, no comparable control city. **Build one from a donor pool.**

```{r sc-data, echo=FALSE, fig.height=3.4}
n_donor <- 9; n_T <- 30; t_treat <- 21
donors <- matrix(NA, n_T, n_donor)
true_w <- c(0.35, 0.25, 0.15, 0.10, 0.10, 0.05, 0, 0, 0)  # the synthetic mix
for (j in 1:n_donor) {
  donors[, j] <- 5 + 0.04 * (1:n_T) + 0.5 * sin((1:n_T) / 5 + j) +
                 j * 0.15 + rnorm(n_T, 0, 0.2)
}
treated_y0 <- as.numeric(donors %*% true_w) + rnorm(n_T, 0, 0.15)
treated_y  <- treated_y0
treated_y[t_treat:n_T] <- treated_y[t_treat:n_T] - 0.8  # policy effect

sc_long <- bind_rows(
  tibble(t = 1:n_T, y = treated_y, kind = "Treated city", unit = "Treated"),
  expand_grid(t = 1:n_T, j = 1:n_donor) |>
    mutate(y = donors[cbind(t, j)], kind = "Donor pool",
           unit = paste0("Donor ", j))
)

ggplot(sc_long, aes(t, y, group = unit, color = kind, linewidth = kind)) +
  geom_line(alpha = 0.7) +
  geom_vline(xintercept = t_treat - 0.5, linetype = "dashed") +
  annotate("text", x = t_treat - 0.5, y = max(sc_long$y) * 1.02,
           label = "policy", hjust = 1.1, size = 3.5, color = "grey40") +
  scale_color_manual(values = c("Treated city" = "firebrick",
                                 "Donor pool"  = "grey60")) +
  scale_linewidth_manual(values = c("Treated city" = 1.2, "Donor pool" = 0.5),
                          guide = "none") +
  labs(x = "Period", y = "Outcome", color = NULL) +
  theme(legend.position = "top")
```

No single donor matches. A **convex combination** might.

---

name: sc-est-main

# Synthetic Control: The Estimator

Pick weights $w_j \ge 0$ with $\sum_j w_j = 1$ to match the treated unit's *pre-treatment* outcomes — a **convex combination** of donors:

$$\min_{w} \sum_{t < T_0} \!\left( Y_{1t} - \sum_{j \ge 2} w_j Y_{jt} \right)^{\!2} \;\;\text{s.t.}\;\; w_j \ge 0,\;\; \sum_j w_j = 1$$

```{r sc-fit, echo=FALSE}
library(quadprog)
pre <- 1:(t_treat - 1)
A <- donors[pre, ]; b <- treated_y[pre]
Dmat <- 2 * t(A) %*% A; dvec <- 2 * t(A) %*% b
Amat <- cbind(rep(1, n_donor), diag(n_donor))
bvec <- c(1, rep(0, n_donor))
w_hat <- solve.QP(Dmat, dvec, Amat, bvec, meq = 1)$solution
synth_y <- as.numeric(donors %*% w_hat)
gap <- treated_y - synth_y
```

```{r sc-counterfactual-plot, echo=FALSE, fig.height=2.6}
tibble(t = 1:n_T, Treated = treated_y, Synthetic = synth_y) |>
  pivot_longer(-t) |>
  ggplot(aes(t, value, color = name, linetype = name)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = t_treat - 0.5, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("Treated" = "firebrick", "Synthetic" = "steelblue")) +
  scale_linetype_manual(values = c("Treated" = "solid", "Synthetic" = "dashed")) +
  labs(x = "Period", y = "Outcome", color = NULL, linetype = NULL) +
  theme(legend.position = "top")
```

The blue dashed line is the synthetic counterfactual: $\sum_j \hat w_j Y_{jt}$ where the $\hat w$ above are `r paste(round(w_hat[w_hat > 0.01], 2), collapse = ", ")`. <a href="#sc-code" class="inline-btn">code</a>

---
name: sc-placebos

# Synthetic Control: Inference via Placebos

Standard errors don't work — we have one treated unit. Instead: **placebo-in-space**. Re-run the procedure pretending each *donor* is the treated unit. If the true treated unit's gap is unusually large vs the placebo distribution, that's evidence of an effect.

```{r sc-placebos, echo=FALSE, fig.height=3.4}
# Placebo: each donor as treated, against the rest
get_gap <- function(treated_idx) {
  y <- donors[, treated_idx]
  D <- donors[, -treated_idx, drop = FALSE]
  Dmat <- 2 * t(D[pre, ]) %*% D[pre, ]; dvec <- 2 * t(D[pre, ]) %*% y[pre]
  k <- ncol(D)
  Amat <- cbind(rep(1, k), diag(k)); bvec <- c(1, rep(0, k))
  w <- tryCatch(solve.QP(Dmat, dvec, Amat, bvec, meq = 1)$solution, error = function(e) NA)
  if (any(is.na(w))) return(rep(NA, n_T))
  y - as.numeric(D %*% w)
}
placebo_gaps <- sapply(1:n_donor, get_gap)

plot_df <- bind_rows(
  tibble(t = 1:n_T, gap = gap, unit = "Treated"),
  as_tibble(placebo_gaps) |> mutate(t = 1:n_T) |>
    pivot_longer(-t, names_to = "unit", values_to = "gap")
)
ggplot(plot_df, aes(t, gap, group = unit, color = unit == "Treated")) +
  geom_line(alpha = 0.7) + geom_hline(yintercept = 0) +
  geom_vline(xintercept = t_treat - 0.5, linetype = "dashed") +
  scale_color_manual(values = c("FALSE" = "grey70", "TRUE" = "firebrick"), guide = "none") +
  labs(x = "Period", y = "Gap (treated − synthetic)")
```

The treated gap (red) drops below the placebo cloud after $t=21$ — the effect is real.

---

name: sdid-main

# Synthetic DiD: The Bridge

Arkhangelsky et al. (2021) generalize both DiD and SC. **Synthetic DiD** uses *two* sets of weights:

.blue-box[
- **Unit weights** $\hat\omega_i$ work like SC: match pre-treatment trajectories of donors to the treated unit.
- **Time weights** $\hat\lambda_t$ match pre-treatment outcomes of the treated unit to its own post-treatment level — down-weighting periods that don't look like "now".
]

.small[
$$\widehat{\tau}_{\text{SDID}} = \arg\min_{\tau, \mu, \alpha, \beta} \sum_{i,t} (Y_{it} - \mu - \alpha_i - \beta_t - \tau D_{it})^2 \, \hat\omega_i \, \hat\lambda_t$$
]

--

Why this matters:
- **DiD** uses uniform weights → biased when donors don't match.
- **SC** uses unit weights only → can over-fit pre-period noise.
- **SDID** uses both → robust to both. In Arkhangelsky's empirical comparisons, lower MSE than either alone.

<a href="#sdid-detail" class="nav-btn">SDID estimator detail</a>

---
name: sc-interview-questions

# Sample Interview Questions: Synthetic Control

.small[
Realistic 45-min round questions on SC and SDiD. The right column gives what's tested and a one-paragraph answer skeleton.

| Question | What it tests / answer skeleton |
|---|---|
| **1.** *"Walk me through donor pool selection for an SC analysis of a single-city policy change."* | Match on pre-treatment outcomes (the load-bearing assumption) and on stable covariates (city size, market structure). Drop donors that experienced their own concurrent shocks; impose a geographic buffer to limit spillover into the donor pool. Report sensitivity to donor exclusion. |
| **2.** *"Pre-treatment fit is bad — RMSPE is bigger than the pre-period outcome SD. What now?"* | Don't trust the SC. The synthetic counterfactual isn't tracking the treated unit's pre-trend, so post-period gaps could be noise rather than treatment. Options: expand the donor pool, lengthen the pre-period, switch to SDiD (which allows an additive intercept shift and doesn't need perfect pre-fit), or move to a different design entirely. <a href="#sc-pre-fit" class="inline-btn">RMSPE detail</a> |
| **3.** *"How do you do inference for SC when there's one treated unit?"* | **Placebo-in-space** (Abadie-Diamond-Hainmueller): assign treatment to each donor in turn, re-estimate, and compute the post/pre RMSPE ratio. The empirical $p$-value is the rank of the actual treated unit's ratio in the placebo distribution. Sanity-check with placebo-in-time (fake treatment date in the pre-period). |
| **4.** *"Why use SDiD over plain SC?"* | SDiD combines unit weights (SC-style) AND time weights (DiD-style), and allows an additive shift. Result: doesn't require near-perfect pre-fit, lower MSE in Arkhangelsky et al.'s empirical comparisons, and naturally extends to multiple treated units. Default to SDiD when you have a moderate-size donor pool and pre-fit is imperfect. |
| **5.** *"You're estimating a fare-structure change in Austin via SC with Houston, Dallas, San Antonio in the donor pool. Drivers near Austin might cross-shop into those cities. What's the bias?"* | Donor pool contamination: treated city's expanded supply pulls drivers from nearby donors → donor trends adjust → synthetic Austin tracks the post-treatment Austin too closely → SC **understates** TTE. See [sc-interference-detail](#sc-interference-detail). Mitigations: geographic buffer, augmented SC (Ben-Michael-Feller-Rothstein 2021), explicit cross-city supply-elasticity model. |
]

---

class: center, middle, inverse

# Part 3 — Causal Forest

---
name: why-hte

# Why HTE — Beyond the ATE

ATE answers: "should we ship?" HTE answers: "**to whom**?"

.highlight-box[
The zone-notification feature has $\widehat{\text{ATE}} = +\$50$/wk. But the effect probably varies — by city density, driver tenure, time-of-week patterns. **Knowing the heterogeneity lets you target rollout, set personalized policies, and forecast aggregate impact under different deployment plans.**
]

--

Classic approach: pre-specify subgroups, run interactions. Two problems:
- **Multiple testing** — interview question from M6.
- **Misspecification** — the right subgroups aren't always the obvious ones.

**Causal forest** (Wager & Athey 2018, Athey-Tibshirani-Wager 2019): non-parametric estimation of $\tau(x) = E[Y(1) - Y(0) \mid X = x]$ using an ensemble of *honest* trees.

---

name: cf-how-main

# Causal Forest — How It Works

.pull-left[
**Honest splitting** is the key trick. Each tree:

1. Splits the sample into two halves.
2. Uses one half to *grow* the tree (decide where to split).
3. Uses the other half to *estimate* the treatment effect within each leaf.

This separation prevents the same observation from informing both the split rule and the estimate inside it — eliminating the over-fitting bias that plagues vanilla trees.
]

.pull-right[
**Forest-level prediction** for a new $x$:
- Each tree drops $x$ down to a leaf.
- The leaf defines a *local neighborhood* — observations with similar covariates.
- $\hat\tau(x)$ = a *weighted* DiD or IV across that neighborhood, with weights coming from how often training points share a leaf with $x$.

The result has pointwise confidence intervals — Wager & Athey prove asymptotic normality.
]

<a href="#cf-splitting-detail" class="nav-btn">splitting algorithm</a>

---

name: cf-practice-main

# Causal Forest in Practice — `grf`

```{r cf-setup, echo=FALSE}
library(grf)
# DGP: zone-notification HTE depends on driver tenure + city density
n <- 4000
X <- tibble(tenure = rexp(n, rate = 1/2), density = runif(n, 0, 1),
            age = sample(20:65, n, replace = TRUE))
W <- rbinom(n, 1, 0.5)
true_tau <- 30 + 60 * X$density - 10 * pmin(X$tenure, 4)
Y <- 700 + 50 * X$age * 0.5 + true_tau * W + rnorm(n, 0, 80)
```

.small[
```{r cf-fit}
cf <- causal_forest(X = as.matrix(X), Y = Y, W = W, num.trees = 1000)
average_treatment_effect(cf)                    # ATE estimate + SE
```
]

```{r cf-plot, echo=FALSE, fig.height=2.6}
tau_hat <- predict(cf)$predictions
bind_cols(X, true = true_tau, hat = tau_hat) |>
  ggplot(aes(density, hat, color = pmin(tenure, 5))) +
  geom_point(alpha = 0.3) +
  geom_smooth(aes(y = true), color = "black", linetype = "dashed", se = FALSE) +
  scale_color_viridis_c(name = "Tenure (yrs)") +
  labs(x = "City density", y = expression(hat(tau)(x)))
```

Predicted $\hat\tau(x)$ rises with density; the forest recovers the truth (dashed). <a href="#cf-dgp" class="inline-btn">DGP code</a>

---
name: policy-learning-teaser

# Policy Learning — A Quick Teaser

Once you have $\hat\tau(x)$, you can learn a **treatment rule** $\pi(x) \in \{0, 1\}$ that maximizes welfare:

$$\pi^*(x) = \mathbb{1}\{\hat\tau(x) > c(x)\}$$

where $c(x)$ is a per-unit cost (e.g., the cost of pushing a notification).

.blue-box[
- **`grf::policy_tree`** — fit a shallow decision tree over the covariates that maximizes estimated welfare. Output: an interpretable rule like *"deploy in dense cities to drivers with <2 years tenure."*
- **Athey & Wager (2021)** prove these are **near-optimal policies**: the learned rule's welfare gets within a vanishing-with-sample-size gap of the *optimal* policy's. So the rule isn't just interpretable — it's provably close to the best rule you could have chosen if you knew the truth.
]

This is what gets used to *deploy* the model. ATE answers "ship?". HTE answers "to whom?". Policy learning answers "**what should we actually do?**"

---

class: center, middle, inverse

# Part 4 — One-Pagers

---
name: bandits-one-pager

# Bandits — Adaptive Experimentation in One Slide

.small[
**Setup:** K arms (e.g., 4 ad creatives). At each step, pick one, observe reward. Goal: maximize cumulative reward, not just identify the best arm.

**Thompson Sampling** is the workhorse: maintain a posterior over each arm's reward, sample from it, play the argmax. Asymptotically optimal regret; minimal tuning.
]

```{r bandit-quick, echo=FALSE, fig.height=2.4}
set.seed(7)
T_steps <- 1000; K <- 4; true_p <- c(0.10, 0.12, 0.15, 0.13)
S <- rep(0, K); F <- rep(0, K); reg <- numeric(T_steps)
for (t in 1:T_steps) {
  theta <- rbeta(K, S + 1, F + 1)
  a <- which.max(theta)
  r <- rbinom(1, 1, true_p[a])
  S[a] <- S[a] + r; F[a] <- F[a] + (1 - r)
  reg[t] <- max(true_p) - true_p[a]
}
tibble(t = 1:T_steps, cum_reg = cumsum(reg)) |>
  ggplot(aes(t, cum_reg)) + geom_line(color = "firebrick", linewidth = 1) +
  labs(x = "Step", y = "Cumulative regret")
```

.small[
**When *not* to use:** when you need an unbiased ATE for a fixed-population decision (a launch / kill call). Bandits make assignment correlated with outcome history — naive analysis is biased.
]

---

name: seq-main

# Sequential Testing in One Slide

.small[
**Problem:** classical p-values are valid only at one fixed sample size. Peeking inflates Type-I error — by the end of the experiment, the running p-value will dip below 0.05 about 30% of the time *under the null*.

**Two fixes used in industry:**
- **Group-sequential / O'Brien-Fleming:** spend the alpha budget on a pre-specified schedule of looks. Conservative early, normal late.
- **Always-valid CIs / mSPRT:** mixture sequential probability ratio. Confidence sequences that hold *uniformly over time*. Stop whenever you want; CIs stay valid.
]

```{r seq-quick, echo=FALSE, fig.height=2.4}
set.seed(1)
n_max <- 2000; sims <- 200
peek_reject <- replicate(sims, {
  x <- rnorm(n_max); rejected <- FALSE
  for (n in seq(50, n_max, by = 50)) {
    if (abs(t.test(x[1:n])$statistic) > 1.96) { rejected <- TRUE; break }
  }
  rejected
})
mean(peek_reject)   # Type-I rate from peeking — should be ~0.30
```

.small[
Under the null, naive peeking yields ≈30% rejection rate, not 5%. mSPRT or OBF would hold this near 5%.
]

<a href="#mSPRT-detail" class="nav-btn">mSPRT formula</a>

---

class: center, middle, inverse

# Part 5 — Wrap-Up

---
name: decision-tree

# Decision Tree: Which Method When

.small[
**Got randomization?** → Use the M5 toolkit (regression adjustment, CUPED, ITT/LATE).

**Staggered rollout, multiple cities, multiple time periods?**
- Heterogeneous effects expected? → **Callaway-Sant'Anna** (or Sun-Abraham). Avoid TWFE.
- Worried about parallel trends? → **Honest DiD** sensitivity bounds.

**Single treated unit?** → **Synthetic control** (or **SDID** if you have a moderate-size donor pool).

**Want effect heterogeneity?** → **Causal forest** for treatment effects as a function of covariates; **policy tree** for an interpretable deployment rule.

**Many arms, online learning?** → **Thompson Sampling**, *unless* you need a clean ATE.

**Repeated looks at one experiment?** → **mSPRT / always-valid CIs**.
]

---
name: interview-cheat-sheet

# Interview Cheat Sheet

.blue-box[
**"Walk me through how you'd analyze a staggered rollout."** Three steps. (1) Plot raw outcomes by cohort and event time — does parallel trends look plausible? (2) Don't run TWFE; run Callaway-Sant'Anna and report the event-study plot. (3) Sensitivity-check with Honest DiD and a robustness column from Sun-Abraham.
]

.blue-box[
**"How do you estimate a causal effect for one city?"** Synthetic control. Build weights from a donor pool to match pre-treatment outcomes. Inference via placebo-in-space. Mention SDID as the modern improvement.
]

.blue-box[
**"How do you get heterogeneous treatment effects?"** Causal forest from `grf`. Honest splitting prevents over-fit. Aggregate to ATE for sanity-check; predict per-unit treatment effects from the covariates; feed into a policy tree if a deployment rule is needed.
]

.highlight-box[
**Red flag in interviews:** running TWFE on staggered data without a Goodman-Bacon diagnostic.
]

---
name: marketplace-matrix

# Marketplace Pros/Cons: AB / Switchback / DiD / SC

.small[
Each method evaluated through the **driver-rider network** lens (M2). "Failure mode" gives the *sign* of the bias and the mechanism.

| Method | Marketplace pro | Marketplace con | Concrete failure mode |
|---|---|---|---|
| **Individual A/B** | Cheap; clean identification *if* SUTVA held | Both arms share the same exposure profile → biased blend of DE and SE | Naive 50/50 recovers DE, *not* TTE (see [M2 worked example](../module-02/slides.html#estimands-formal-2)) — [more](#ab-interference-detail) |
| **Switchback** | Treat-all vs treat-none periods → targets **TTE directly** | Carryover; demands homogeneous time effects | Driver positioning persists across switches; OFF period isn't truly $\mathbf{0}$ → bias toward zero — [more](#switchback-interference-detail) |
| **DiD (CS / staggered)** | Uses observational rollout data; matches how features actually launch | Parallel trends ≈ "no cross-unit spillover from treated to controls" — strong in marketplaces | Drivers reposition across geo borders; control trend bends down → DiD **overstates** TTE — [more](#did-interference-detail) |
| **Synthetic control** | Targets one-treated-city rollout cleanly | Donor pool assumed unaffected by treatment | Treated city pulls supply from donor neighbors; synthetic counterfactual bends *toward* treated trend → SC **understates** TTE — [more](#sc-interference-detail) |

**Pattern.** Methods anchored on *control units* (DiD, SC) bias when treatment leaks into the control set. Methods comparing *exposure profiles directly* (switchback, full-saturation cluster A/B) target TTE but pay other costs.
]

---

# Going Deeper

This module is a tour, not a treatment. Each of the three core methods has a course's worth of material behind it.

.blue-box[
**Companion course (in development):** *Causal Inference Beyond A/B Tests* — a deep dive into modern DiD (formal CS / SA / BJS estimators, full Honest DiD), synthetic control variants (augmented SC, generalized SC, matrix completion), and the Athey-Wager causal-forest stack (policy learning, contextual bandits via causal trees).

Repo will live alongside this one at `~/Desktop/sandbox/courses/causal-inference-beyond-ab/` and link from the same landing page.
]

For the read-now references behind today's slides:

.small[
- **Goodman-Bacon (2021)** — *Difference-in-differences with variation in treatment timing*, J. Econometrics.
- **Callaway & Sant'Anna (2021)** — *Difference-in-differences with multiple time periods*, J. Econometrics.
- **Arkhangelsky et al. (2021)** — *Synthetic difference in differences*, AER.
- **Wager & Athey (2018)** — *Estimation and inference of heterogeneous treatment effects using random forests*, JASA.
- **Rambachan & Roth (2023)** — *A more credible approach to parallel trends*, ReStud.
]

---

name: course-map-end

# Course Map

<table>
<tr><th>#</th><th>Module</th><th>Status</th></tr>
<tr><td>1</td><td><a href="../module-01/slides.html">The Experimental Ideal</a></td><td>✓ done</td></tr>
<tr><td>2</td><td><a href="../module-02/slides.html">SUTVA and When It Breaks</a></td><td>✓ done</td></tr>
<tr><td>3</td><td><a href="../module-03/slides.html">Designing Around Interference</a></td><td>✓ done</td></tr>
<tr><td>4</td><td><a href="../module-04/slides.html">Power and Sample Size</a></td><td>✓ done</td></tr>
<tr><td>5</td><td><a href="../module-05/slides.html">Analyzing Experiments</a></td><td>✓ done</td></tr>
<tr><td>6</td><td><a href="../module-06/slides.html">Multiple Testing & Subgroups</a></td><td>✓ done</td></tr>
<tr><td>7</td><td><a href="../module-07/slides.html">External Validity</a></td><td>✓ done</td></tr>
<tr><td><b>8</b></td><td><b>Beyond the A/B Test</b> <i>(just finished)</i></td><td>✓ done</td></tr>
</table>

---

class: center, middle, inverse

# Backup Slides

---
name: pretrend-power-detail

# Backup: The Roth (2022) Pretrend Critique

**The standard pretrend test** has low power against linear (or near-linear) violations — exactly the kind that would bias the post-treatment estimate.

```{r roth-power, echo=FALSE, fig.height=2.6}
sim_power <- function(slope, n_pre = 4, sd = 1, n_sim = 1000) {
  mean(replicate(n_sim, {
    pre_diffs <- slope * (-n_pre:-1) + rnorm(n_pre, 0, sd)
    abs(mean(pre_diffs)) / (sd / sqrt(n_pre)) > 1.96
  }))
}
df <- tibble(slope = seq(0, 0.5, 0.05)) |>
  mutate(power = map_dbl(slope, sim_power))
ggplot(df, aes(slope, power)) + geom_line(linewidth = 1, color = "firebrick") +
  geom_hline(yintercept = 0.8, linetype = "dashed") +
  labs(x = "True linear pre-trend slope", y = "Power to detect")
```

A meaningful linear violation is missed most of the time. Roth's recommendation: use Honest DiD bounds, not the pretrend test.

<a href="#parallel-trends-main" class="nav-btn">← back</a>

---
name: staggered-dgp-detail

# Backup: Staggered-Adoption DGP

.small[
```{r staggered-dgp-show, eval=FALSE}
n_cities <- 12; n_t <- 24
cohorts <- tibble(
  city = 1:n_cities,
  g = c(rep(8, 3), rep(12, 3), rep(16, 3), rep(Inf, 3))
)
panel <- expand_grid(city = 1:n_cities, t = 1:n_t) |>
  left_join(cohorts, by = "city") |>
  mutate(
    treated = t >= g,
    # heterogeneous, time-varying effect: bigger for earlier cohorts
    eff = if_else(treated, 0.6 * (1 + 0.15 * (t - g)) *
                            (1 + 0.7 * (g == 8) - 0.7 * (g == 16)), 0),
    y = 5 + 0.05 * t + city * 0.1 + eff + rnorm(n(), 0, 0.3)
  )
```
]

Three cohorts adopt at $t = 8, 12, 16$; three never adopt. Effect grows over time and is **larger for earlier cohorts** — the classic case where TWFE breaks.

<a href="#staggered-main" class="nav-btn">← back</a>

---
name: goodman-bacon-derivation

# Backup: The Goodman-Bacon Decomposition

For TWFE with staggered timing, the OLS coefficient $\hat\beta_{\text{TWFE}}$ decomposes as:

$$\hat\beta_{\text{TWFE}} = \sum_{k} s_k \, \widehat{\text{ATT}}_k$$

where each $\widehat{\text{ATT}}_k$ is a 2×2 DiD between two cohorts (or a cohort and the never-treated), and the weights $s_k$ depend on:

- size of each cohort,
- timing of treatment,
- variance of treatment exposure within the panel.

.highlight-box[
**The bug:** when an *earlier-treated* cohort is used as a comparison group for a *later-treated* one, the post-period for the comparison cohort is also under treatment. So that 2×2 has the form $(\Delta Y_{\text{treated}}) - (\Delta Y_{\text{also treated, but longer ago}})$, and the second piece subtracts a *treatment-effect trend*, not a counterfactual.

If treatment effects grow over time, this 2×2 gets a *negative* contribution. TWFE's "weighted average" puts positive weight on this, biasing $\hat\beta$ downward.
]

<a href="#twfe-main" class="nav-btn">← back</a>

---
name: honest-did-detail

# Backup: Honest DiD Visualization

The robust CI grows as you allow more violation. The point estimate stays — only the uncertainty changes.

```{r honest-did-viz, echo=FALSE, fig.height=3.5}
# Stylized illustration
df <- tibble(
  M = seq(0, 2, 0.05),
  est = 0.5,
  lb = 0.5 - 0.15 - 0.4 * M,
  ub = 0.5 + 0.15 + 0.4 * M
)
ggplot(df, aes(M, est)) +
  geom_ribbon(aes(ymin = lb, ymax = ub), alpha = 0.3, fill = "firebrick") +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  annotate("text", x = 1.4, y = 0.55, label = "Robust CI",
           color = "firebrick", size = 4) +
  labs(x = expression("Allowed violation magnitude " * (M)),
       y = "Treatment effect estimate")
```

The smallest $M$ at which the CI crosses zero is the **breakdown value** — how much PT violation you'd need to tolerate before your conclusion flips.

<a href="#honest-did-main" class="nav-btn">← back</a>

---
name: sc-code

# Backup: Synthetic Control via `solve.QP`

.small[
```{r sc-fit-show, eval=FALSE}
library(quadprog)
pre <- 1:(t_treat - 1)
A <- donors[pre, ]; b <- treated_y[pre]

# Quadratic program:
#   minimize  || A w - b ||^2     subject to  w_j >= 0,  sum_j w_j = 1
Dmat <- 2 * t(A) %*% A
dvec <- 2 * t(A) %*% b
Amat <- cbind(rep(1, n_donor), diag(n_donor))   # constraints stacked
bvec <- c(1, rep(0, n_donor))                   # 1 for sum=1, 0 for w >= 0
w_hat   <- solve.QP(Dmat, dvec, Amat, bvec, meq = 1)$solution
synth_y <- as.numeric(donors %*% w_hat)
gap     <- treated_y - synth_y
```
]

`solve.QP` minimizes $\frac{1}{2} w^\top D w - d^\top w$ subject to $A^\top w \ge b$, with `meq` flagging the first constraint as equality. Production code uses `Synth::synth()`, which adds a covariate-importance V-matrix and a nested optimization.

<a href="#sc-est-main" class="nav-btn">← back</a>

---
name: sdid-detail

# Backup: SDID Estimator Mechanics

The SDID estimator solves:

$$(\hat\tau, \hat\mu, \hat\alpha, \hat\beta) = \arg\min \sum_{it} (Y_{it} - \mu - \alpha_i - \beta_t - \tau D_{it})^2 \, \hat\omega_i \, \hat\lambda_t$$

where $\hat\omega_i$ are unit weights from a regularized SC-style fit, and $\hat\lambda_t$ are time weights from an analogous time-domain fit.

.blue-box[
**Why the regularization matters.** SC's unit weights can over-fit pre-period noise on a short pre-period — picking a donor combination that matches noise rather than the underlying trend. SDID adds an $L_2$ penalty on the weights, smoothing the solution and reducing variance.

**Why the time weights matter.** They down-weight pre-periods that don't resemble the treatment period, focusing the comparison on "comparable" history. This is what makes SDID robust to the donor-pool drift that breaks SC.
]

R implementation: `synthdid::synthdid_estimate()`. The package returns the estimate, jackknife SE, and a built-in SC vs DiD vs SDID comparison plot.

<a href="#sdid-main" class="nav-btn">← back</a>

---
name: sc-pre-fit

# Backup: When Is Pre-Fit "Bad"?

.small[
**RMSPE** = root mean squared prediction error in the pre-period — *the objective the SC weights minimize:*
$$\text{RMSPE}_{\text{pre}} = \sqrt{\frac{1}{T_0}\sum_{t < T_0}\!\Bigl(Y_{1t} - \sum_j \hat w_j\, Y_{jt}\Bigr)^{\!2}}$$

By construction, this is the best the donor pool can do under the SC constraints ($w_j \geq 0$, $\sum w_j = 1$).

**The diagnostic.** Compare $\text{RMSPE}_{\text{pre}}$ to $\text{SD}(Y_{1t})_{\text{pre}}$ — the SD of the treated unit's own pre-period outcomes (the naive "predict the mean" baseline):

| Ratio $\text{RMSPE}_{\text{pre}} / \text{SD}(Y_{1t})$ | Interpretation |
|---|---|
| $\ll 1$ (e.g., 0.1–0.3) | Synthetic tracks treated well — pre-fit is good, SC trustworthy. |
| $\approx 1$ | Synthetic is no better than "predict the pre-period mean" — donor pool can't reconstruct the trajectory. |
| $> 1$ | Synthetic is *worse than a constant*. Very bad — post-period gap is uninterpretable. |

**Why ratio $> 1$ kills SC.** Identification rests on the pre-fit weights $\hat w$ also generating a valid counterfactual *post*-treatment. If $\hat w$ doesn't even reproduce the pre-period (worse than a constant!), there's no evidence the donor pool tracks the treated unit in the absence of treatment — any post-period gap could be fit error spilling forward, not the treatment effect. Q2's fixes (expand donors, lengthen pre-period, switch to SDiD) all target one of those failure modes.
]

<a href="#sc-interview-questions" class="nav-btn-br">← back</a>

---
name: cs-code

# Backup: Hand-Coded Callaway-Sant'Anna

.small[
```{r cs-handcoded-show, eval=FALSE}
cohort_means <- panel |> filter(is.finite(g)) |>
  group_by(g, t) |> summarise(y = mean(y), .groups = "drop")
never_means <- panel |> filter(is.infinite(g)) |>
  group_by(t) |> summarise(y_nev = mean(y), .groups = "drop")

# ATT(g, t) = (Y_g,t - Y_g,g-1) - (Y_never,t - Y_never,g-1)
att_gt <- cohort_means |> left_join(never_means, by = "t") |>
  group_by(g) |>
  mutate(att = (y - y[t == g - 1]) - (y_nev - y_nev[t == g - 1])) |>
  ungroup() |> mutate(e = t - g)
```
]

This is the simplest version: each cohort vs never-treated, anchored at $t = g - 1$. Production CS adds: not-yet-treated (vs only never-treated) controls, doubly-robust adjustment, and influence-function based standard errors. The `did` package implements all three.

<a href="#cs-main" class="nav-btn">← back</a>

---
name: cf-dgp

# Backup: Causal-Forest DGP

.small[
```{r cf-dgp-show, eval=FALSE}
library(grf)
# DGP: zone-notification HTE depends on driver tenure + city density
n <- 4000
X <- tibble(tenure = rexp(n, rate = 1/2),    # years
            density = runif(n, 0, 1),        # city density 0-1
            age = sample(20:65, n, replace = TRUE))
W <- rbinom(n, 1, 0.5)
# Heterogeneous true effect: rises with density, falls with tenure (capped)
true_tau <- 30 + 60 * X$density - 10 * pmin(X$tenure, 4)
Y <- 700 + 50 * X$age * 0.5 + true_tau * W + rnorm(n, 0, 80)
```
]

The HTE pattern: drivers in dense cities benefit most ($+\$60$/wk per density unit); long-tenured drivers benefit less (already routing efficiently). The forest must recover this without us specifying the functional form.

<a href="#cf-practice-main" class="nav-btn">← back</a>

---
name: cf-splitting-detail

# Backup: Causal-Forest Splitting Algorithm

For each candidate split $(j, c)$ on covariate $j$ at threshold $c$, define left/right child sets.

**Vanilla regression-tree criterion:** maximize variance of the *outcome mean* across the two children — i.e., split where $Y$ differs most.

**Causal-forest criterion** (Athey, Tibshirani & Wager 2019): maximize the **heterogeneity of the treatment effect** across children:

$$\Delta(j, c) = n_L \, n_R \, (\hat\tau_L - \hat\tau_R)^2 / n$$

where $\hat\tau_L, \hat\tau_R$ are local treatment-effect estimates inside each child. The `grf` implementation uses a fast gradient-based approximation rather than evaluating $\hat\tau$ exactly at every candidate split.

.highlight-box[
**Honesty:** within a tree, sample-splitting separates the data used to *grow* the tree from the data used to *estimate $\hat\tau$ inside leaves*. Without this, the same noise that drove the split would inflate the estimated effect inside the leaf — exactly the over-fit that vanilla trees suffer.
]

<a href="#cf-how-main" class="nav-btn">← back</a>

---
name: mSPRT-detail

# Backup: mSPRT — Always-Valid Inference

The mSPRT (mixture sequential probability ratio test) constructs a *confidence sequence* $\{(L_t, U_t)\}$ such that:

$$P\!\left(\theta \in (L_t, U_t)\;\;\forall\, t\right) \ge 1 - \alpha$$

The interval is *uniformly* valid — you can stop at any time.

The construction (Howard et al. 2021): for a Gaussian outcome with variance $\sigma^2$, the running CI for $\theta = E[X]$ is

$$\hat\theta_t \pm \sigma \sqrt{\frac{2 \log(1/\alpha) + \log(1 + t \rho^2)}{t \rho^2}}$$

for a chosen mixing scale $\rho$. The CI shrinks like $\sqrt{(\log t) / t}$ — slightly slower than fixed-$n$ $\sqrt{1/t}$, the price of always-valid coverage.

.blue-box[
**Used in industry by:** Optimizely (mSPRT), Microsoft (group-sequential + mSPRT), Netflix (sequential testing platform).
]

<a href="#seq-main" class="nav-btn">← back</a>

---
name: ab-interference-detail

# Backup: Individual A/B Under Driver-Rider Interference

.small[
**Setup.** 50% of drivers in a city get a zone-notification feature; the other 50% don't. Both arms operate in the same zones simultaneously.

**The hidden bias.** Treated drivers rush the zone → crowding pushes treated accept-rate *down*. Control drivers in the same zone benefit from less competition → control accept-rate goes *up*. The naive estimator $\hat\mu(1) - \hat\mu(0)$ subtracts a *contaminated control* from a *deflated treated*.

**The killer.** In the canonical DGP with symmetric crowd-out, at saturation 0.5 the naive estimator equals the **direct effect**, not TTE. The policy-relevant quantity (TTE) can have the *opposite sign* from what the A/B reports — a +5 pp A/B can mask a −2 pp rollout. (Worked example: [M2 §29](../module-02/slides.html#estimands-formal-2).)

**What to do.**
- **Cluster A/B at the geographic level** (whole cities) — if clusters are large enough to contain spillovers, the within-cluster estimate approaches TTE.
- **Switchback** — treat-all vs treat-none periods alternate; targets TTE directly.
- **Saturation experiment** — randomize the *fraction* treated across markets; trace the TTE/ADE curve and extrapolate to 100%.
- **Cluster-robust SE alone doesn't help.** It fixes variance, not bias. Don't conflate the two.
]

<a href="#marketplace-matrix" class="nav-btn">← back to matrix</a>

---
name: switchback-interference-detail

# Backup: Switchback Under Driver-Rider Interference

.small[
**Strength.** Whole-city ON/OFF periods → at any given moment every driver sees the *same* regime → the comparison is between a near $\mathbf{1}$ and a near $\mathbf{0}$ exposure regime → the estimator targets TTE directly. No treated-vs-control crowding within a period.

**Network-side weakness — physical carryover.** Driver positioning is *physical*. A treated period concentrates drivers in zone X; when the period flips OFF, those drivers don't teleport home. The first 5–10 minutes of OFF are still partially treated.

**Sign of the bias.** Carryover makes OFF periods look more like ON periods → the ON−OFF gap shrinks → switchback **underestimates** TTE in absolute value (biases toward zero).

**What to do.**
- **Washout windows.** Drop the first τ minutes after each switch from analysis. τ = 1× the median driver repositioning time is a good starting heuristic.
- **Period-length tuning.** Too short → high carryover bias. Too long → too few independent comparisons (reduces effective $n$). Bojinov et al. (2023) derive the optimal period length — see [switchback-optimal in M3](../module-03/slides.html#switchback-optimal).
- **Asymmetric carryover.** If ON→OFF and OFF→ON have different lag profiles (e.g., features that trigger habit formation), use asymmetric washouts.

**Fundamental limitation.** Switchback gives you TTE as a single number. It can't decompose into DE and SE — for that you need a saturation design.
]

<a href="#marketplace-matrix" class="nav-btn">← back to matrix</a>

---
name: did-interference-detail

# Backup: DiD Under Driver-Rider Interference

.small[
**Setup.** Uber turns on a routing-algorithm change in San Francisco at $t^*$, leaves Oakland untouched. Standard 2×2 DiD:

$$\text{DiD} = (\bar Y_{SF, post} - \bar Y_{SF, pre}) - (\bar Y_{Oak, post} - \bar Y_{Oak, pre})$$

**The hidden assumption.** Parallel trends says: *absent treatment*, $Y_{SF}$ and $Y_{Oak}$ would have moved in parallel. But "absent treatment" really means *no treatment anywhere in the network*. If the routing change in SF reroutes drivers across the bay bridge, Oakland is **not** at the no-treatment counterfactual — it loses supply.

**Sign of the bias.** Spillover *from* SF *into* Oakland → Oakland trend bends DOWN → control rises *less* than it should → DiD = (treated rise) − (control rise) gets *bigger* → **overestimates** the SF-only treatment effect. The same logic applies to staggered DiD when treated cities pull from not-yet-treated controls.

**What to do.**
- **Distance-based donor exclusion.** Drop control geos within a buffer radius of treated; use only "isolated" controls.
- **Saturation-aware adjustment.** Estimate the spillover function from cross-border movement data and subtract the implied control-unit shift.
- **HonestDiD with a spillover-driven $\bar M$.** Set the bound on parallel-trends violation to the worst-case implied by your spillover model — see [honest-did-detail](#honest-did-detail).
- **Switch to switchback or cluster A/B if randomization is feasible.** They target TTE directly without the parallel-trends-equals-no-spillover leap.
]

<a href="#marketplace-matrix" class="nav-btn">← back to matrix</a>

---
name: sc-interference-detail

# Backup: Synthetic Control Under Driver-Rider Interference

.small[
**Setup.** Uber rolls out a new fare structure in **Austin** only; estimates a synthetic Austin from a weighted combination of Houston, Dallas, San Antonio, etc.

**The hidden assumption.** The donor pool is *not* affected by treatment. With a fare change in Austin: drivers near the metro might cross-shop, riders might delay travel, operational adjustments (driver supply hours, vehicle deployment) propagate through Uber's national supply pool. *Houston is no longer at counterfactual.*

**Sign of the bias.** Donor cities lose drivers to Austin (or absorb returning supply) → donor trends adjust → synthetic Austin trends with donor adjustments → counterfactual *underestimates* "Austin if untreated" → SC **understates** the true treatment effect. (Sign flips if spillover *helps* donors — e.g., riders priced out of Austin substitute to Houston.)

**What to do.**
- **Geographic buffer.** Restrict donors to cities far enough that supply cross-shopping is negligible.
- **Pre-treatment covariate matching.** Choose donors with similar driver labor markets so spillover impact is balanced (won't eliminate bias but reduces it).
- **Augmented SC** (Ben-Michael, Feller, Rothstein 2021). The outcome model can absorb part of the spillover.
- **Explicit spillover model.** Estimate cross-city supply elasticity and adjust the synthetic counterfactual.
- **Triangulate.** If feasible, validate the SC estimate with a switchback or cluster A/B in a *different* market — divergence flags unmeasured spillover.
]

<a href="#marketplace-matrix" class="nav-btn">← back to matrix</a>
