---
title: "Module 8: Advanced Topics for Tech Interviews"
subtitle: "Synthetic Control, Sequential Testing, Bandits, and DiD"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{css, echo=FALSE}
.remark-code, .remark-inline-code { font-size: 80%; }
.remark-slide-content { padding: 1em 2em; }
.small { font-size: 80%; }
.tiny { font-size: 65%; }
.highlight-box {
  background: #fff3e0;
  border-left: 4px solid #e65100;
  padding: 0.5em 1em;
  margin: 0.5em 0;
}
.blue-box {
  background: #e3f2fd;
  border-left: 4px solid #1565c0;
  padding: 0.5em 1em;
  margin: 0.5em 0;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, fig.align = "center",
  fig.width = 10, fig.height = 4.2, dpi = 150,
  message = FALSE, warning = FALSE
)
library(tidyverse)
theme_set(theme_minimal(base_size = 14))
set.seed(42)
```

# Course Map

<table>
<tr><th>#</th><th>Module</th><th>Status</th></tr>
<tr><td>1</td><td><a href="../module-01/slides.html">The Experimental Ideal</a></td><td>✓ done</td></tr>
<tr><td>2</td><td>SUTVA and When It Breaks</td><td>done</td></tr>
<tr><td>3</td><td>Designing Around Interference</td><td>done</td></tr>
<tr><td>4</td><td>Power and Sample Size</td><td>done</td></tr>
<tr><td>5</td><td>Analyzing Experiments</td><td>done</td></tr>
<tr><td>6</td><td>Multiple Testing & Subgroups</td><td>done</td></tr>
<tr><td>7</td><td>External Validity</td><td>done</td></tr>
<tr><td><b>8</b></td><td><b>Advanced Topics for Tech Interviews</b> <i>(you are here)</i></td><td>current</td></tr>
</table>

---

# When You Can't Randomize: One Treated City

A ride-sharing company rolls out a new regulatory policy in **one** city. No control group, no randomization. How do you estimate the effect?

--

```{r synth-data, echo=FALSE, fig.height=3.8}
# Simulate 10 cities, 20 time periods, treatment at t=11 for city 1
n_cities <- 10
n_time <- 20
treat_time <- 11

city_fe <- c(5, 3, 4, 6, 2, 3.5, 4.5, 5.5, 2.5, 3.8)
time_trend <- seq(0, 2, length.out = n_time)
synth_df <- expand_grid(city = 1:n_cities, time = 1:n_time) |>
  mutate(
    y_base = city_fe[city] + time_trend[time] + rnorm(n(), 0, 0.3),
    treated = (city == 1 & time >= treat_time),
    effect = if_else(treated, -1.5, 0),
    y = y_base + effect,
    city_label = if_else(city == 1, "Treated city", "Donor pool"),
    city_f = factor(city)
  )

ggplot(synth_df, aes(time, y, group = city_f, color = city_label)) +
  geom_line(alpha = 0.7, linewidth = 0.8) +
  geom_vline(xintercept = treat_time - 0.5, linetype = "dashed") +
  annotate("text", x = treat_time - 0.3, y = 8.5, label = "Policy change",
           hjust = 1, size = 4) +
  scale_color_manual(values = c("grey60", "firebrick")) +
  labs(title = "One treated city, nine potential controls",
       x = "Time", y = "Outcome (e.g., rides per capita)", color = "")
```

--

No single donor city is a good control. But a **weighted combination** might be.

---

# Synthetic Control: Building a Counterfactual

The idea: find weights $w_j \geq 0$, $\sum w_j = 1$ so the weighted donor pool matches the treated city pre-treatment.

```{r synth-weights, echo=FALSE}
# Simple synthetic control via pre-treatment OLS
pre_data <- synth_df |> filter(time < treat_time)
pre_wide <- pre_data |>
  select(city, time, y) |>
  pivot_wider(names_from = city, values_from = y)

# Target: city 1. Donors: cities 2-10. Constrained regression (simplified)
y_target <- pre_wide$`1`
X_donors <- as.matrix(pre_wide[, 3:11])

# Non-negative least squares (simple projection + clamp)
raw_w <- coef(lm(y_target ~ X_donors - 1))
raw_w[is.na(raw_w) | raw_w < 0] <- 0
weights <- raw_w / sum(raw_w)

donor_names <- paste0("City ", 2:10)
keep <- weights > 0.01
w_df <- tibble(City = donor_names[keep],
               Weight = round(weights[keep], 3))
knitr::kable(w_df, caption = "Synthetic control weights (cities with w > 0.01)")
```

--

```{r synth-result, echo=FALSE, fig.height=3}
# Construct synthetic control outcome
post_wide <- synth_df |>
  select(city, time, y) |>
  pivot_wider(names_from = city, values_from = y)

synth_y <- as.matrix(post_wide[, 3:11]) %*% weights

sc_plot_df <- tibble(
  time = 1:n_time,
  Treated = post_wide$`1`,
  `Synthetic control` = as.numeric(synth_y)
) |> pivot_longer(-time, names_to = "Series", values_to = "y")

ggplot(sc_plot_df, aes(time, y, color = Series)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = treat_time - 0.5, linetype = "dashed") +
  scale_color_manual(values = c("steelblue", "firebrick")) +
  labs(title = "Synthetic control tracks pre-treatment, reveals post-treatment gap",
       x = "Time", y = "Outcome", color = "")
```

The gap after treatment = estimated causal effect.

---

# Synthetic Control: Key Points

.pull-left[
**When to use it:**
- Only one (or very few) treated units
- Good donor pool of untreated units
- Long pre-treatment period available

**Strengths:**
- Transparent: you see the weights
- Pre-treatment fit validates the approach
- No parallel trends assumption needed
]

.pull-right[
**Limitations:**
- Inference is hard (permutation-based)
- Requires good pre-treatment fit
- Sensitive to donor pool selection

**Interview tip:**
"How would you measure the effect of a policy change in one city?"
$\rightarrow$ Synthetic control is the go-to answer when the unit count is too small for DiD.
]

.highlight-box[
**Key insight:** Synthetic control constructs the counterfactual *from data*, rather than assuming parallel trends with any single comparison unit.
]

---

# Regression Discontinuity: Local Randomization

A ride-sharing platform offers a bonus to drivers who complete 100+ trips/month. Drivers at 99 vs 101 trips are essentially identical --- but only those above the cutoff get the bonus.

```{r rd-sim, echo=FALSE, fig.height=3.8}
n_rd <- 800
rd_df <- tibble(
  trips = runif(n_rd, 60, 140),
  above = trips >= 100,
  # True effect = 5 more rides next month for those above cutoff
  y = 40 + 0.3 * (trips - 100) + 5 * above + rnorm(n_rd, 0, 4)
)

ggplot(rd_df, aes(trips, y, color = above)) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  geom_vline(xintercept = 100, linetype = "dashed") +
  scale_color_manual(values = c("grey50", "steelblue"),
                     labels = c("Below cutoff", "Above cutoff")) +
  labs(title = "Regression Discontinuity: jump at the cutoff = causal effect",
       x = "Trips this month (running variable)",
       y = "Rides next month", color = "") +
  theme(legend.position = "top")
```

--

.highlight-box[
**RD intuition:** near the cutoff, treatment is "as good as random." The jump in the outcome at the threshold estimates a **local** causal effect (LATE at the cutoff).
]

---

# RD: Sharp vs Fuzzy

.pull-left[
**Sharp RD:** treatment is a deterministic function of the running variable.

$D_i = \mathbf{1}[X_i \geq c]$

Everyone above the cutoff is treated, everyone below is not.

Estimate: local regression on each side, take the difference at $c$.
]

.pull-right[
**Fuzzy RD:** crossing the cutoff changes the *probability* of treatment, but compliance is imperfect.

$P(D_i = 1 \mid X_i \geq c) > P(D_i = 1 \mid X_i < c)$

Use the cutoff as an **instrument** for treatment.

Estimate: ratio of outcome jump to treatment jump (Wald/IV).
]

--

.blue-box[
**Bandwidth choice matters:** too narrow = noisy; too wide = bias from functional form. Use data-driven methods (Imbens-Kalyanaraman, Calonico-Cattaneo-Titiunik). In interviews, mention the tradeoff.
]

---

# Difference-in-Differences: The Parallel Trends Bet

Two groups of cities. One group gets a new feature at $t = 5$. The identifying assumption: **absent treatment, both groups would have trended the same way.**

```{r did-sim, echo=FALSE, fig.height=3.8}
n_cities_did <- 20
n_time_did <- 10
treat_time_did <- 5

did_df <- expand_grid(city = 1:n_cities_did, time = 1:n_time_did) |>
  mutate(
    treated_group = city <= 10,
    post = time > treat_time_did,
    city_fe = if_else(treated_group, 2, 0),
    # Parallel trends hold
    y = 10 + city_fe + 0.5 * time + 3 * (treated_group & post) + rnorm(n(), 0, 1),
    Group = if_else(treated_group, "Treatment cities", "Control cities")
  )

did_means <- did_df |>
  group_by(Group, time) |>
  summarise(y = mean(y), .groups = "drop")

ggplot(did_means, aes(time, y, color = Group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  geom_vline(xintercept = treat_time_did + 0.5, linetype = "dashed") +
  scale_color_manual(values = c("grey50", "steelblue")) +
  labs(title = "DiD: parallel pre-trends, then divergence = treatment effect",
       x = "Time", y = "Mean outcome", color = "")
```

--

The DiD estimate: $\hat{\tau} = (\bar{Y}_{T,post} - \bar{Y}_{T,pre}) - (\bar{Y}_{C,post} - \bar{Y}_{C,pre})$

---

# DiD: What If Parallel Trends Fail?

```{r did-violated, echo=FALSE, fig.height=3.8}
did_bad <- expand_grid(city = 1:n_cities_did, time = 1:n_time_did) |>
  mutate(
    treated_group = city <= 10,
    post = time > treat_time_did,
    city_fe = if_else(treated_group, 2, 0),
    # Non-parallel trends: treatment group was already accelerating
    trend = if_else(treated_group, 0.5 + 0.08 * time, 0.5),
    y = 10 + city_fe + trend * time + 1.5 * (treated_group & post) +
      rnorm(n(), 0, 1),
    Group = if_else(treated_group, "Treatment cities", "Control cities")
  )

did_bad_means <- did_bad |>
  group_by(Group, time) |>
  summarise(y = mean(y), .groups = "drop")

ggplot(did_bad_means, aes(time, y, color = Group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  geom_vline(xintercept = treat_time_did + 0.5, linetype = "dashed") +
  scale_color_manual(values = c("grey50", "firebrick")) +
  annotate("text", x = 3, y = 18, label = "Pre-trends diverge!",
           color = "firebrick", size = 5) +
  labs(title = "Violation: treatment group was already trending differently",
       subtitle = "DiD overestimates the effect when treated group was accelerating",
       x = "Time", y = "Mean outcome", color = "")
```

--

.highlight-box[
**How to test:** Plot pre-treatment trends. Run event-study regressions (leads should be zero). If pre-trends diverge, DiD is biased --- consider synthetic control or other methods.
]

---

# Bandits: Explore vs Exploit

Standard A/B test: fixed allocation (50/50), wait for statistical significance, then decide. What if you could **shift traffic toward the winner** during the experiment?

--

```{r bandit-intro, echo=FALSE, fig.height=3.5}
# Thompson Sampling bandit vs fixed A/B
n_rounds <- 1000
true_p <- c(A = 0.10, B = 0.13)  # B is better

# Fixed A/B
set.seed(42)
fixed_choices <- rep(c("A", "B"), each = n_rounds / 2) |> sample()
fixed_rewards <- rbinom(n_rounds, 1, true_p[fixed_choices])
fixed_cumrew <- cumsum(fixed_rewards)

# Thompson Sampling
ts_alpha <- c(A = 1, B = 1)
ts_beta <- c(A = 1, B = 1)
ts_rewards <- numeric(n_rounds)
ts_choices <- character(n_rounds)

for (i in 1:n_rounds) {
  theta <- c(rbeta(1, ts_alpha["A"], ts_beta["A"]),
             rbeta(1, ts_alpha["B"], ts_beta["B"]))
  choice <- c("A", "B")[which.max(theta)]
  reward <- rbinom(1, 1, true_p[choice])
  ts_alpha[choice] <- ts_alpha[choice] + reward
  ts_beta[choice] <- ts_beta[choice] + (1 - reward)
  ts_choices[i] <- choice
  ts_rewards[i] <- reward
}

ts_cumrew <- cumsum(ts_rewards)

bandit_plot <- tibble(
  round = rep(1:n_rounds, 2),
  cumulative_reward = c(fixed_cumrew, ts_cumrew),
  Method = rep(c("Fixed A/B (50/50)", "Thompson Sampling"), each = n_rounds)
)

ggplot(bandit_plot, aes(round, cumulative_reward, color = Method)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("grey50", "steelblue")) +
  labs(title = "Bandit earns more total reward by shifting to the winner",
       x = "Round", y = "Cumulative conversions", color = "") +
  theme(legend.position = "top")
```

--

Thompson Sampling: sample from posterior of each arm's conversion rate. Pick the arm with the highest draw.

---

# Bandits: The Trade-off

.pull-left[
**Bandits win at:**
- Minimizing opportunity cost (regret) during the experiment
- Continuous optimization (no fixed end date)
- Personalization (contextual bandits)
]

.pull-right[
**Bandits lose at:**
- **Statistical inference**: non-uniform allocation biases standard estimators
- Estimating the treatment effect precisely
- Regulatory / pre-registration contexts
]

--

```{r bandit-allocation, echo=FALSE, fig.height=2.8}
ts_frac_b <- cumsum(ts_choices == "B") / 1:n_rounds

ggplot(tibble(round = 1:n_rounds, frac_B = ts_frac_b), aes(round, frac_B)) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
  labs(title = "Thompson Sampling shifts traffic toward B over time",
       x = "Round", y = "Fraction allocated to B") +
  ylim(0, 1)
```

.highlight-box[
**Interview answer:** "Bandits optimize *reward*, A/B tests optimize *learning*. Use bandits when the cost of suboptimal allocation is high and you don't need a precise effect estimate. Use A/B tests when you need to *measure* the effect."
]

---

# Sequential Testing: The Peeking Problem

You run an A/B test. After each day, you check: "is it significant yet?" This is **peeking**, and it inflates your false positive rate.

--

```{r peeking-sim, echo=FALSE, fig.height=3.5}
# Simulate A/A test (no true effect) with repeated peeking
n_sims <- 5000
n_obs <- 1000
peek_times <- seq(50, n_obs, by = 50)  # peek every 50 observations

set.seed(42)
ever_significant <- replicate(n_sims, {
  x <- rnorm(n_obs)  # A arm
  y <- rnorm(n_obs)  # B arm (same distribution = no effect)
  any(sapply(peek_times, function(k) {
    t.test(x[1:k], y[1:k])$p.value < 0.05
  }))
})

no_peek_sig <- replicate(n_sims, {
  x <- rnorm(n_obs)
  y <- rnorm(n_obs)
  t.test(x, y)$p.value < 0.05
})

tibble(
  Method = c("Fixed sample (test once)", "Peek every 50 obs"),
  `False positive rate` = c(mean(no_peek_sig), mean(ever_significant))
) |> knitr::kable(digits = 3)
```

--

.highlight-box[
Peeking at 20 time points inflates the false positive rate from 5% to >25%. The more you peek, the worse it gets. This is **not** a minor technicality --- it's the most common mistake in industry A/B testing.
]

---

# Sequential Testing: The Fix

**Group sequential designs** (Pocock, O'Brien-Fleming) adjust the significance threshold at each interim look so the *overall* Type I error stays at $\alpha$.

```{r seq-boundaries, echo=FALSE, fig.height=3.8}
# O'Brien-Fleming-like spending function (simplified)
n_looks <- 5
info_frac <- (1:n_looks) / n_looks
# Approximate OBF boundaries
obf_z <- qnorm(1 - 0.025 / sqrt(info_frac))
pocock_z <- rep(qnorm(1 - 0.025 / 1.5), n_looks)  # approximate

bounds_df <- tibble(
  Look = rep(1:n_looks, 2),
  info = rep(info_frac, 2),
  z_boundary = c(obf_z, pocock_z),
  Method = rep(c("O'Brien-Fleming", "Pocock"), each = n_looks)
)

ggplot(bounds_df, aes(Look, z_boundary, color = Method)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_hline(yintercept = 1.96, linetype = "dashed", color = "grey50") +
  annotate("text", x = 1.5, y = 2.05, label = "Fixed z = 1.96", color = "grey50") +
  scale_color_manual(values = c("steelblue", "firebrick")) +
  labs(title = "Sequential boundaries: harder to reject early, easier late",
       x = "Interim look", y = "z-boundary for rejection") +
  theme(legend.position = "top")
```

--

.blue-box[
**O'Brien-Fleming:** very conservative early (hard to stop), nearly standard at the end. **Pocock:** equal boundary at each look, but final-look threshold is stricter than 1.96. Both control overall $\alpha = 0.05$.
]

---

# Always-Valid Inference: Confidence Sequences

Modern approach: instead of pre-specifying look times, construct a **confidence sequence** that is valid at *any* stopping time.

--

.pull-left[
**Traditional CI:** valid only at fixed sample size $n$.

$$\bar{X}_n \pm z_{\alpha/2} \cdot \frac{s}{\sqrt{n}}$$

Shrinks as $n$ grows. If you stop when it excludes zero, $\alpha > 0.05$.
]

.pull-right[
**Confidence sequence:** valid at *all* $n$ simultaneously.

$$\bar{X}_n \pm \sqrt{\frac{2 s^2 (1 + 1/n)}{n} \log\left(\frac{\sqrt{n + 1}}{\alpha}\right)}$$

Wider than fixed-$n$ CI, but you can peek freely.
]

--

.highlight-box[
**Interview tip:** "We use always-valid p-values / confidence sequences so product teams can check results anytime without inflating false positives. The cost is ~20-30% more samples to reach the same power."
]

---

# Metric Selection: What Are You Optimizing?

.pull-left[
**Success metrics** (primary):
- What you're trying to improve
- e.g., bookings per user, revenue per session
- Needs to be sensitive enough to detect changes

**Guardrail metrics**:
- Must not get worse
- e.g., crash rate, latency, customer complaints
- Often tested one-sided
]

.pull-right[
**Surrogate outcomes:**
- Long-term outcome (12-mo retention) is slow
- Use a *proxy* that's measurable quickly
- e.g., 1st-week engagement predicts 12-mo retention
- Requires validated relationship

**OEC (Overall Evaluation Criterion):**
- Single composite metric for decision-making
- Weighted combination of success + guardrails
- Makes launch decisions mechanical
]

--

.highlight-box[
**Interview framework:** "What's the primary metric? What are the guardrails? How long do we need to run? Is there a surrogate that lets us decide faster?"
]

---

# Interview Framework: Designing an Experiment

When an interviewer asks "How would you design an experiment to test X?", use this structure:

--

**1. Define the question:** What's the causal effect we want to estimate? On whom?

--

**2. Unit of randomization:** User? Session? City? (Think about SUTVA)

--

**3. Metrics:** Primary success metric, guardrails, surrogates

--

**4. Power:** MDE, sample size, how long to run

--

**5. Analysis plan:** What estimator? How to handle non-compliance? Multiple comparisons?

--

**6. Threats:** Interference/spillovers, novelty effects, survivorship bias, peeking

--

.blue-box[
Walk through this checklist out loud. Interviewers want to see structured thinking, not just the "right" answer. Every step is a chance to demonstrate depth.
]

---

# Application: Ride-sharing Policy Change

**Scenario:** A city bans surge pricing. Only one city is affected.

.pull-left[
**Synthetic control approach:**
1. Donor pool: similar cities without the ban
2. Match on pre-ban metrics (rides, revenue, wait times)
3. Construct synthetic city
4. Gap after ban = causal effect
5. Placebo tests: apply method to untreated cities
]

.pull-right[
```{r app-synth, echo=FALSE, fig.height=3.5}
# Re-use synth data, relabel
sc_app <- tibble(
  week = 1:n_time,
  `Treated city` = post_wide$`1`,
  `Synthetic control` = as.numeric(synth_y)
) |> pivot_longer(-week, names_to = "Series", values_to = "rides")

ggplot(sc_app, aes(week, rides, color = Series)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = treat_time - 0.5, linetype = "dashed") +
  annotate("text", x = 15, y = 5.5, label = "Surge ban\neffect",
           color = "firebrick", size = 4) +
  scale_color_manual(values = c("steelblue", "firebrick")) +
  labs(title = "Estimated effect of surge pricing ban",
       x = "Week", y = "Rides per capita", color = "")
```
]

---

# Application: Feature Rollout with Sequential Testing

**Scenario:** Rolling out a new matching algorithm. Want to detect a 2% improvement in completion rate, but product wants to check results weekly.

--

.small[
```{r app-seq, echo=FALSE, fig.height=3.5}
# Simulate a sequential test with a real effect
set.seed(123)
n_per_week <- 5000
n_weeks <- 8
true_effect <- 0.02
base_rate <- 0.70

weekly_data <- tibble(week = 1:n_weeks) |>
  mutate(
    n_cum = n_per_week * week,
    control_conv = map_dbl(n_cum, ~ mean(rbinom(.x, 1, base_rate))),
    treat_conv = map_dbl(n_cum, ~ mean(rbinom(.x, 1, base_rate + true_effect))),
    diff = treat_conv - control_conv,
    se = sqrt(treat_conv * (1 - treat_conv) / n_cum +
                control_conv * (1 - control_conv) / n_cum),
    z_stat = diff / se,
    # O'Brien-Fleming boundary (approximate)
    obf_bound = qnorm(1 - 0.025 / sqrt(week / n_weeks)),
    reject = z_stat > obf_bound
  )

ggplot(weekly_data, aes(week)) +
  geom_line(aes(y = z_stat, color = "Test statistic"), linewidth = 1.2) +
  geom_point(aes(y = z_stat, color = "Test statistic"), size = 3) +
  geom_line(aes(y = obf_bound, color = "OBF boundary"), linewidth = 1,
            linetype = "dashed") +
  geom_point(aes(y = obf_bound, color = "OBF boundary"), size = 2) +
  scale_color_manual(values = c("firebrick", "steelblue")) +
  labs(title = "Sequential test: cross the boundary to reject",
       x = "Week", y = "z-statistic / boundary", color = "") +
  theme(legend.position = "top")
```
]

--

Stops early if the effect is large; runs to completion if it's small. Controls Type I error throughout.

---

# Application: Bandit for Driver Incentives

**Scenario:** Testing 4 incentive levels ($5, $10, $15, $20) to maximize driver sign-ups. Each "wrong" choice costs money.

```{r app-bandit, echo=FALSE, fig.height=3.5}
set.seed(42)
true_rates <- c(0.08, 0.12, 0.14, 0.11)  # $15 is best
arms <- paste0("$", c(5, 10, 15, 20))
n_rounds_b <- 2000
K <- length(true_rates)

# Thompson Sampling
alpha_ts <- rep(1, K)
beta_ts <- rep(1, K)
ts_arm <- integer(n_rounds_b)
ts_rew <- numeric(n_rounds_b)

for (i in 1:n_rounds_b) {
  draws <- rbeta(K, alpha_ts, beta_ts)
  chosen <- which.max(draws)
  reward <- rbinom(1, 1, true_rates[chosen])
  alpha_ts[chosen] <- alpha_ts[chosen] + reward
  beta_ts[chosen] <- beta_ts[chosen] + (1 - reward)
  ts_arm[i] <- chosen
  ts_rew[i] <- reward
}

alloc_df <- tibble(round = 1:n_rounds_b, arm = ts_arm) |>
  mutate(arm_label = arms[arm]) |>
  group_by(arm_label) |>
  mutate(cum_count = row_number()) |>
  ungroup()

alloc_summary <- alloc_df |>
  group_by(arm_label) |>
  summarise(n = n(), .groups = "drop") |>
  mutate(frac = n / sum(n))

ggplot(alloc_summary, aes(arm_label, frac, fill = arm_label)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(round(100 * frac), "%")), vjust = -0.5, size = 5) +
  scale_fill_brewer(palette = "Blues") +
  labs(title = "Thompson Sampling concentrates on the best arm ($15)",
       x = "Incentive level", y = "Fraction of traffic") +
  ylim(0, 0.7)
```

The bandit learns the $15 incentive is best and allocates most traffic there, minimizing wasted spend.

---

# Putting It All Together

| Method | When to use | Key assumption | Estimates |
|--------|------------|----------------|-----------|
| **Synthetic control** | 1 treated unit, donor pool | Pre-treatment fit | Treated unit's effect |
| **RD** | Treatment assigned by cutoff | Continuity at cutoff | LATE at cutoff |
| **DiD** | Panel data, staggered adoption | Parallel trends | ATT |
| **Bandits** | Optimizing allocation | Stationarity | Regret, not ATE |
| **Sequential testing** | Interim monitoring | Pre-specified boundaries | ATE with valid $\alpha$ |

--

.highlight-box[
**Meta-lesson:** Each method trades off something. Synthetic control needs a good donor pool. RD is local. DiD needs parallel trends. Bandits sacrifice inference. Sequential testing sacrifices power. Know the trade-offs --- that's what interviewers test.
]

---

# Key Takeaways

1. **Synthetic control** builds a data-driven counterfactual from weighted donors when you have only one treated unit.

2. **RD** exploits cutoff-based assignment --- local randomization gives you a credible causal estimate near the threshold.

3. **DiD** relies on parallel trends. Test with pre-trends; if violated, consider alternative methods.

4. **Bandits** optimize reward during the experiment but make inference harder. Use when the cost of suboptimal allocation outweighs the need for measurement.

5. **Sequential testing** lets you peek at results without inflating false positives, at the cost of wider confidence intervals.

6. **Metric selection** is often the most important design decision. Define success, guardrails, and surrogates before running anything.

7. **Use the structured framework** (question, unit, metrics, power, analysis, threats) for any experiment design interview question.

---

# Exercise Preview

In the exercise you will:

1. Implement Thompson Sampling from scratch and compare cumulative regret to a fixed A/B test
2. Simulate peeking in an A/A test and measure false positive inflation
3. Implement a simple group sequential boundary and show it controls $\alpha$
4. Build a basic synthetic control and estimate the treatment effect

See `exercise.R` for the starter code.

---
class: center, middle, inverse

# Backup Slides

---

# Backup: Why Peeking Inflates False Positives

Under $H_0$ (no effect), the test statistic $Z_n = \frac{\bar{X}_n - \bar{Y}_n}{\text{SE}_n}$ follows a random walk:

--

- At any single time $n$, $P(|Z_n| > 1.96) = 0.05$

--

- But across many times, the process will *eventually* cross 1.96 with probability 1 (by the law of the iterated logarithm)

--

- With $k$ equally-spaced looks, $P(\text{ever reject}) \approx \alpha \cdot \rho(k)$ where $\rho(k)$ grows with $k$

--

.highlight-box[
The Brownian motion analogy: a random walk that's currently at 0 will eventually hit any finite boundary. More looks = more chances to cross. Sequential boundaries grow over time to compensate.
]

---

# Backup: Thompson Sampling Derivation

**Setup:** $K$ arms, arm $k$ has unknown conversion rate $\theta_k$. Prior: $\theta_k \sim \text{Beta}(1, 1)$ (uniform).

--

**Update rule:** After observing $s_k$ successes and $f_k$ failures on arm $k$:

$$\theta_k \mid \text{data} \sim \text{Beta}(1 + s_k, 1 + f_k)$$

--

**Selection:** At each round, draw $\tilde{\theta}_k \sim \text{Beta}(1 + s_k, 1 + f_k)$ for each arm. Play the arm with the highest draw.

--

**Why it works:** The probability of selecting arm $k$ equals the posterior probability that arm $k$ is optimal. This naturally balances exploration (uncertain arms get sampled) and exploitation (high-reward arms get sampled more).

---

# Backup: Synthetic Control Inference

Standard inference (t-tests, CIs) doesn't apply --- you have one treated unit.

--

**Permutation inference:** Apply the synthetic control method to each *untreated* unit (pretend it was treated). This gives you a distribution of placebo effects.

--

**p-value:** Fraction of placebo effects as large as the actual effect.

$$p = \frac{\sum_{j \neq 1} \mathbf{1}[|\hat{\tau}_j| \geq |\hat{\tau}_1|]}{J}$$

--

.blue-box[
With 10 donor units, the smallest possible p-value is 0.10. You need a large donor pool for precise inference --- another limitation of the method.
]

---

# Backup: DiD Event Study Specification

To test parallel trends, estimate the event-study regression:

$$Y_{it} = \alpha_i + \lambda_t + \sum_{k \neq -1} \beta_k \cdot \mathbf{1}[t - t^* = k] \cdot D_i + \varepsilon_{it}$$

--

where $t^*$ is the treatment time, $D_i$ indicates the treated group, and $k = -1$ is the omitted reference period.

--

**What to look for:**
- Pre-treatment coefficients $\beta_k$ for $k < -1$ should be close to zero (parallel pre-trends)
- Post-treatment coefficients $\beta_k$ for $k \geq 0$ estimate dynamic treatment effects
- Confidence intervals on pre-treatment $\beta_k$ tell you about power to detect violations
