---
title: "Module 4: Power and Sample Size"
subtitle: "How Big Does Your Experiment Need to Be?"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{css, echo=FALSE}
.remark-code, .remark-inline-code { font-size: 80%; }
.remark-slide-content { padding: 1em 2em; }
.small { font-size: 80%; }
.tiny { font-size: 65%; }
.highlight-box {
  background: #fff3e0;
  border-left: 4px solid #e65100;
  padding: 0.5em 1em;
  margin: 0.5em 0;
}
.blue-box {
  background: #e3f2fd;
  border-left: 4px solid #1565c0;
  padding: 0.5em 1em;
  margin: 0.5em 0;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, fig.align = "center",
  fig.width = 10, fig.height = 4.2, dpi = 150,
  message = FALSE, warning = FALSE
)
library(tidyverse)
theme_set(theme_minimal(base_size = 14))
set.seed(42)
```

# Course Map

<table>
<tr><th>#</th><th>Module</th><th>Status</th></tr>
<tr><td>1</td><td><a href="../module-01/slides.html">The Experimental Ideal</a></td><td>✓ done</td></tr>
<tr><td>2</td><td><a href="../module-02/slides.html">SUTVA and When It Breaks</a></td><td>✓ done</td></tr>
<tr><td>3</td><td><a href="../module-03/slides.html">Designing Around Interference</a></td><td>✓ done</td></tr>
<tr><td><b>4</b></td><td><b>Power and Sample Size</b> <i>(you are here)</i></td><td>current</td></tr>
<tr><td>5</td><td><a href="../module-05/slides.html">Analyzing Experiments</a></td><td>✓ done</td></tr>
<tr><td>6</td><td>Multiple Testing & Subgroups</td><td>upcoming</td></tr>
<tr><td>7</td><td><a href="../module-07/slides.html">External Validity</a></td><td>✓ done</td></tr>
<tr><td>8</td><td>Advanced Topics</td><td>upcoming</td></tr>
</table>

---

# The Problem: An Underpowered Experiment

A ride-sharing platform runs an A/B test: does a new splash screen increase rider sign-ups? They test on 200 users per arm for one week.

```{r underpowered-setup, echo=FALSE}
n_small <- 200
true_effect <- 0.02  # 2pp lift on a 10% baseline
sim_small <- function(i) {
  control <- rbinom(n_small, 1, prob = 0.10)
  treated <- rbinom(n_small, 1, prob = 0.10 + true_effect)
  t.test(treated, control)$p.value
}
pvals_small <- map_dbl(1:2000, sim_small)
power_small <- mean(pvals_small < 0.05)
```

--

Result: p = 0.38. "No significant effect. Kill the feature."

--

But the true effect **is** 2 percentage points (a 20% relative lift!). With only 200 per arm, the experiment had **`r round(power_small * 100)`% power** -- it almost never would have detected the real effect.

--

.highlight-box[
**The most common mistake in experimentation is running underpowered tests.** You don't learn "the feature doesn't work" -- you learn nothing.
]

---

# What Is Statistical Power?

.blue-box[
**Power** = $P(\text{reject } H_0 \mid H_1 \text{ is true})$ -- the probability of detecting a real effect when it exists.
]

--

```{r power-diagram, echo=FALSE, fig.height=3.5}
x <- seq(-4, 6, length.out = 500)
null_dist <- dnorm(x, mean = 0, sd = 1)
alt_dist <- dnorm(x, mean = 2.5, sd = 1)
crit <- qnorm(0.975)

df_plot <- tibble(x = x, null = null_dist, alt = alt_dist)

ggplot(df_plot) +
  geom_line(aes(x, null), color = "grey40", linewidth = 1) +
  geom_line(aes(x, alt), color = "steelblue", linewidth = 1) +
  geom_area(data = df_plot |> filter(x > crit),
            aes(x, null), fill = "firebrick", alpha = 0.3) +
  geom_area(data = df_plot |> filter(x > crit),
            aes(x, alt), fill = "steelblue", alpha = 0.3) +
  geom_vline(xintercept = crit, linetype = "dashed") +
  annotate("text", x = -1.2, y = 0.3, label = "H0 (no effect)", color = "grey30", size = 4.5) +
  annotate("text", x = 3.7, y = 0.3, label = "H1 (real effect)", color = "steelblue", size = 4.5) +
  annotate("text", x = crit + 0.15, y = 0.42, label = "critical\nvalue", hjust = 0, size = 3.5) +
  annotate("text", x = 2.5, y = 0.05, label = "Power", color = "steelblue", fontface = "bold", size = 5) +
  annotate("text", x = 2.7, y = 0.015, label = expression(alpha), color = "firebrick", size = 4) +
  labs(x = "Test statistic", y = "Density",
       title = "Power = shaded blue area under H1 beyond the critical value") +
  theme(axis.text.y = element_blank())
```

---

# The Four Knobs

Power depends on exactly **four** things:

| Knob | Turn it this way | Power goes... |
|------|-----------------|---------------|
| Effect size $(\delta)$ | Larger | Up |
| Sample size $(n)$ | Larger | Up |
| Variance $(\sigma^2)$ | Smaller | Up |
| Significance level $(\alpha)$ | Larger | Up |

--

The formula for a two-sample test with equal arms:

$$n_{\text{per arm}} = \frac{(z_{1-\alpha/2} + z_\beta)^2 \cdot 2\sigma^2}{\delta^2}$$

--

.highlight-box[
**But you should rarely use this formula.** Simulation-based power analysis is more flexible, more intuitive, and handles real-world complications (clustering, non-normal outcomes, covariates).
]

---

# Simulation-Based Power Analysis

The idea is dead simple:

1. **Simulate** data under $H_1$ (with a known effect)
2. **Analyze** it the way you would analyze the real experiment
3. **Check** if you reject $H_0$
4. **Repeat** 1,000+ times. Power = fraction of rejections.

--

```{r sim-power-basic}
sim_power <- function(n_per_arm, effect, baseline = 0.10, nsim = 2000) {
  rejections <- map_lgl(1:nsim, function(i) {
    control <- rbinom(n_per_arm, 1, prob = baseline)
    treated <- rbinom(n_per_arm, 1, prob = baseline + effect)
    t.test(treated, control)$p.value < 0.05
  })
  mean(rejections)
}
```

--

```{r sim-power-check}
sim_power(n_per_arm = 200, effect = 0.02)   # too small
sim_power(n_per_arm = 5000, effect = 0.02)  # much better
```

---

# Power Curves: Seeing the Trade-offs

```{r power-curve, echo=FALSE, fig.height=4}
sample_sizes <- c(100, 200, 500, 1000, 2000, 5000, 10000, 15000, 20000)
power_results <- tibble(
  n_per_arm = sample_sizes,
  power = map_dbl(sample_sizes, ~ sim_power(.x, effect = 0.02, nsim = 1000))
)

ggplot(power_results, aes(n_per_arm, power)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_point(size = 3, color = "steelblue") +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "firebrick") +
  annotate("text", x = 15000, y = 0.83, label = "80% power threshold",
           color = "firebrick", size = 4) +
  scale_x_continuous(labels = scales::comma) +
  labs(title = "Power curve: detecting a 2pp lift on 10% baseline",
       x = "Sample size per arm", y = "Power")
```

--

You need roughly **5,000-10,000 users per arm** to reliably detect a 2 percentage point lift on a 10% baseline conversion rate.

---

# Minimum Detectable Effect (MDE)

.blue-box[
**MDE** = the smallest effect your experiment can reliably detect (at 80% power, alpha = 0.05).
]

--

The business doesn't ask "what's our power?" They ask:

> "We can run this test for 2 weeks and get 10,000 users per arm. **What's the smallest lift we can detect?**"

--

```{r mde-curve, echo=FALSE, fig.height=3.2}
effects <- seq(0.005, 0.06, by = 0.005)
mde_results <- tibble(
  effect = effects,
  power = map_dbl(effects, ~ sim_power(n_per_arm = 5000, effect = .x, nsim = 1000))
)

ggplot(mde_results, aes(effect, power)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_point(size = 3, color = "steelblue") +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "firebrick") +
  scale_x_continuous(labels = scales::percent) +
  labs(title = "MDE: What effect can 5,000 per arm detect?",
       subtitle = "MDE ~ 2% (where the curve crosses 80% power)",
       x = "True effect size", y = "Power")
```

---

# Application: Online Ads

An ad platform wants to test a new ad format. Baseline click-through rate (CTR) is 3%. They want to detect a 2 percentage point lift (to 5%).

--

.pull-left[
```{r ads-power}
# How many impressions per arm?
sizes <- c(100, 200, 500, 1000, 2000)
ads_power <- tibble(
  n = sizes,
  power = map_dbl(sizes, ~ sim_power(
    .x, effect = 0.02,
    baseline = 0.03, nsim = 1000
  ))
)
ads_power |> knitr::kable(digits = 2)
```
]

--

.pull-right[
At CTR = 3%, the variance is very low ( $p(1-p) \approx 0.03$ ), so detecting a 2pp absolute lift is easier than at a 50% baseline.

You need about **500-1,000 impressions per arm** -- quite feasible for an ad platform.

But what if you want to detect a **0.1pp** lift? That's ~500,000 per arm.
]

---

# The Clustering Problem

What if randomization is at the **city level**, not the individual level?

--

```{r cluster-sim-setup}
sim_power_cluster <- function(n_clusters, n_per_cluster, effect,
                              icc = 0.05, nsim = 1000) {
  rejections <- map_lgl(1:nsim, function(i) {
    cluster_id <- rep(1:n_clusters, each = n_per_cluster)
    treatment <- rep(sample(rep(c(0, 1), each = n_clusters / 2)),
                     each = n_per_cluster)
    # Cluster random effect
    cluster_effect <- rep(rnorm(n_clusters, 0,
                                sd = sqrt(icc / (1 - icc))),
                          each = n_per_cluster)
    y <- rnorm(n_clusters * n_per_cluster,
               mean = treatment * effect + cluster_effect, sd = 1)
    # Cluster-robust inference
    df <- data.frame(y = y, treatment = treatment, cluster = cluster_id)
    coef(summary(lm(y ~ treatment, data = df)))["treatment", "Pr(>|t|)"] < 0.05
  })
  mean(rejections)
}
```

---

# Clustering Kills Power

```{r cluster-comparison, echo=FALSE, fig.height=4}
# Compare: 1000 individuals vs 20 clusters of 50
individual_power <- map_dbl(
  seq(0.05, 0.4, by = 0.05),
  ~ sim_power(500, effect = .x, baseline = 0.5, nsim = 500)
)

cluster_power <- map_dbl(
  seq(0.05, 0.4, by = 0.05),
  ~ sim_power_cluster(20, 50, effect = .x, icc = 0.05, nsim = 500)
)

bind_rows(
  tibble(effect = seq(0.05, 0.4, by = 0.05),
         power = individual_power, design = "1,000 individuals"),
  tibble(effect = seq(0.05, 0.4, by = 0.05),
         power = cluster_power, design = "20 clusters of 50")
) |>
  ggplot(aes(effect, power, color = design)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "grey50") +
  labs(title = "Same total N = 1,000, but clustering destroys power",
       subtitle = "With 20 clusters and ICC = 0.05, you need much larger effects to reach 80%",
       x = "True effect size", y = "Power", color = "Design") +
  theme(legend.position = "top")
```

---

# The Design Effect and ICC

.blue-box[
**ICC** (Intra-cluster Correlation) = fraction of total variance that is *between* clusters.

**Design effect** $= 1 + (m - 1) \times \text{ICC}$, where $m$ = cluster size.
]

--

```{r design-effect, echo=FALSE}
tibble(
  `Cluster size (m)` = c(10, 50, 100, 500),
  `ICC = 0.01` = 1 + (c(10, 50, 100, 500) - 1) * 0.01,
  `ICC = 0.05` = 1 + (c(10, 50, 100, 500) - 1) * 0.05,
  `ICC = 0.10` = 1 + (c(10, 50, 100, 500) - 1) * 0.10
) |> knitr::kable(digits = 1)
```

--

.highlight-box[
**Design effect** = how many times more samples you need compared to individual-level randomization. With 500 users per cluster and ICC = 0.05, you need **25x** the sample size. Adding more users per cluster barely helps -- you need **more clusters**.
]

---

# Application: Ride-Sharing City-Level Experiment

A ride-sharing company tests the **zone-notification** feature (Modules 1–2) in 20 cities (10 treatment, 10 control), ~500 drivers per city. The ICC for driver accept rate within a city is 0.08.

--

.pull-left[
**Design effect:** $1 + (500 - 1) \times 0.08 = 40.9$

**Effective sample size:** $\frac{10{,}000}{40.9} \approx 245$ per arm

Even with 10,000 drivers total, you effectively have 245.
]

--

.pull-right[
```{r city-power, echo=FALSE}
city_powers <- tibble(
  n_cities = c(10, 20, 40, 60, 100),
  power = map_dbl(c(10, 20, 40, 60, 100),
                  ~ sim_power_cluster(.x, 500, effect = 0.15,
                                       icc = 0.08, nsim = 500))
)
city_powers |> knitr::kable(
  col.names = c("Cities per arm", "Power"), digits = 2)
```

With only 10 cities per arm, power is low even for a large effect (0.15 SD). You need **40+ cities per arm** to get adequate power.
]

---

# More Clusters vs. Bigger Clusters

```{r more-vs-bigger, echo=FALSE, fig.height=4}
# Fix total N = 2000, vary cluster count
configs <- tibble(
  n_clusters = c(10, 20, 40, 100, 200),
  n_per_cluster = 2000 / c(10, 20, 40, 100, 200)
) |>
  mutate(
    power = map2_dbl(n_clusters, n_per_cluster,
                     ~ sim_power_cluster(.x, .y, effect = 0.2,
                                          icc = 0.05, nsim = 500)),
    label = paste0(n_clusters, " clusters x ", n_per_cluster, " each")
  )

ggplot(configs, aes(n_clusters, power)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_point(size = 3, color = "steelblue") +
  geom_text(aes(label = label), vjust = -1, size = 3.5) +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "firebrick") +
  labs(title = "Total N = 2,000 (fixed). More clusters >> bigger clusters.",
       subtitle = "ICC = 0.05. Power improves by splitting into more, smaller clusters.",
       x = "Number of clusters", y = "Power") +
  ylim(0, 1)
```

---

# Reducing Variance with Pre-Treatment Covariates

You can boost power without more users by **reducing residual variance**:

$$n \propto \frac{\sigma^2}{\delta^2} \quad \Rightarrow \quad \text{cut } \sigma^2 \text{ in half} \Leftrightarrow \text{double } n$$

--

```{r covariate-power, fig.height=3.2}
sim_power_covariate <- function(n_per_arm, effect, r_squared, nsim = 1000) {
  rejections <- map_lgl(1:nsim, function(i) {
    x_pre <- rnorm(2 * n_per_arm)
    treat <- rep(c(0, 1), each = n_per_arm)
    y <- sqrt(r_squared) * x_pre + treat * effect +
         rnorm(2 * n_per_arm, sd = sqrt(1 - r_squared))
    coef(summary(lm(y ~ treat + x_pre)))[2, 4] < 0.05
  })
  mean(rejections)
}
cat("Without covariate:", sim_power(2000, effect = 0.1, baseline = 0.5, nsim = 1000), "\n")
cat("With covariate (R^2=0.3):", sim_power_covariate(2000, effect = 0.1, r_squared = 0.3), "\n")
```

--

.highlight-box[
**CUPED preview:** Using pre-experiment data (last week's clicks, last month's rides) as a covariate can reduce variance by 30-50%, equivalent to doubling your sample size for free. Module 5 covers the mechanics.
]

---

# Summary: The Power Checklist

Before running any experiment, answer these:

--

1. **What is the MDE?** What's the smallest effect that matters for the business?

--

2. **What is the baseline rate and variance?** Look at historical data.

--

3. **Is randomization at the individual level or clustered?** If clustered, how many clusters do you have? What is the ICC?

--

4. **Can you reduce variance?** Pre-treatment covariates, stratification, CUPED?

--

5. **How long do you need to run?** More time = more users, but watch out for novelty effects and contamination.

--

.highlight-box[
**Rules of thumb:** (1) 80% power is the standard minimum. (2) You almost always need more samples than you think. (3) For clustered designs, the number of *clusters* matters far more than the number of units per cluster.
]

---

# Key Takeaways

1. **Power** = $P(\text{reject } H_0 \mid H_1 \text{ true})$. Low power means you learn nothing from a null result.

2. **Four knobs:** effect size, sample size, variance, significance level. You usually control $n$ and can reduce $\sigma^2$.

3. **Simulation-based power** is more flexible than formulas. Simulate your analysis pipeline directly.

4. **MDE** is the question the business actually asks: "Given our traffic, what can we detect?"

5. **Clustering** destroys power via the design effect $1 + (m-1) \times \text{ICC}$. Add more clusters, not more units per cluster.

6. **Pre-treatment covariates** can reduce variance dramatically -- a free lunch for power. (Details in Module 5.)

---

# Exercise Preview

In the exercise you will:

1. Build a simulation-based power calculator for proportions
2. Generate power curves for different sample sizes and effect sizes
3. Demonstrate how clustering kills power (compare individual vs. cluster randomization)
4. Compute the design effect and effective sample size
5. Show how adding a covariate recovers power lost to noise

See `exercise.R` for the starter code.

---
class: center, middle, inverse

# Backup Slides

---
name: power-formula-derivation

# Backup: Power Formula Derivation .tiny[(for two-sample z-test)]

Under $H_0$: $\bar{Y}_T - \bar{Y}_C \sim N\left(0, \frac{2\sigma^2}{n}\right)$

Under $H_1$: $\bar{Y}_T - \bar{Y}_C \sim N\left(\delta, \frac{2\sigma^2}{n}\right)$

--

Reject $H_0$ when $\frac{\bar{Y}_T - \bar{Y}_C}{\sqrt{2\sigma^2/n}} > z_{1-\alpha/2}$

--

Power = $P\left(\frac{\bar{Y}_T - \bar{Y}_C}{\sqrt{2\sigma^2/n}} > z_{1-\alpha/2} \;\middle|\; H_1\right)$

--

$= P\left(Z + \frac{\delta}{\sqrt{2\sigma^2/n}} > z_{1-\alpha/2}\right)$ where $Z \sim N(0,1)$

$= P\left(Z > z_{1-\alpha/2} - \frac{\delta\sqrt{n}}{\sqrt{2}\sigma}\right)$

--

Setting power $= 1 - \beta$ and solving for $n$:

$$n = \frac{2\sigma^2 (z_{1-\alpha/2} + z_{1-\beta})^2}{\delta^2}$$

---
name: icc-derivation

# Backup: ICC and Design Effect

**ICC** = $\rho = \frac{\sigma_b^2}{\sigma_b^2 + \sigma_w^2}$, where $\sigma_b^2$ = between-cluster variance and $\sigma_w^2$ = within-cluster variance.

--

The variance of the cluster mean is:

$$\text{Var}(\bar{Y}_{c}) = \frac{\sigma_w^2}{m} + \sigma_b^2$$

--

The variance of the treatment effect estimator with $J$ clusters per arm of size $m$:

$$\text{Var}(\hat{\tau}) = \frac{2(\sigma_w^2 + m\sigma_b^2)}{Jm} = \frac{2\sigma^2}{Jm} \cdot [1 + (m-1)\rho]$$

--

The term $\text{DEFF} = 1 + (m-1)\rho$ is the **design effect**. It inflates the variance relative to individual-level randomization.

Effective sample size: $n_{\text{eff}} = \frac{Jm}{\text{DEFF}}$

---
name: mde-formula

# Backup: MDE Formula

From the power formula, solving for $\delta$ instead of $n$:

$$\text{MDE} = (z_{1-\alpha/2} + z_{1-\beta}) \cdot \sqrt{\frac{2\sigma^2}{n}}$$

--

For a binary outcome with baseline $p$:

$$\text{MDE} = (z_{0.975} + z_{0.80}) \cdot \sqrt{\frac{2p(1-p)}{n}} \approx 2.8 \cdot \sqrt{\frac{2p(1-p)}{n}}$$

--

For clustered designs, multiply the standard error by $\sqrt{\text{DEFF}}$:

$$\text{MDE}_{\text{clustered}} = (z_{1-\alpha/2} + z_{1-\beta}) \cdot \sqrt{\frac{2\sigma^2 \cdot [1 + (m-1)\rho]}{Jm}}$$
