---
title: "Module 2: SUTVA and When It Breaks"
subtitle: "Interference, Spillovers, and Why Your Experiment Might Be Lying"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{css, echo=FALSE}
.remark-code, .remark-inline-code { font-size: 80%; }
.remark-slide-content { padding: 1em 2em; }
.small { font-size: 80%; }
.tiny { font-size: 65%; }
.highlight-box {
  background: #fff3e0;
  border-left: 4px solid #e65100;
  padding: 0.5em 1em;
  margin: 0.5em 0;
}
.blue-box {
  background: #e3f2fd;
  border-left: 4px solid #1565c0;
  padding: 0.5em 1em;
  margin: 0.5em 0;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, fig.align = "center",
  fig.width = 10, fig.height = 4.2, dpi = 150,
  message = FALSE, warning = FALSE
)
library(tidyverse)
theme_set(theme_minimal(base_size = 14))
set.seed(42)
```

# Course Map

<table>
<tr><th>#</th><th>Module</th><th>Status</th></tr>
<tr><td>1</td><td>The Experimental Ideal</td><td>done</td></tr>
<tr><td><b>2</b></td><td><b>SUTVA and When It Breaks</b> <i>(you are here)</i></td><td>current</td></tr>
<tr><td>3</td><td>Designing Around Interference</td><td>upcoming</td></tr>
<tr><td>4</td><td>Power and Sample Size</td><td>upcoming</td></tr>
<tr><td>5</td><td>Analyzing Experiments</td><td>upcoming</td></tr>
<tr><td>6</td><td>Multiple Testing & Subgroups</td><td>upcoming</td></tr>
<tr><td>7</td><td>External Validity</td><td>upcoming</td></tr>
<tr><td>8</td><td>Advanced Topics</td><td>upcoming</td></tr>
</table>

---

# Last Time: The Experimental Ideal

Module 1 showed that randomization kills selection bias:

$$E[Y \mid D=1] - E[Y \mid D=0] = \text{ATE}$$

--

But this result relied on an assumption we barely mentioned:

.highlight-box[
**SUTVA (Stable Unit Treatment Value Assumption):** Each unit's outcome depends *only* on its own treatment assignment, not on anyone else's.
]

--

Today: what happens when SUTVA breaks, why it almost always breaks in marketplace experiments, and how to detect the problem.

---

# The Setup: Driver Bonus Experiment (Again)

Same experiment as Module 1: randomize 500 drivers to a $500 bonus, 500 to control. Measure 6-month retention.

```{r setup-data, echo=FALSE}
n <- 1000
drivers <- tibble(
  driver_id = 1:n,
  skill = rnorm(n),
  p_retain_0 = plogis(-0.5 + 0.3 * skill),
  p_retain_1 = plogis(-0.5 + 0.3 * skill + 0.5)
)
true_ate <- mean(drivers$p_retain_1 - drivers$p_retain_0)
```

But now let's think about what happens in the **marketplace**...

--

- Bonus drivers are more motivated and take **more rides**
- The total number of ride requests is roughly fixed (in the short run)
- More rides for treated drivers = **fewer rides for control drivers**
- Control drivers earn less = **lower retention** (even without the bonus effect)

--

.highlight-box[
The control group outcome is *contaminated* by the treatment. The naive ATE will be **biased upward** -- it captures the direct bonus effect PLUS the harm to the control group.
]

---

# SUTVA: Stated Precisely

**Component 1 -- No interference:**

$$Y_i = Y_i(D_i) \quad \text{for all } i$$

Unit $i$'s outcome depends only on $i$'s treatment. Not on $D_j$ for any $j \neq i$.

--

**Component 2 -- No hidden versions of treatment:**

If $D_i = D_i' = 1$, then $Y_i(D_i) = Y_i(D_i')$. There is one "treated" state and one "control" state.

--

When SUTVA holds: $2$ potential outcomes per unit $(Y_i(0), Y_i(1))$.

When SUTVA fails: potential outcomes depend on the **entire** treatment vector $\mathbf{D} = (D_1, \ldots, D_N)$.

.highlight-box[
With $N$ units, there are $2^N$ possible treatment vectors. Each unit has $2^N$ potential outcomes instead of $2$. The ATE is no longer well-defined without specifying $\mathbf{D}$.
]

<a href="#sutva-formal" style="position:absolute; bottom:12px; left:40px; font-size:11px; background:#e8eaf6; padding:2px 8px; border-radius:3px; z-index:100;">Formal notation</a>

---

# Seeing the Bias: A Simulation

```{r interference-sim, echo=FALSE, fig.height=4}
sim_marketplace <- function(interference_strength) {
  bonus <- sample(rep(c(0, 1), each = n / 2))
  frac_treated <- mean(bonus)
  interference <- -interference_strength * frac_treated * (1 - bonus)
  y_obs <- rbinom(n, 1, prob = plogis(-0.5 + 0.3 * drivers$skill +
                                        0.5 * bonus + interference))
  mean(y_obs[bonus == 1]) - mean(y_obs[bonus == 0])
}

estimates_clean <- replicate(1000, sim_marketplace(0))
estimates_dirty <- replicate(1000, sim_marketplace(0.4))

bind_rows(
  tibble(est = estimates_clean, scenario = "No interference"),
  tibble(est = estimates_dirty, scenario = "With interference")
) |>
  ggplot(aes(est, fill = scenario)) +
  geom_histogram(bins = 40, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = true_ate, color = "firebrick",
             linewidth = 1.2, linetype = "dashed") +
  annotate("text", x = true_ate + 0.005, y = 55,
           label = paste("True ATE =", round(true_ate, 3)),
           color = "firebrick", hjust = 0, size = 4) +
  scale_fill_manual(values = c("steelblue", "darkorange")) +
  labs(title = "Interference biases the ATE upward",
       subtitle = "Control drivers are harmed by treatment -> naive effect is too large",
       x = "Estimated ATE", y = "Count") +
  theme(legend.position = "top")
```

---

# Where Does the Bias Come From?

.pull-left[
**Without interference:**
- $E[Y(0) \mid D=0]$ = true control outcome
- $E[Y(1) \mid D=1]$ = true treated outcome
- Difference = ATE

**With interference:**
- $E[Y(0) \mid D=0]$ is **too low** (control drivers harmed)
- $E[Y(1) \mid D=1]$ is roughly correct
- Difference = ATE + bias
]

.pull-right[
```{r bias-direction, echo=FALSE, fig.height=4.5}
tibble(
  Group = rep(c("Control", "Treated"), each = 2),
  Scenario = rep(c("No interference", "With interference"), 2),
  Retention = c(
    mean(drivers$p_retain_0),
    mean(drivers$p_retain_0) - 0.04,  # interference reduces control
    mean(drivers$p_retain_1),
    mean(drivers$p_retain_1)   # treated unaffected
  )
) |>
  ggplot(aes(Group, Retention, fill = Scenario)) +
  geom_col(position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("steelblue", "darkorange")) +
  labs(title = "Interference pushes control\nretention DOWN") +
  theme(legend.position = "top")
```
]

---

# The Bias Grows with Interference Strength

```{r bias-vs-strength, echo=FALSE, fig.height=3.8}
strengths <- seq(0, 0.8, by = 0.05)
bias_results <- tibble(
  interference = strengths,
  mean_ate = map_dbl(strengths, ~ mean(replicate(300, sim_marketplace(.x))))
)

ggplot(bias_results, aes(interference, mean_ate)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_point(size = 2) +
  geom_hline(yintercept = true_ate, linetype = "dashed", color = "firebrick") +
  annotate("text", x = 0.5, y = true_ate + 0.005,
           label = "True ATE", color = "firebrick", size = 4) +
  labs(title = "Stronger interference = more bias",
       x = "Interference strength parameter",
       y = "Mean estimated ATE (over 300 simulations)")
```

In a real marketplace, interference strength depends on market thickness (how tight supply is), substitutability, and network density.

---

# Hidden Versions of Treatment

SUTVA also requires **one version** of treatment. This can fail when:

--

.pull-left[
**The treatment differs by context:**
- A $500 bonus in San Francisco vs. rural Iowa
- A new algorithm in a thick market vs. thin market
- Same nudge email, but recipient has 1 co-author vs. 10
]

--

.pull-right[
**The treatment differs by implementation:**
- New driver app version with different bugs on iOS vs. Android
- "Same" coupon delivered via email vs. push notification
- Treatment intensity varies (some drivers get reminded, others don't)
]

--

.highlight-box[
**Interview framing:** "The treatment effect is 5%." Which treatment? For whom? Delivered how? If the answer is "it depends," you may have a hidden versions problem.
]

---

# Why Interviewers Love This Topic

In tech, the three most common SUTVA violations are:

--

**1. Marketplace interference** (ride-sharing, food delivery, e-commerce)
- Supply and demand are linked: treating one side affects the other
- "We tested a new surge algorithm on 10% of riders" -- but those riders affect driver allocation for everyone

--

**2. Network interference** (social media, communication tools)
- Treating user A changes what user B sees in their feed
- Viral features: a treated user shares content with control users

--

**3. Shared resources** (cloud infrastructure, ad auctions)
- Bidding experiments: treating some advertisers changes auction outcomes for all
- Capacity experiments: allocating more servers to treated users reduces capacity for control

---

# Application: Marketplace Interference in Detail

```{r marketplace-diagram, echo=FALSE, fig.height=4}
# Simulate a simple supply-demand marketplace
set.seed(123)
n_drivers <- 200
n_riders <- 150

market <- tibble(
  period = rep(c("Before experiment", "During experiment"), each = 2),
  side = rep(c("Treated drivers", "Control drivers"), 2),
  rides_per_driver = c(
    3.0, 3.0,   # Before: equal
    3.8, 2.2    # During: treated take more, control get fewer
  )
)

ggplot(market, aes(side, rides_per_driver, fill = period)) +
  geom_col(position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("grey70", "steelblue")) +
  geom_hline(yintercept = 3.0, linetype = "dashed", color = "firebrick") +
  annotate("text", x = 2.5, y = 3.1, label = "Pre-experiment average",
           color = "firebrick", size = 3.5) +
  labs(title = "Marketplace interference: a zero-sum game",
       subtitle = "Total rides are ~fixed. Bonus drivers capture more -> control drivers get fewer.",
       x = "", y = "Average rides per driver per day") +
  theme(legend.position = "top")
```

---

# Application: Co-Author Network Spillovers

Consider an experiment on academic researchers. 80 studies are randomized into treatment arms (T0-T3). Each study has 1-3 authors drawn from a shared pool.

```{r coauthor-network, echo=FALSE, fig.height=3.5}
# Simulate co-author network
n_studies <- 80
n_auth_pool <- 60

set.seed(42)
study_data <- tibble(
  study_id = 1:n_studies,
  n_auth = sample(1:3, n_studies, replace = TRUE, prob = c(0.3, 0.4, 0.3)),
  arm = sample(rep(0:3, each = 20))
)

# Create author assignments
author_links <- study_data |>
  rowwise() |>
  mutate(author_id = list(sample(1:n_auth_pool, n_auth))) |>
  ungroup() |>
  unnest(author_id)

# Find co-authored study pairs
coauth_pairs <- author_links |>
  inner_join(author_links, by = "author_id", suffix = c("_a", "_b"),
             relationship = "many-to-many") |>
  filter(study_id_a < study_id_b) |>
  distinct(study_id_a, study_id_b, .keep_all = TRUE)

# Compute spillover exposure
spillover_df <- author_links |>
  inner_join(author_links, by = "author_id", suffix = c("_focal", "_other"),
             relationship = "many-to-many") |>
  filter(study_id_focal != study_id_other) |>
  group_by(study_id_focal) |>
  summarise(
    n_treated_neighbors = sum(arm_other > 0),
    n_total_neighbors = n()
  )

study_data <- study_data |>
  left_join(spillover_df, by = c("study_id" = "study_id_focal")) |>
  mutate(across(c(n_treated_neighbors, n_total_neighbors),
                ~ replace_na(.x, 0)))

# Show distribution of spillover exposure
study_data |>
  mutate(arm_label = paste0("T", arm)) |>
  ggplot(aes(n_treated_neighbors, fill = arm_label)) +
  geom_histogram(binwidth = 1, alpha = 0.7) +
  facet_wrap(~ arm_label, nrow = 1) +
  scale_fill_manual(values = c("grey60", "#66c2a5", "#fc8d62", "#8da0cb")) +
  labs(title = "Co-author spillover exposure by treatment arm",
       subtitle = "Each bar: number of studies with X treated neighbors through shared co-authors",
       x = "Number of treated co-author links", y = "Count") +
  theme(legend.position = "none")
```

---

# Co-Author Spillovers: The Problem

```{r spillover-bias, echo=FALSE, fig.height=3.8}
# Simulate outcomes with spillover
set.seed(42)
study_data <- study_data |>
  mutate(
    direct_effect = case_when(
      arm == 0 ~ 0,
      arm == 1 ~ 0.05,
      arm == 2 ~ 0.10,
      arm == 3 ~ 0.15
    ),
    spillover = 0.02 * n_treated_neighbors,
    reporting_rate = pmin(1, 0.40 + direct_effect + spillover +
                           rnorm(n(), 0, 0.08))
  )

# Naive vs corrected
naive <- mean(study_data$reporting_rate[study_data$arm > 0]) -
  mean(study_data$reporting_rate[study_data$arm == 0])
true_direct <- mean(c(0.05, 0.10, 0.15))

study_data |>
  mutate(treated = if_else(arm > 0, "Treated (T1-T3)", "Control (T0)")) |>
  ggplot(aes(n_treated_neighbors, reporting_rate, color = treated)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  scale_color_manual(values = c("firebrick", "steelblue")) +
  labs(title = "Control studies with more treated co-authors report MORE",
       subtitle = "Spillovers contaminate the control group -> naive effect is biased",
       x = "Number of treated co-author links",
       y = "Reporting rate") +
  theme(legend.position = "top")
```

Control studies connected to treated co-authors show **higher** reporting. The naive treatment effect **underestimates** the true effect (spillovers help controls).

---

# Partial vs General Equilibrium

A crucial distinction for marketplace experiments:

--

.pull-left[
**Partial equilibrium** (small experiment)
- Treat 1% of drivers
- Market barely moves
- Estimate $\approx$ direct effect of bonus
- Interference is negligible
]

--

.pull-right[
**General equilibrium** (full rollout)
- Treat 100% of drivers
- Market adjusts: more drivers → lower wages per ride → some drivers exit anyway
- The effect you measured at 1% **does not predict** the rollout outcome
]

--

.highlight-box[
**The core problem:** experiments estimate partial equilibrium effects. Business decisions require general equilibrium predictions. The gap between them can be large in thick markets.
]

---

# Partial vs GE: Simulated

```{r partial-ge, echo=FALSE, fig.height=3.8}
sim_ge <- function(frac_treated, interference = 0.3) {
  skill <- drivers$skill
  bonus <- sample(c(rep(1, round(n * frac_treated)),
                    rep(0, n - round(n * frac_treated))))
  interf <- -interference * frac_treated * (1 - bonus)
  y <- rbinom(n, 1, prob = plogis(-0.5 + 0.3 * skill + 0.5 * bonus + interf))
  if (sum(bonus == 0) == 0 | sum(bonus == 1) == 0) return(NA_real_)
  mean(y[bonus == 1]) - mean(y[bonus == 0])
}

fracs <- c(0.01, 0.05, 0.10, 0.25, 0.50, 0.75, 0.90)
ge_data <- tibble(
  frac = fracs,
  ate = map_dbl(fracs, ~ mean(replicate(300, sim_ge(.x)), na.rm = TRUE))
)

ggplot(ge_data, aes(frac, ate)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_point(size = 3) +
  geom_hline(yintercept = true_ate, linetype = "dashed", color = "firebrick") +
  annotate("text", x = 0.5, y = true_ate - 0.008,
           label = "True direct effect (no interference)",
           color = "firebrick", size = 3.5) +
  annotate("segment", x = 0.05, xend = 0.05, y = ge_data$ate[2],
           yend = ge_data$ate[2] + 0.03,
           arrow = arrow(length = unit(0.15, "cm")), color = "grey40") +
  annotate("text", x = 0.08, y = ge_data$ate[2] + 0.035,
           label = "Your experiment\n(5% treated)", size = 3, color = "grey40") +
  labs(title = "The experiment predicts the wrong rollout effect",
       subtitle = "At 5% treated, interference is small. At 90%, it dominates.",
       x = "Fraction of drivers treated", y = "Estimated ATE")
```

---

# Detecting SUTVA Violations

How do you know if interference is a problem in your experiment?

--

**1. Domain knowledge:** Ask "does my treatment change the environment for control units?"
- Marketplace: does treating suppliers change demand/supply for control suppliers?
- Network: do treated and control users interact?

--

**2. Check for dose-response in control outcomes:**
- If control units in markets with many treated units have different outcomes than control units in markets with few treated units, interference is present.

--

**3. Vary the treatment fraction:**
- Run the experiment at different saturation levels (e.g., 10% vs 50% treated)
- If the estimated ATE differs, SUTVA is violated

--

**4. Look at control group outcomes over time:**
- If control outcomes *worsen* after the experiment starts, the control group is being contaminated.

---

# How Treatment Fraction Reveals Interference

```{r saturation-test, echo=FALSE, fig.height=3.5}
# Simulate varying treatment fractions in separate "markets"
set.seed(42)
saturation_test <- function(frac, n_per_market = 200) {
  skill <- rnorm(n_per_market)
  bonus <- sample(c(rep(1, round(n_per_market * frac)),
                    rep(0, n_per_market - round(n_per_market * frac))))
  interf <- -0.3 * frac * (1 - bonus)
  y <- rbinom(n_per_market, 1,
              prob = plogis(-0.5 + 0.3 * skill + 0.5 * bonus + interf))
  tibble(
    frac_treated = frac,
    mean_control = if (sum(bonus == 0) > 0) mean(y[bonus == 0]) else NA,
    mean_treated = mean(y[bonus == 1])
  )
}

sat_data <- map_dfr(rep(c(0.1, 0.3, 0.5, 0.7), each = 50), saturation_test)

sat_data |>
  group_by(frac_treated) |>
  summarise(control = mean(mean_control, na.rm = TRUE),
            treated = mean(mean_treated, na.rm = TRUE)) |>
  pivot_longer(c(control, treated), names_to = "group", values_to = "retention") |>
  ggplot(aes(factor(frac_treated), retention, fill = group)) +
  geom_col(position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("firebrick", "steelblue"),
                    labels = c("Control", "Treated")) +
  labs(title = "Saturation test: control outcomes worsen with more treatment",
       subtitle = "Treated outcomes are stable; control outcomes drop as interference increases",
       x = "Fraction treated in market", y = "Mean retention", fill = "") +
  theme(legend.position = "top")
```

.highlight-box[
If control outcomes vary with treatment saturation, SUTVA is violated. The treated-control gap grows even though the *direct* effect is constant.
]

---

# Summary: Types of SUTVA Violations

| Type | Mechanism | Bias direction | Example |
|------|-----------|---------------|---------|
| **Marketplace** | Supply/demand reallocation | Overestimates ATE | Driver bonus steals rides from control |
| **Network** | Information/behavior spreads | Underestimates ATE | Co-author learns about reporting |
| **Shared resource** | Capacity competition | Overestimates ATE | Server allocation experiment |
| **General equilibrium** | Prices/wages adjust | Direction depends | Full rollout changes market structure |

--

.highlight-box[
**The key question for any experiment:** "If I treat more units, does the control group outcome change?" If yes, SUTVA is violated.
]

---

# What Can We Do About It?

Preview of Module 3:

--

1. **Cluster randomization** -- randomize at a level where interference is contained (cities, markets, time periods)

--

2. **Switchback designs** -- alternate treatment on/off over time within the same unit

--

3. **Two-sided marketplace designs** -- randomize on one side, measure effects on the other

--

4. **Ego-cluster randomization** -- for network experiments, randomize the cluster around each unit

--

Each approach trades **bias** for **variance**. Bigger clusters reduce interference bias but give you fewer independent units.

---

# Key Takeaways

1. **SUTVA** = no interference + no hidden versions of treatment. It's an assumption, not a fact.

2. In **marketplaces**, treating one side (drivers) almost always affects the other side (riders) and within-side competitors (other drivers). Naive ATE is biased.

3. In **networks**, treatment spills over through connections. Co-authors, friends, neighbors can all be contaminated.

4. The **direction** of bias depends on whether spillovers help or harm the control group.

5. **Partial vs general equilibrium**: your small experiment estimates a partial equilibrium effect. The full rollout will be different.

6. **Detection**: vary the treatment fraction, check control outcomes against treatment saturation, use domain knowledge.

---

# Exercise Preview

In the exercise you will:

1. Simulate a marketplace with and without interference
2. Show that interference biases the naive ATE upward
3. Sweep over interference strengths and see the bias grow
4. Build a co-author network and simulate spillover contamination
5. Compare partial vs general equilibrium predictions

See `exercise.R` for the starter code.

---
class: center, middle, inverse

# Backup Slides

---
name: sutva-formal

# Backup: SUTVA Formal Notation

Without SUTVA, each unit $i$'s potential outcome depends on the entire treatment vector:

$$Y_i = Y_i(\mathbf{D}) = Y_i(D_1, D_2, \ldots, D_N)$$

With $N$ binary-treatment units, each unit has $2^N$ potential outcomes.

--

**SUTVA simplifies this:**

$$Y_i(\mathbf{D}) = Y_i(D_i) \quad \forall \mathbf{D}, \mathbf{D}' \text{ such that } D_i = D_i'$$

--

Under SUTVA, we can write the observed outcome as:

$$Y_i^{obs} = D_i \cdot Y_i(1) + (1 - D_i) \cdot Y_i(0)$$

and the ATE is well-defined as:

$$\tau = E[Y_i(1) - Y_i(0)]$$

Without SUTVA, we need to define which treatment vector we mean. The "average direct effect" becomes:

$$\tau(\mathbf{d}_{-i}) = E[Y_i(1, \mathbf{d}_{-i}) - Y_i(0, \mathbf{d}_{-i})]$$

which depends on $\mathbf{d}_{-i}$, the treatment assignments of all other units.

---
name: interference-types

# Backup: Taxonomy of Interference

**Direct interference:** unit $j$'s treatment directly changes unit $i$'s outcome.
- Example: vaccinating $j$ reduces $i$'s infection probability.

--

**Indirect interference (through market/prices):** treatment changes an equilibrium variable that affects everyone.
- Example: subsidizing some sellers lowers the market price, affecting all sellers.

--

**Behavioral interference:** treatment changes $j$'s behavior, and $i$ observes and responds.
- Example: co-author $j$ learns about reporting standards and shares this knowledge.

--

**Mechanical interference:** the experimental design itself creates interference.
- Example: in a waitlist design, adding someone to treatment removes them from control, changing the control group composition.

---
name: ge-formal

# Backup: Partial vs General Equilibrium -- Formal

Let $p$ denote a market-clearing price and $\alpha$ the fraction treated.

**Partial equilibrium** (small $\alpha$): price is approximately fixed at $p_0$.

$$\hat{\tau}_{PE} = E[Y_i(1, p_0)] - E[Y_i(0, p_0)]$$

--

**General equilibrium** (large $\alpha$): price adjusts to $p(\alpha)$.

$$\tau_{GE}(\alpha) = E[Y_i(1, p(\alpha))] - E[Y_i(0, p(\alpha))]$$

--

The experiment estimates $\hat{\tau}_{PE}$, but the rollout delivers $\tau_{GE}(1)$.

The gap depends on how much $p$ moves and how sensitive outcomes are to $p$:

$$\hat{\tau}_{PE} - \tau_{GE}(1) = \underbrace{(p_0 - p(1))}_{\text{price change}} \times \underbrace{\frac{\partial}{\partial p}[E[Y_i(1,p)] - E[Y_i(0,p)]]}_{\text{differential price sensitivity}}$$

This is a first-order approximation. In practice, you often cannot sign this without structural modeling.
